---
title: "Getting Set Up With debiasR"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Timing

10 minutes.

## Purpose

Get everyone into the same package and data workflow.

## Setup

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)

has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work examples.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 25)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work
  coverage <- example_data$coverage
  covariates <- example_data$covariates
}
```

## Code-Along

- Load `debiasR`
- Load tidy data and simple mapping packages
- Load the workshop MSOA travel-to-work datasets from `debiasRdata`
- Use `msoa_OD_travel2work` as the observed MPD OD matrix
- Use `census_msoa_OD_travel2work` as the Census benchmark OD matrix
- Inspect core fields: area ID, census population, observed user or device count,
  and contextual covariates

```{r}
example_data$metadata
```

```{r}
data.frame(
  dataset = c(
    "msoa_OD_travel2work",
    "census_msoa_OD_travel2work",
    "coverage",
    "covariates"
  ),
  rows = c(
    nrow(msoa_OD_travel2work),
    nrow(census_msoa_OD_travel2work),
    nrow(coverage),
    nrow(covariates)
  ),
  columns = c(
    ncol(msoa_OD_travel2work),
    ncol(census_msoa_OD_travel2work),
    ncol(coverage),
    ncol(covariates)
  )
)
```

```{r}
head(msoa_OD_travel2work)
head(census_msoa_OD_travel2work)
head(coverage)
```
