---
title: "Measuring Coverage Bias"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette demonstrates the package functions used to measure coverage and
representativeness bias before adjusting OD flows.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
```

## Coverage Score and Coverage Bias

`measure_bias()` starts from area-level benchmark population $P_i$ and observed
active-user count $U_i$. It returns:

$$
p_i = \frac{U_i}{P_i}
$$

$$
b_i = 1 - p_i
$$

where $p_i$ is the coverage score and $b_i$ is the coverage bias.

```{r}
bias <- measure_bias(example_data$coverage)

bias |>
  summarise(
    areas = n(),
    global_coverage_score = sum(user_count) / sum(population),
    median_coverage_score = median(coverage_score),
    min_coverage_score = min(coverage_score),
    max_coverage_score = max(coverage_score)
  )
```

```{r}
bias |>
  arrange(desc(coverage_bias)) |>
  select(origin, population, user_count, coverage_score, coverage_bias) |>
  head(10)
```

## Representativeness Residuals

`validate_bias_residual_structure()` extends the coverage calculation by asking
whether coverage residuals are structured. The default residual is:

$$
e_i = p_i - \bar{p}
$$

where $\bar{p} = \sum_i U_i / \sum_i P_i$. Positive values mean higher
coverage than expected under the global coverage score; negative values mean
lower coverage.

```{r}
bias_structure <- validate_bias_residual_structure(
  coverage_df = example_data$coverage,
  benchmark_od_df = example_data$census_msoa_OD_travel2work,
  covariate_df = example_data$covariates,
  covariate_col = "census_inflow"
)

bias_structure$summary
```

```{r}
bias_structure$benchmark_flow_correlation
bias_structure$covariate_correlation
```

```{r}
bias_structure$area_level |>
  arrange(coverage_score_residual) |>
  select(area, population, user_count, coverage_score, coverage_score_residual) |>
  head(10)
```

## Interpretation

The aim is not only to compute a coverage number. The aim is to locate where
coverage departs from the benchmark population and whether those departures
align with benchmark mobility structure or area-level characteristics.
