---
title: "Measuring Coverage Bias"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Timing

20 minutes.

## Purpose

Show the first useful thing the package does.

## Code-Along

- Use `measure_bias()`
- Compute coverage bias as `1 - user_count / population`
- Interpret under-coverage, apparent over-coverage, and missingness
- Summarise the distribution of bias
- Plot or map bias across MSOAs

## Teaching Point

Participants should be able to say: I can quantify how far the digital trace
population departs from a census benchmark.

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work examples.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 25)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work
  coverage <- example_data$coverage

  bias <- measure_bias(coverage)
}
```

The coverage table is derived from matched origin totals in
`msoa_OD_travel2work` and `census_msoa_OD_travel2work`.

```{r}
bias |>
  summarise(
    areas = n(),
    median_coverage_score = median(coverage_score),
    min_coverage_score = min(coverage_score),
    max_coverage_score = max(coverage_score)
  )
```

```{r}
bias |>
  arrange(desc(coverage_bias)) |>
  select(origin, population, user_count, coverage_score, coverage_bias) |>
  head(10)
```

## Exercise

Which MSOAs are most under-covered?
