---
title: "Identifying and Explaining Bias"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Timing

25-30 minutes.

## Purpose

Move from "there is bias" to "what kinds of places are biased?"

## Code-Along and Demo

- Join bias estimates to census or contextual variables
- Explore bias by deprivation, age structure, urban/rural status, income,
  ethnicity, housing, and related area characteristics
- Use the simple population-only linear residual diagnostic to spot areas whose
  active-user counts sit above or below the benchmark population trend
- Show variable importance, partial dependence, or SHAP-style outputs if
  available
- Connect to the finding that bias is spatial, social, and not random

## Exercise

Which area characteristics seem associated with under-coverage?

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work examples.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 25)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work
  coverage <- example_data$coverage
  covariates <- example_data$covariates

  diagnostics <- validate_bias_residual_structure(
    coverage_df = coverage,
    benchmark_od_df = census_msoa_OD_travel2work,
    covariate_df = covariates,
    covariate_col = "census_inflow"
  )

  population_lm_diagnostics <- validate_bias_residual_structure(
    coverage_df = coverage,
    benchmark_od_df = census_msoa_OD_travel2work,
    covariate_df = covariates,
    covariate_col = "census_inflow",
    residual_type = "population_lm"
  )
}
```

```{r}
diagnostics$summary
```

```{r}
diagnostics$covariate_correlation
diagnostics$benchmark_flow_correlation
```

```{r}
population_lm_diagnostics$population_lm
```

```{r}
diagnostics$area_level |>
  arrange(coverage_score_residual) |>
  select(area, population, user_count, coverage_score, coverage_score_residual) |>
  head(10)
```

```{r}
population_lm_diagnostics$area_level |>
  arrange(desc(abs(population_lm_residual))) |>
  select(
    area,
    population,
    user_count,
    population_lm_expected_user_count,
    population_lm_residual
  ) |>
  head(10)
```
