---
title: "Adjusting Biases in Population or Mobility Estimates"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Timing

25-30 minutes.

## Purpose

Show the practical payoff from measuring and explaining bias.

## Code-Along

- Derive correction factors from measured coverage
- Apply adjustment to counts or OD flows
- Use inverse penetration weighting or the relevant `debiasR` adjustment method
- Compare raw and adjusted estimates
- Discuss what adjustment can and cannot fix

## Teaching Point

Adjustment is useful, but it rests on assumptions. The workshop should make
participants more capable and more cautious.

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work examples.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 25)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work
  coverage <- example_data$coverage
}
```

```{r}
adj_inverse <- adjust_inverse_penetration(
  mpd_od_df = msoa_OD_travel2work,
  coverage_df = coverage,
  weight_by = "origin"
)

adj_raking <- adjust_raking_ratio(
  mpd_od_df = msoa_OD_travel2work,
  benchmark_od_df = census_msoa_OD_travel2work,
  max_iter = 100
)

adj_coefficient <- adjust_coefficient(
  mpd_od_df = msoa_OD_travel2work,
  benchmark_od_df = census_msoa_OD_travel2work,
  level = "origin"
)
```

```{r}
bind_rows(
  adj_inverse |> mutate(method = "inverse_penetration"),
  adj_raking |> mutate(method = "raking"),
  adj_coefficient |> mutate(method = "coefficient")
) |>
  group_by(method) |>
  summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_ratio = median(flow_adj / pmax(flow, 1), na.rm = TRUE),
    .groups = "drop"
  )
```

## Exercise

How much do estimates change, and where?
