---
title: "Adjusting Biases in OD Flows"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette demonstrates the adjustment functions in `debiasR`. The Bayesian
multilevel model is the central innovation; deterministic methods are shown as
transparent baselines and comparators.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = Inf, complete_grid = TRUE)
```

## Bayesian Multilevel Adjustment

`adjust_multilevel_bayes()` estimates adjusted OD flows with a multilevel count
model. The core idea is:

\[
\log(\mu_{ij}) =
f(X_i, X_j, d_{ij}, e_i, u)
\]

where \(X_i\) and \(X_j\) are origin and destination attributes, \(d_{ij}\) is
OD distance, \(e_i\) is the origin coverage residual or bias term, and \(u\) is
an optional hierarchical term. The adjusted flow \(F^{adj}_{ij}\) is obtained by
removing the estimated coverage-bias contribution from posterior linear
predictor draws.

The vignette uses a render-feasible square OD subset for the Bayesian model.
Set `DEBIASR_BAYES_MAX_OD_ROWS` to increase or decrease the maximum grid size.

```{r}
bayes_n_areas <- choose_bayes_area_count()
bayes_example <- load_workshop_example(
  n_areas = bayes_n_areas,
  complete_grid = TRUE
)

bayes_example$od_audit
```

```{r}
bayes_ready <- requireNamespace("rstanarm", quietly = TRUE) &&
  nrow(bayes_example$distance) > 0L

adj_bayes <- NULL
if (bayes_ready) {
  adj_bayes <- adjust_multilevel_bayes(
    mpd_od_df = bayes_example$msoa_OD_travel2work,
    coverage_df = bayes_example$coverage,
    covariates_df = bayes_example$covariates,
    distance_df = bayes_example$distance,
    prediction_scope = "complete_grid",
    random_intercept = "origin",
    model_family = "poisson",
    flow_adj_summary = "median",
    iter = 500,
    chains = 2,
    seed = 123,
    refresh = 0
  )

  attr(adj_bayes, "result_metadata")
} else {
  "Bayesian empirical rendering requires `rstanarm` and real OD distance from `debiasRdata`. The complete-grid interface is available, but the final empirical Bayesian vignette should be rendered after those inputs are present."
}
```

## Deterministic Baselines

The deterministic methods provide interpretable comparators.

### Inverse Penetration

\[
w_i = \frac{P_i}{U_i}
\]

\[
F^{adj}_{ij} = F^{mpd}_{ij} w_i
\]

or, when both origin and destination coverage are used:

\[
F^{adj}_{ij} = F^{mpd}_{ij}\sqrt{w_i w_j}
\]

### Selection Rate

The selection-rate methods use coverage and a covariate-scaled selection
parameter to construct multiplicative weights.

### Raking Ratio

Raking adjusts flows so that supplied origin and destination margins match
trusted targets:

\[
\sum_j F^{adj}_{ij} = T^{O}_i,\qquad
\sum_i F^{adj}_{ij} = T^{D}_j
\]

### Coefficient Calibration

The coefficient method estimates a global mapping from observed MPD flows to
benchmark flows:

\[
E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}
\]

## Fit Baselines

```{r}
baseline_results <- fit_deterministic_baselines(example_data)

tibble::tibble(
  method = names(baseline_results),
  rows = vapply(baseline_results, nrow, integer(1)),
  adjusted_total = vapply(
    baseline_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

```{r}
baseline_results$raking_ratio |>
  summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_weight = median(weight_ipf, na.rm = TRUE),
    ipf_converged = attr(baseline_results$raking_ratio, "ipf_converged"),
    ipf_iterations = attr(baseline_results$raking_ratio, "ipf_iterations")
  )
```

## Compare Bayesian and Baselines

If the Bayesian model is rendered, compare it against the same benchmark in
`07-validation.qmd`. If it is not rendered, the deterministic baselines still
show the complete-grid adjustment workflow and the Bayesian preconditions are
explicit.
