---
title: "Adjusting Biases in Origin-Destination Flows"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette describes the set of adjustment methods included in `debiasR` to correct mobile-phone-derived origin-destination mobility counts. It describes how each adjustment method works, associated functions and how to implement it.
All the methods in `debiasR` use mobile-phone-derived flows as input data and return an adjusted flow `flow_adj`. However, they differ in fundamental ways, in terms of data input requirements, how they derive adjustement weights and extent of bias adjustment. These differences are described in more detail in this vignette. The practical choice should be guided by what data
are available and validation against a credible benchmark. A key innovation of `debiasR` is the introduction of a Bayesian multilevel modelling approach which we explain below.


## Intuition

Let us very quickly recall the challenge we face when working with mobile-phone-derived mobility flows. These flows are typically represent an small proportion of the resident population, hence we end up under-counting these populations. So the idea of the methods included in `debiasR` is to adjust these counts to make these counts more representative of the actual observed flow. Formally, let $F^{mpd}_{ij}$ denote the observed mobile-phone-derived flow from origin $i$ to destination $j$, and let $F^{adj}_{ij}$ denote the adjusted flow. If a
method treats the observed MPD flow as under-counted, it increases
$F^{adj}_{ij}$. If it treats the observed MPD flow as over-counted, it reduces
$F^{adj}_{ij}$.

Where, let us recall:

- $F^{mpd}_{ij}$: observed MPD flow from origin $i$ to destination $j$.
- $F^{bench}_{ij}$: benchmark flow from origin $i$ to destination $j$.
- $F^{adj}_{ij}$: adjusted flow returned by an adjustment method.
- $P_i, P_j$: population at origin and destination.
- $U_i, U_j$: active-user count at origin and destination.
- $p_i = U_i/P_i$: active-user coverage rate.
- $d_{ij}$: distance between origin $i$ and destination $j$.
- $X_i, X_j$: origin and destination covariates, such as income.

## Methods In Scope

This vignette covers the adjustment functions currently available in
`debiasR`:

- `adjust_inverse_penetration()`
- `adjust_selection_rate()`
- `adjust_selection_rate2()`
- `adjust_raking_ratio()`
- `adjust_coefficient()`
- `adjust_multilevel_bayes()`

The first five methods are stable adjustment options. The multilevel function
contains the Bayesian innovation and now also exposes a fast development engine
for testing S1-S4 source/time data contracts before the full Bayesian transfer
is completed.

## Data

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
```

## Adjustment Method Summary

The table below follows the same comparison logic as the conceptual method
guide. It is meant to help users identify which method is feasible before they
run it.

```{r results='asis', echo=FALSE}
eq_stack <- function(...) {
  paste0(
    "<div class='eq-stack'>",
    paste0("<div class='eq-line'>", c(...), "</div>", collapse = ""),
    "</div>"
  )
}

method_summary <- tibble::tribble(
  ~Method, ~Rationale, ~Equation, ~`Benchmark<br>population`, ~`Active<br>users`, ~`Benchmark<br>OD flows`, ~`MPD<br>OD flows`, ~Covariates, ~Assumptions, ~Strengths, ~Limitations, ~`Validation<br>strategy`,
  "<code>adjust_inverse_penetration()</code>",
  "Inflates observed flows using population-to-user ratios where coverage is thin.",
  eq_stack(
    "<i>w</i><sub>i</sub> = P<sub>i</sub> / U<sub>i</sub>",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>i</sub>, <i>w</i><sub>j</sub>, or &radic;(<i>w</i><sub>i</sub><i>w</i><sub>j</sub>)"
  ),
  "Yes", "Yes", "No", "Yes", "No",
  "Coverage bias is roughly proportional to active-user penetration.",
  "Simple, transparent, fast, and does not require benchmark OD flows.",
  "Can overinflate sparse areas and misses compositional selection bias.",
  "Use <code>validate_flow_overall()</code> for summary fit and <code>validate_flow_residuals()</code> for large OD corrections.",

  "<code>adjust_selection_rate()</code>",
  "Uses coverage, covariates, and optional benchmark calibration to model differential selection.",
  eq_stack(
    "<i>w</i><sub>i</sub>(<i>r</i><sub>t</sub>) = 1 / (I<sub>i</sub><i>p</i><sub>i</sub> + (1 - I<sub>i</sub>)<i>r</i><sub>t</sub>)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>"
  ),
  "Yes", "Yes", "Optional", "Yes", "Yes",
  "The covariate is informative about selection and the calibration benchmark is representative when used.",
  "More flexible than inverse penetration and can use benchmark calibration.",
  "Depends on covariate quality, calibration choices, and a meaningful selection parameter.",
  "Compare calibrated and uncalibrated outputs with <code>validate_flow_overall()</code> and inspect OD residuals.",

  "<code>adjust_selection_rate2()</code>",
  "Applies a compact coverage correction with optional <code>k</code> calibration.",
  eq_stack(
    "CF(<i>p</i>; <i>k</i>) = <i>p</i> &times; (exp(-<i>k</i>) - 1) / (exp(-<i>k</i><i>p</i>) - 1)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; CF(...)"
  ),
  "Yes", "Yes", "Optional", "Yes", "No",
  "A single curvature parameter can summarize the selection correction.",
  "Parsimonious, interpretable, and lighter than a covariate-rich adjustment.",
  "Less flexible when bias varies strongly across areas.",
  "Use summary validation metrics and residual diagnostics to check whether the compact correction is adequate.",

  "<code>adjust_raking_ratio()</code>",
  "Uses iterative proportional fitting so adjusted flows match trusted origin and destination margins.",
  eq_stack(
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>",
    "&Sigma;<sub>j</sub><i>F</i><sup>adj</sup><sub>ij</sub> = O<sub>i</sub> and &Sigma;<sub>i</sub><i>F</i><sup>adj</sup><sub>ij</sub> = D<sub>j</sub>"
  ),
  "Optional as targets", "No", "Yes or user targets", "Yes", "No",
  "Trusted margins are available and the observed OD support is adequate.",
  "Guarantees consistency with supplied marginal totals.",
  "Matching margins does not guarantee correct destination allocation within each origin.",
  "Check IPF convergence and margin matching, then validate OD-pair residuals and allocation fidelity.",

  "<code>adjust_coefficient()</code>",
  "Learns a direct mapping from MPD flows to benchmark flows and applies it to the MPD table.",
  eq_stack(
    "E[<i>F</i><sup>bench</sup><sub>ij</sub>] = &beta; <i>F</i><sup>mpd</sup><sub>ij</sub>",
    "log(&mu;<sub>ij</sub>) = &alpha; + log(<i>F</i><sup>mpd</sup><sub>ij</sub>)"
  ),
  "No", "No", "Yes", "Yes", "No",
  "The MPD-benchmark relationship is stable enough to estimate on overlapping support.",
  "Direct benchmark-driven calibration with several model families.",
  "Benchmark quality is critical and a global coefficient can miss local heterogeneity.",
  "Compare families and levels with <code>validate_flow_overall()</code>; inspect systematic residuals.",

  "<code>adjust_multilevel_bayes()</code>",
  "Fits a multilevel count model with coverage bias, covariates, distance, and optional hierarchical structure.",
  eq_stack(
    "log(&mu;<sup>obs</sup><sub>ij</sub>) = &alpha; + &beta;<sub>o</sub>X<sub>i</sub> + &beta;<sub>d</sub>X<sub>j</sub> + &gamma; log(<i>d</i><sub>ij</sub>) + &omega;e<sub>i</sub> + <i>u</i>",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = summary(exp(&eta;<sup>fix</sup><sub>ij</sub> - &omega;e<sub>i</sub>))"
  ),
  "Yes", "Yes", "No", "Yes", "Yes",
  "The model family and hierarchy are well specified and coverage bias is captured by the fitted bias term.",
  "Richer structure, no benchmark OD needed for fitting, and a path to uncertainty-aware partial pooling.",
  "Heavier runtime and dependencies; S2-S4 Bayesian fitting is still being transferred from the development contract.",
  "Use model diagnostics plus the same validation functions used for other methods when a benchmark is available."
)

table_html <- knitr::kable(
  method_summary,
  format = "html",
  escape = FALSE,
  table.attr = "class='table method-summary-table'"
)

cat("<div class='method-table-wrap'>", table_html, "</div>")
```

<style>
.method-table-wrap {
  overflow-x: auto;
  margin: 1.25rem 0 1.75rem 0;
  padding: 0.6rem 0.8rem;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  background: #fff;
}

.method-summary-table {
  min-width: 1850px;
  font-size: 0.9rem;
  line-height: 1.45;
}

.method-summary-table th,
.method-summary-table td {
  vertical-align: top;
  padding: 0.75rem;
  white-space: normal;
}

.eq-stack {
  font-family: "Cambria Math", "STIX Two Text", "Times New Roman", serif;
  line-height: 1.45;
}

.eq-line {
  display: block;
}

.eq-line + .eq-line {
  margin-top: 0.25rem;
}
</style>

Two contrasts are especially important:

- `adjust_coefficient()` directly learns from benchmark flows; the benchmark is
  part of model fitting.
- `adjust_multilevel_bayes()` does not use benchmark OD flows in model fitting.
  It adjusts MPD flows using coverage, covariates, distance, and multilevel
  structure, then should be validated against a benchmark when one is available.

## Fit The Main Adjustment Methods

The non-Bayesian adjustment examples use the same empirical LAD subset loaded
above. Fitting them once makes it easier to compare output shape and totals
across methods.

```{r}
method_results <- fit_adjustment_methods(example_data)

tibble::tibble(
  method = names(method_results),
  rows = vapply(method_results, nrow, integer(1)),
  adjusted_total = vapply(
    method_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

## Method 1: `adjust_inverse_penetration()`

### Concept

If only a fraction of the population is observed in MPD, inflate flows by the
inverse of that fraction. The method can weight by origin coverage, destination
coverage, or the geometric mean of both.

### Equation

$$
p_i = \frac{U_i}{P_i}, \qquad
w_i = \frac{1}{p_i} = \frac{P_i}{U_i}
$$

Origin-weighted adjustment:

$$
F^{adj}_{ij} = w_i F^{mpd}_{ij}
$$

Destination-weighted adjustment:

$$
F^{adj}_{ij} = w_j F^{mpd}_{ij}
$$

Both-side adjustment:

$$
F^{adj}_{ij} = \sqrt{w_iw_j}F^{mpd}_{ij}
$$

### Implementation In `debiasR`

```{r}
method_results$inverse_penetration |>
  dplyr::select(origin, destination, flow, flow_adj) |>
  dplyr::slice_head(n = 5)
```

This method is a good first adjustment when the main known bias is uneven
coverage. Its limitation is that it treats coverage as the correction mechanism;
it does not model the social or spatial composition of who appears in the MPD
source.

## Method 2: `adjust_selection_rate()`

### Concept

`adjust_selection_rate()` extends coverage weighting by using an income-like
covariate and an optional benchmark calibration step. It is useful when simple
penetration weighting is too crude because selection is expected to vary with
area characteristics.

### Equation

Origin-side weight:

$$
w_i^{(O)}(r_t) =
\frac{1}{I_i p_i^{(O)} + (1 - I_i) r_t}
$$

Destination-side weight:

$$
w_j^{(D)}(r_t) =
\frac{1}{I_j p_j^{(D)} + (1 - I_j) r_t}
$$

Adjustment:

$$
F^{adj}_{ij} = F^{mpd}_{ij}w_{ij}
$$

where $I_i$ and $I_j$ are covariate values, $r_t$ is the selection-rate
parameter, and $w_{ij}$ is the origin, destination, or both-side correction.

### Implementation In `debiasR`

```{r}
method_results$selection_rate |>
  dplyr::select(origin, destination, flow, flow_adj) |>
  dplyr::slice_head(n = 5)

attr(method_results$selection_rate, "r_global")
```

This method can improve on inverse penetration when the chosen covariate is
meaningful. It requires more judgement because covariate choice, weighting side,
and calibration scale affect the output.

## Method 3: `adjust_selection_rate2()`

### Concept

`adjust_selection_rate2()` applies a compact coverage correction with an
optional calibration parameter, $k$. It is a parsimonious selection-style
alternative when users want a lighter correction than the covariate-rich
selection-rate method.

### Equation

$$
CF(p; k) =
p\frac{\exp(-k) - 1}{\exp(-kp) - 1}
$$

$$
F^{adj}_{ij} = F^{mpd}_{ij}CF(p; k)
$$

The same correction can be applied on the origin side, destination side, or both
sides depending on `weight_by`.

### Implementation In `debiasR`

```{r}
method_results$selection_rate2 |>
  dplyr::select(origin, destination, flow, flow_adj) |>
  dplyr::slice_head(n = 5)

attr(method_results$selection_rate2, "k")
```

This method is useful when coverage is the dominant concern and a compact
correction is adequate. It can underfit when bias varies strongly across places.

## Method 4: `adjust_raking_ratio()`

### Concept

`adjust_raking_ratio()` uses iterative proportional fitting so adjusted flows
match trusted origin and destination margins. This is useful when consistency
with known aggregate totals is the main requirement.

### Equation

Find $F^{adj}_{ij}$ close to $F^{mpd}_{ij}$ such that:

$$
\sum_j F^{adj}_{ij} = O_i^{bench},
\qquad
\sum_i F^{adj}_{ij} = D_j^{bench}
$$

where $O_i^{bench}$ and $D_j^{bench}$ are benchmark outflow and inflow totals.

### Implementation In `debiasR`

```{r}
method_results$raking_ratio |>
  dplyr::summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_weight = median(weight_ipf, na.rm = TRUE),
    ipf_converged = attr(method_results$raking_ratio, "ipf_converged"),
    ipf_iterations = attr(method_results$raking_ratio, "ipf_iterations")
  )
```

Raking improves marginal consistency, but matching margins does not guarantee
that the internal destination allocation is correct. Validate it at the OD-pair
level and with allocation-fidelity metrics.

## Method 5: `adjust_coefficient()`

### Concept

`adjust_coefficient()` learns a mapping from MPD flows to benchmark flows, then
applies that mapping to the MPD table. It is most appropriate when a benchmark
OD table exists and direct calibration is the main goal.

### Equation

Core proportional form:

$$
E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}
$$

For count families:

$$
\log(\mu_{ij}) = \alpha + \log(F^{mpd}_{ij})
$$

so that:

$$
\mu_{ij} = e^{\alpha}F^{mpd}_{ij},
\qquad
\beta = e^{\alpha}
$$

### Implementation In `debiasR`

```{r}
method_results$coefficient |>
  dplyr::select(origin, destination, flow, flow_adj) |>
  dplyr::slice_head(n = 5)

attr(method_results$coefficient, "coef")
```

This method is direct and easy to validate because it uses the benchmark
relationship explicitly. Its main limitation is that a global coefficient can
miss local heterogeneity unless the chosen model family and level capture the
relevant structure.

## Method 6: `adjust_multilevel_bayes()`

### Concept

`adjust_multilevel_bayes()` fits a multilevel count model to observed MPD flows
using origin and destination covariates, OD distance, an explicit coverage-bias
term, and optional hierarchical structure. It then removes the estimated
coverage-bias contribution from the model predictor to obtain adjusted flows.

The full S1-S4 Bayesian scenario implementation is still being completed. This
vignette therefore uses `model_engine = "frequentist"` as a fast placeholder for
the same data and output contract. The example is S1: one mobile-phone data
source and one observation period. The intended inferential target remains the
Bayesian engine.

### Equation

Observation model:

$$
F^{mpd}_{ij} \sim \text{Poisson}(\mu^{obs}_{ij})
$$

Observed-flow model:

$$
\log \mu^{obs}_{ij}
= \alpha + \beta_o X_i + \beta_d X_j +
\gamma \log d_{ij} + \omega e_i + u_i
$$

Coverage-bias term:

$$
e_i = 1 - \frac{U_i}{P_i}
$$

Adjusted fixed-effect predictor:

$$
\eta^{adj}_{ij} = \eta^{fix}_{ij} - \omega e_i
$$

Adjusted flow:

$$
F^{adj}_{ij} = \exp(\eta^{adj}_{ij})
$$

In the Bayesian implementation, this is done draw-by-draw from the posterior
and summarized by mean or median. The frequentist engine is used here only to
test the shared data contract, formula, and output shape quickly.

### Implementation In `debiasR`

The vignette uses a render-feasible square OD subset for the multilevel model.
Set `DEBIASR_BAYES_MAX_OD_ROWS` to increase or decrease the maximum grid size.

```{r}
bayes_n_areas <- choose_bayes_area_count()
bayes_example <- load_workshop_example(
  n_areas = bayes_n_areas,
  complete_grid = TRUE
)

bayes_example$od_audit
```

```{r}
mpd_s1 <- bayes_example$mpd_od |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

coverage_s1 <- bayes_example$coverage |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

multilevel_ready <- nrow(bayes_example$distance) > 0L

adj_multilevel <- NULL
if (multilevel_ready) {
  adj_multilevel <- adjust_multilevel_bayes(
    mpd_od_df = mpd_s1,
    coverage_df = coverage_s1,
    covariates_df = bayes_example$covariates,
    distance_df = bayes_example$distance,
    model_engine = "frequentist",
    scenario = "s1",
    source_col = "mpd_source",
    time_col = "mpd_time",
    repeated_observation = "none",
    prediction_scope = "complete_grid",
    random_intercept = "none",
    model_family = "poisson",
    flow_adj_summary = "median"
  )

  adj_multilevel |>
    dplyr::select(origin, destination, mpd_source, mpd_time, flow, flow_adj) |>
    dplyr::slice_head(n = 5)
} else {
  "The multilevel example needs OD distance inputs. The complete-grid interface is available, but this render skipped the placeholder fit because distance data were unavailable."
}
```

Inspect the metadata to confirm the scenario and formula contract.

```{r}
if (!is.null(adj_multilevel)) {
  attr(adj_multilevel, "result_metadata")[c(
    "model_engine",
    "scenario",
    "repeated_observation",
    "n_sources",
    "n_time_periods",
    "prediction_scope"
  )]
}

if (!is.null(adj_multilevel)) {
  attr(adj_multilevel, "model_terms")
}
```

To fit alternative source/time structures, keep the same function and change the
scenario columns and repeated-observation contract:

```{r}
tibble::tribble(
  ~data_structure, ~scenario, ~source_col, ~time_col, ~repeated_observation, ~typical_random_intercept,
  "S1: one source, one time", "s1", "optional constant source column", "optional constant time column", "none", "none or origin",
  "S2: one source, multiple times", "s2", "optional constant source column", "time-period column", "time", "time or origin",
  "S3: multiple sources, one time", "s3", "source column", "optional constant time column", "source", "source or origin",
  "S4: multiple sources, multiple times", "s4", "source column", "time-period column", "source_time", "source_time, source, time, or origin"
)
```

Use `model_engine = "frequentist"` for S1-S4 examples until Bayesian transfer is
complete. Use `model_engine = "bayesian"` for the existing S1 Bayesian path when
the required backend and runtime budget are available.

## How To Read Outputs

All adjustment functions return a data frame with at least `origin`,
`destination`, `flow`, and `flow_adj`. Additional columns and attributes depend
on the method. Raking stores IPF convergence metadata; selection methods store
their calibrated parameters; the multilevel method stores model, formula,
scenario, diagnostic, and prediction-scope metadata.

```{r}
validation_row(
  method_name = "inverse_penetration",
  adj_df = method_results$inverse_penetration,
  benchmark_od_df = example_data$benchmark_od
)
```

```{r}
validate_flow_pairs(
  adj_df = method_results$inverse_penetration,
  benchmark_od_df = example_data$benchmark_od
) |>
  dplyr::select(
    origin,
    destination,
    mpd_flow,
    adj_flow,
    benchmark_flow,
    diff_mpd_benchmark,
    diff_adj_benchmark
  ) |>
  dplyr::slice_head(n = 5)
```

## Practical Takeaway

No adjustment method is universally best. Start with the method whose
assumptions match the data you actually have:

- Need fast, transparent scaling from coverage inputs: use
  `adjust_inverse_penetration()` or `adjust_selection_rate2()`.
- Need a covariate-informed coverage correction: use
  `adjust_selection_rate()`.
- Need agreement with trusted origin and destination margins: use
  `adjust_raking_ratio()`.
- Need direct benchmark calibration: use `adjust_coefficient()`.
- Need a structured model for coverage bias, covariates, distance, and
  hierarchy: use `adjust_multilevel_bayes()`, validating runtime and diagnostics
  carefully.

The next vignette, `07-validation.qmd`, demonstrates how to compare adjusted OD
flows against benchmark data using complementary validation metrics.
