---
title: "Validation"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Timing

10 minutes.

## Purpose

Close the workshop by showing how to check adjusted estimates against appropriate
benchmarks and report limitations.

## Topics

- How to validate adjusted estimates
- Metrics for validation
- What benchmark data to use
- Nuances of validation

## Practical Checklist

- What is the benchmark?
- What population does the digital trace source actually cover?
- At what geography and time scale is bias measured?
- Which groups or places are underrepresented?
- What assumptions does the adjustment make?
- How should uncertainty and limitations be reported?

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work examples.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 25)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work

  raw_as_adjusted <- msoa_OD_travel2work |>
    transmute(origin, destination, flow_adj = flow)

  adj_raking <- adjust_raking_ratio(
    mpd_od_df = msoa_OD_travel2work,
    benchmark_od_df = census_msoa_OD_travel2work,
    max_iter = 100
  )
}
```

```{r}
raw_fit <- validate_flow_overall(
  adj_df = raw_as_adjusted,
  benchmark_od_df = census_msoa_OD_travel2work,
  method_name = "raw_mpd"
)

raked_fit <- validate_flow_overall(
  adj_df = adj_raking,
  benchmark_od_df = census_msoa_OD_travel2work,
  method_name = "raking"
)

bind_rows(raw_fit, raked_fit) |>
  select(method, n_pairs, pearson_r, rmse, mae, mape)
```

```{r}
validate_flow_residuals(
  adj_df = adj_raking,
  benchmark_od_df = census_msoa_OD_travel2work,
  method_name = "raking",
  top_n = 10
)$summary
```

```{r}
validate_flow_distribution(
  adj_df = adj_raking,
  benchmark_od_df = census_msoa_OD_travel2work,
  method_name = "raking",
  weight_by = "benchmark_origin_total"
)$summary
```
