---
title: "Data"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette describes the empirical data inputs used in the package examples,
their sources, their row/column structure, and the OD checks applied before
adjustment.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
```

## Sources

The empirical MPD input is `msoa_OD_travel2work`, the mobile-phone-derived MSOA
travel-to-work OD matrix from the Zenodo-derived `debiasRdata` route.

The benchmark is `census_msoa_OD_travel2work`, the matching Census 2021
`ODWP01EW` MSOA workplace-flow benchmark.

`debiasR_example_data()` normalises both sources to:

- `origin`
- `destination`
- `flow`

When `complete_grid = TRUE`, absent OD pairs are zero-filled and tagged with row
status indicators.

## Tables Returned

```{r}
tibble::tibble(
  object = names(example_data),
  class = vapply(example_data, function(x) paste(class(x), collapse = ", "), character(1)),
  rows = vapply(example_data, function(x) if (is.data.frame(x)) nrow(x) else NA_integer_, integer(1)),
  columns = vapply(example_data, function(x) if (is.data.frame(x)) ncol(x) else NA_integer_, integer(1))
)
```

## OD Audit

OD data used for method evaluation should have strict support:

- the same set of origins and destinations
- one row per OD pair
- no duplicated OD pairs
- finite nonnegative flows
- balanced total outflow and inflow at the matrix level

```{r}
example_data$od_audit
```

## MPD OD Matrix

```{r}
names(example_data$msoa_OD_travel2work)
head(example_data$msoa_OD_travel2work)
```

The MPD table includes row-status indicators when complete-grid mode is used:

- `mpd_observed`
- `mpd_zero_filled`
- `mpd_row_status`

These distinguish originally observed source rows from zero-filled OD cells.

## Benchmark OD Matrix

```{r}
names(example_data$census_msoa_OD_travel2work)
head(example_data$census_msoa_OD_travel2work)
```

The benchmark table includes analogous indicators:

- `benchmark_observed`
- `benchmark_zero_filled`
- `benchmark_row_status`

## Derived Area Tables

```{r}
head(example_data$coverage)
head(example_data$active_users)
head(example_data$population)
head(example_data$covariates)
```

The coverage table supplies $P_i$, $U_i$, $p_i$, and $b_i$ inputs for
measurement and adjustment. The covariate table supplies area-level predictors
used by diagnostics and adjustment models.

## Distance Inputs

The Bayesian model should use real OD distance when available:

```{r}
example_data$metadata |>
  select(distance_source)
```

```{r}
head(example_data$distance)
```

If `distance_source` is `not_available`, final empirical Bayesian rendering is
blocked until `debiasRdata` exposes a real OD distance table. The package can
still demonstrate the interface, but real-distance Bayesian validation should
wait for that input.
