---
title: "debiasR: Conceptual Guide to Adjustment Methods"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This short notebook explains, intuitively, how each `debiasR` adjustment method works and how it is implemented in the package.

## Setup

```{r}
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

data(simulated_mpd.od, package = "debiasR")
data(simulated_benchmark.od, package = "debiasR")
data(simulated_coverage, package = "debiasR")
data(simulated_covariates, package = "debiasR")
data(simulated_distance, package = "debiasR")
```

## Shared Intuition

All methods start from observed MPD flows (`flow`) and produce adjusted flows (`flow_adj`).

- If a method thinks MPD under-observes a flow, it increases `flow_adj`.
- If it thinks MPD over-observes a flow, it decreases `flow_adj`.

The methods differ in *how* they infer that correction.

### Notation Used

- $F^{mpd}_{ij}$: observed MPD flow from origin $i$ to destination $j$.
- $F^{bench}_{ij}$: benchmark (census-like) flow from $i$ to $j$.
- $F^{adj}_{ij}$: adjusted flow output by a method.
- $P_i, P_j$: population at origin/destination.
- $U_i, U_j$: active user count at origin/destination.
- $p_i = U_i/P_i$: coverage (penetration) rate.
- $d_{ij}$: distance between $i$ and $j$.
- $X_i, X_j$: origin/destination covariates (for example, income).

## Method Comparison Table

```{r}
method_cmp <- tibble::tribble(
  ~method, ~uses_benchmark_in_fit, ~uses_coverage, ~uses_covariates, ~uses_distance, ~uses_random_effects, ~core_calculation,
  "adjust_inverse_penetration", "No", "Yes", "No", "No", "No", "Inverse penetration weights from population/user_count (origin/destination/both).",
  "adjust_selection_rate", "Optional (for calibration)", "Yes", "Yes", "No", "No", "Selection-rate style weight combining coverage + covariates, then optional benchmark calibration.",
  "adjust_selection_rate2", "Optional (for calibration)", "Yes", "No", "No", "No", "Simplified selection-rate correction, often close to multiplicative rescaling.",
  "adjust_raking_ratio", "Yes", "No", "No", "No", "No", "IPF/raking to match benchmark marginals.",
  "adjust_coefficient", "Yes", "No", "No", "No", "No", "Regression mapping from MPD to benchmark flow (coefficient-based adjustment).",
  "adjust_multilevel_bayes", "No", "Yes", "Yes", "Yes", "Yes (configurable)", "Bayesian count model on MPD; de-biases posterior linear predictor using estimated bias components."
)

method_cmp
```

Key contrast:
- `adjust_coefficient()` directly learns from benchmark flows.
- `adjust_multilevel_bayes()` does **not** use benchmark flows in model fitting; it adjusts MPD using coverage/covariates/distance/hierarchical structure.

## Method 1: `adjust_inverse_penetration()`

### Concept

If only a fraction of the population is observed in MPD, inflate flows by the inverse of that fraction.

- Coverage at area `i`: `user_count_i / population_i`
- Inverse penetration weight: `population_i / user_count_i`

### Equation

\[
p_i = \frac{U_i}{P_i}, \quad w_i = \frac{1}{p_i} = \frac{P_i}{U_i}
\]

Origin-weighted version:
\[
F^{adj}_{ij} = w_i F^{mpd}_{ij}
\]

Destination-weighted version:
\[
F^{adj}_{ij} = w_j F^{mpd}_{ij}
\]

Both:
\[
F^{adj}_{ij} = \sqrt{w_i w_j} \, F^{mpd}_{ij}
\]

### Implementation in `debiasR`

- Function: `adjust_inverse_penetration()`
- Uses `coverage_df` columns `origin`, `population`, `user_count` (and optionally `mpd_source`).
- Applies weights by origin, destination, or both (`weight_by`).

```{r}
adj_m1 <- adjust_inverse_penetration(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "both"
)

adj_m1 |> select(origin, destination, flow, flow_adj) |> slice_head(n = 5)
```

## Method 2: `adjust_selection_rate()`

### Concept

Adjust with a selection-rate style factor that combines:

- coverage terms (population/users),
- covariates (for example, income),
- optional calibration to benchmark totals.

### Equation (conceptual form)

\[
F^{adj}_{ij} = r_{ij} \, F^{mpd}_{ij}
\]

with a selection-rate factor $r_{ij}$ built from coverage and covariates, for example:
\[
r_{ij} \propto f\!\left(\frac{P_i}{U_i}, \frac{P_j}{U_j}, X_i, X_j\right)
\]

and optional calibration:
\[
\tilde{F}^{adj}_{ij} = c_g \, F^{adj}_{ij}
\]
where $c_g$ is a calibration factor at global/origin/destination aggregation.

### Implementation in `debiasR`

- Function: `adjust_selection_rate()`
- Requires `covariates_df` and coverage inputs.
- Can calibrate globally/origin/destination using `benchmark_od_df`.

```{r}
adj_m2 <- adjust_selection_rate(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  covariates_df = simulated_covariates,
  income_col = "income_norm",
  weight_by = "origin",
  benchmark_od_df = simulated_benchmark.od,
  calibration_aggregate = "origin"
)

attr(adj_m2, "r_global")
```

## Method 3: `adjust_selection_rate2()`

### Concept

A simplified/alternative selection-rate adjustment that often behaves like a multiplicative scaling of MPD flows.

### Equation (conceptual form)

\[
F^{adj}_{ij} = k \, r^{(2)}_{ij} \, F^{mpd}_{ij}
\]

where:
- $r^{(2)}_{ij}$ is a simplified coverage-based correction,
- $k$ is an optional calibration constant estimated from benchmark aggregates.

### Implementation in `debiasR`

- Function: `adjust_selection_rate2()`
- Uses coverage-derived correction and optional calibration constant `k`.
- Often yields near-linear `flow_adj` vs `flow`.

```{r}
adj_m3 <- adjust_selection_rate2(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "origin",
  benchmark_od_df = simulated_benchmark.od,
  calibration_aggregate = "origin"
)

attr(adj_m3, "k")
```

## Method 4: `adjust_raking_ratio()`

### Concept

Use iterative proportional fitting (IPF/raking) so adjusted OD flows match known marginals (from benchmark).

### Equation (constraint form)

Find $F^{adj}_{ij}$ close to $F^{mpd}_{ij}$ such that:
\[
\sum_j F^{adj}_{ij} = O_i^{bench}, \quad
\sum_i F^{adj}_{ij} = D_j^{bench}
\]

where $O_i^{bench}$ and $D_j^{bench}$ are benchmark outflow/inflow marginals.
IPF alternates row and column scaling until constraints are met.

### Implementation in `debiasR`

- Function: `adjust_raking_ratio()`
- Reweights the OD matrix iteratively to satisfy margin constraints.
- Good when matching aggregate totals is the priority.

```{r}
adj_m4 <- adjust_raking_ratio(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od
)

list(
  ipf_converged = attr(adj_m4, "ipf_converged"),
  ipf_iterations = attr(adj_m4, "ipf_iterations")
)
```

## Method 5: `adjust_coefficient()`

### Concept

Learn a mapping from MPD flows to benchmark flows via a regression coefficient model, then apply that mapping.

### Equation (OLS-style example)

Calibration model:
\[
F^{bench}_{ij} = \alpha + \beta F^{mpd}_{ij} + \varepsilon_{ij}
\]

Adjustment map:
\[
F^{adj}_{ij} = \hat{\alpha} + \hat{\beta} F^{mpd}_{ij}
\]

Depending on options, implementation may use alternative families/forms, but the key point is direct learning from benchmark flows.

### Implementation in `debiasR`

- Function: `adjust_coefficient()`
- Supports model families (for example, OLS).
- Can fit at OD level and project to adjusted flows.

```{r}
adj_m5 <- adjust_coefficient(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od,
  model_family = "ols",
  level = "od"
)

attr(adj_m5, "coef")
```

## Method 6: `adjust_multilevel_bayes()`

### Concept

Fit a Bayesian multilevel count model to observed MPD flows:

- fixed effects for origin/destination predictors and distance,
- explicit bias term from coverage (`bias_e_origin`),
- random intercept structure (for example, origin).

Then remove estimated bias contribution from posterior linear predictors to obtain `flow_adj`.

### Equation (prototype in package)

Observation model:
\[
F^{mpd}_{ij} \sim \text{Poisson}(\mu^{obs}_{ij})
\]

Linear predictor:
\[
\log \mu^{obs}_{ij}
= \alpha + \beta_o X_i + \beta_d X_j + \gamma \log d_{ij} + \omega e_i + u_i
\]

with coverage-bias term:
\[
e_i = 1 - \frac{U_i}{P_i}
\]

Bias-adjusted latent mean:
\[
\log \mu^{true}_{ij}
= \log \mu^{obs}_{ij} - (\omega e_i + u_i)
\]

and:
\[
F^{adj}_{ij} = \exp(\log \mu^{true}_{ij})
\]

In implementation, this is done draw-by-draw from the posterior, then summarized by mean/median.

### Implementation in `debiasR`

- Function: `adjust_multilevel_bayes()`
- Backend: `rstanarm`/`brms` (depends on family/backend choice).
- Uses posterior draws and summarizes adjusted flows (mean/median).

```{r}
if (requireNamespace("rstanarm", quietly = TRUE)) {
  adj_m6 <- adjust_multilevel_bayes(
    mpd_od_df = simulated_mpd.od |> slice_sample(n = 3000),
    coverage_df = simulated_coverage,
    covariates_df = simulated_covariates,
    distance_df = simulated_distance,
    random_intercept = "origin",
    model_family = "poisson",
    flow_adj_summary = "median",
    iter = 200,
    chains = 1,
    seed = 123,
    refresh = 0
  )

  attr(adj_m6, "coefficients") |> select(any_of(c("term", "mean", "q2.5", "q97.5")))
} else {
  "Install 'rstanarm' to run this chunk."
}
```

## How to Read Outputs

For all methods:

- Primary output: tibble with `origin`, `destination`, `flow`, `flow_adj`.
- Compare methods with:
  - `validate_flow_benchmark()`
  - scatterplots (`flow_adj` vs benchmark, `flow_adj` vs MPD).

```{r}
val_m1 <- validate_flow_benchmark(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration",
  return_joined = FALSE
)

tibble::tibble(
  method = val_m1$method,
  n = val_m1$n,
  pearson_r = val_m1$pearson_r,
  rmse = val_m1$rmse,
  mae = val_m1$mae,
  r_squared = val_m1$r_squared
)
```

## Practical Takeaway

- Need fast transparent scaling: start with Method 1 or 3.
- Need margin consistency: Method 4.
- Need model-based mapping: Method 5.
- Need uncertainty + hierarchical bias structure: Method 6.
