---
title: "Measuring population coverage bias"
vignette: >
  %\VignetteIndexEntry{Measuring population coverage bias}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette shows how to measure population coverage bias in mobile-phone-derived data. The goal is simple: compare how many people are captured in the digital trace data with how many people live in each area according to a trusted benchmark.

We use the notation introduced in the previous vignette. For each Local Authority District (LAD) $i$, let:

- $U_i$ be the observed digital trace user count
- $P_i$ be the benchmark population count from the 2021 Census
- $c_i$ be the active-user coverage rate
- $b_i$ be the coverage bias

The two core quantities are:

$$
c_i = \frac{U_i}{P_i}
$$

$$
b_i = 1 - c_i
$$

So, if an area has $c_i = 0.04$, the digital trace data cover about 4 people per 100 residents, and the coverage bias is $b_i = 0.96$. Lower values of $b_i$ mean better coverage. Values above zero usually mean undercoverage relative to the Census benchmark.

## Load data

We use the LAD coverage table from `debiasRdata`. The table contains benchmark Census population counts and Locomizer-derived digital trace population counts for each LAD.

```{r setup}
library(debiasR)
library(debiasRdata)
library(dplyr)
library(ggplot2)
```

```{r}
coverage <- debiasRdata::coverage_lad

head(coverage)
```

`debiasR::measure_bias()` expects the benchmark population column to be called `population` and the digital trace population column to be called `user_count`. These names are already used in `coverage_lad`; we only rename the area code to `area` so the later residual diagnostic has a standard area identifier.

```{r}
coverage <- coverage |>
  rename(
    area = code
  )

head(coverage)
```

## Measure coverage bias

`measure_bias()` calculates the active-user coverage rate $c_i = U_i / P_i$ and the coverage bias $b_i = 1 - c_i$.

The function returns these as:

- `coverage_score`, which is $c_i$
- `coverage_bias`, which is $b_i$

```{r}
bias_lad <- debiasR::measure_bias(coverage)

bias_lad |>
  select(area, name, population, user_count, coverage_score, coverage_bias) |>
  head()
```

A useful first summary is the overall coverage across all LADs. This answers: across England and Wales as a whole, what share of the Census population is captured in the digital trace population count?

```{r}
bias_lad |>
  summarise(
    lad_count = n(),
    census_population = sum(population),
    digital_trace_population = sum(user_count),
    overall_coverage_score = sum(user_count) / sum(population),
    overall_coverage_bias = 1 - overall_coverage_score
  )
```

This overall number is useful, but it can hide important local differences. A data source may have reasonable national coverage while still covering some places much better than others.

## Compare LADs

The next step is to look at variation across LADs. The areas with the lowest values of $b_i$ have the strongest digital trace coverage. The areas with the highest values of $b_i$ have the weakest coverage.

```{r}
bias_lad |>
  arrange(coverage_bias) |>
  select(area, name, population, user_count, coverage_score, coverage_bias) |>
  head(10)
```

```{r}
bias_lad |>
  arrange(desc(coverage_bias)) |>
  select(area, name, population, user_count, coverage_score, coverage_bias) |>
  head(10)
```

A histogram gives a quick view of how uneven coverage is across LADs.

```{r}
ggplot(bias_lad, aes(x = coverage_bias)) +
  geom_histogram(bins = 30, colour = "white", fill = "#356859") +
  labs(
    x = expression("Coverage bias, " * b[i]),
    y = "Number of LADs",
    title = "Distribution of population coverage bias across LADs"
  ) +
  theme_minimal()
```

## Measure distributional bias

Coverage rates tell us whether each area is over- or under-represented relative
to its own benchmark population. A related question is whether the active-user
population is distributed across areas in the same proportions as the Census
population. For example, if an LAD contains 3% of the benchmark population, a
proportionally representative active-user dataset would also place about 3% of
active users in that LAD.

`measure_bias_distribution()` summarises this distributional comparison. It
uses the benchmark population shares as the reference distribution and active
user shares as the comparison distribution. Lower KL divergence and
Jensen-Shannon divergence mean the active-user distribution is closer to the
benchmark population distribution.

```{r}
distribution_bias <- debiasR::measure_bias_distribution(
  coverage_df = coverage,
  area_col = "area",
  population_col = "population",
  user_count_col = "user_count"
)

distribution_bias$summary
```

The area-level output shows which LADs contribute most to the difference
between the two distributions. Positive share differences mean an LAD accounts
for a larger share of active users than of the benchmark population. Negative
values mean the opposite.

```{r}
distribution_bias$area_level |>
  dplyr::left_join(
    coverage |> dplyr::select(area, name),
    by = "area"
  ) |>
  dplyr::mutate(
    abs_share_difference = abs(share_difference_user_minus_population)
  ) |>
  dplyr::arrange(dplyr::desc(abs_share_difference)) |>
  dplyr::select(
    area,
    name,
    population_share,
    user_share,
    share_difference_user_minus_population,
    kl_contribution,
    jsd_contribution
  ) |>
  head(10)
```

Before looking at residuals, it is useful to check the basic relationship
between benchmark population and observed digital trace user counts. Larger LADs
should generally have larger user counts. If they did not, the data would be a
poor population proxy from the start.

The figure below turns this check into a proportional baseline. The line shows
the user count we would expect if every LAD had the same overall active-user
coverage rate. The points do follow the line reasonably well, which is
reassuring. But the second panel also shows that some LADs sit above or below
that expectation. These departures are the beginning of the residual bias
question.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "Observed user counts broadly increase with Census population, but some LADs depart from the constant-coverage baseline."
baseline_plot_data <- bias_lad |>
  mutate(
    overall_coverage_score = sum(user_count) / sum(population),
    expected_user_count = population * overall_coverage_score,
    baseline_coverage_bias = 1 - overall_coverage_score,
    residual_bias = coverage_bias - baseline_coverage_bias
  )

baseline_line <- baseline_plot_data |>
  select(population, expected_user_count) |>
  arrange(population)

plot_baseline <- ggplot(baseline_plot_data, aes(x = population, y = user_count)) +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_line(
    data = baseline_line,
    aes(y = expected_user_count),
    colour = "#8B3A3A",
    linewidth = 1
  ) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = expression("Census population, " * P[i]),
    y = expression("Observed user count, " * U[i]),
    title = "Observed counts and baseline"
  ) +
  theme_minimal()

plot_residuals <- ggplot(baseline_plot_data, aes(x = population, y = residual_bias)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(alpha = 0.7, colour = "#356859") +
  scale_x_log10() +
  labs(
    x = expression("Census population, " * P[i]),
    y = expression("Bias residual, " * b[i] - bar(b)),
    title = "Departures from baseline"
  ) +
  theme_minimal()

grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2)))
print(
  plot_baseline,
  vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1)
)
print(
  plot_residuals,
  vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2)
)
```

## Proportional baseline

The proportional baseline is useful because it gives us a simple expectation:
if the digital trace data captured the same share of the population in every
LAD, then differences in user counts would mainly reflect differences in Census
population size.

That is a sensible first check, but it is not the same as saying the data are
representative. A dataset can scale well with population overall while still
covering some places better than others. The residuals from the baseline show
where those local departures occur.

`debiasR::validate_bias_residual_structure()` calculates this baseline and
returns expected user counts plus several residual measures for each LAD. These
include count residuals, defined as observed user counts minus the counts
expected under constant coverage, and the default `coverage_score_residual`,
defined as local coverage minus overall coverage.

```{r}
bias_residuals <- debiasR::validate_bias_residual_structure(
  coverage_df = coverage,
  coverage_area_col = "area",
  population_col = "population",
  user_count_col = "user_count"
)

```

The object returned by `validate_bias_residual_structure()` is a list. Each
component gives different information:

- `summary` gives the overall active-user coverage rate, the number of areas and summary
  measures of the residuals
- `residual_definitions` records how the different residuals are defined and how
  to interpret their sign
- `moran_i` reports spatial autocorrelation diagnostics when a neighbour table
  is supplied; here it is returned with missing values because we have not
  provided spatial neighbours
- `benchmark_flow_correlation` reports correlations between residuals and
  benchmark origin or destination flow totals when a benchmark OD-flow table is
  supplied; here it is empty because this section only uses population counts
- `covariate_correlation` reports the correlation between residuals and a
  selected area-level covariate when one is supplied; here it is returned with
  missing values because we have not yet added covariates
- `area_level` contains one row per LAD with the observed count, expected count,
  active-user coverage rate and residuals
- `map_data` contains the same area-level residuals, optionally joined to
  geometry or coordinate data when these are supplied

Some optional components may also appear if extra inputs are provided. For
example, `benchmark_flow_data` is returned when benchmark OD flows are supplied,
`covariate_data` is returned when a covariate is supplied and `plots` is returned
when `make_plots = TRUE`.

For the rest of this vignette, we focus on three components: `summary`,
`residual_definitions` and `area_level`.

```{r}
names(bias_residuals)
```

The `summary` component provides a compact description of the proportional
baseline. Here, `global_coverage_score` is the share of the total benchmark
population captured by the digital trace data across all LADs combined. This is
the constant coverage rate used to calculate the expected user count in each
LAD.

```{r}
bias_residuals$summary
```

The `residual_definitions` component is a useful reminder of the residuals
available in the output. By default, the selected residual is
`coverage_score_residual`.

```{r}
bias_residuals$residual_definitions
```

The `area_level` component is the main table used for interpretation. It keeps
the original population and user counts, adds the expected user count under the
proportional baseline and then calculates residuals for each LAD.

```{r}
bias_residuals$area_level |>
  select(
    area,
    population,
    user_count,
    expected_user_count,
    coverage_score,
    coverage_score_residual,
    selected_residual
  ) |>
  head()
```

The default residual, `coverage_score_residual`, is local coverage minus overall coverage:

$$
c_i - \bar{c}
$$

Here, $\bar{c}$ is the overall active-user coverage rate. Positive values mean the LAD is covered more than expected under the proportional baseline. Negative values mean the LAD is covered less than expected.

Because the vignette focuses on bias, we also create a bias residual:

$$
r_i = b_i - \bar{b}
$$

where $\bar{b} = 1 - \bar{c}$. Positive values of $r_i$ mean the LAD has more coverage bias than expected under the proportional baseline. Negative values mean it has less coverage bias than expected.

```{r}
residual_lad <- bias_residuals$area_level |>
  mutate(
    baseline_coverage_score = global_coverage_score,
    baseline_coverage_bias = 1 - global_coverage_score,
    residual_bias = coverage_bias - baseline_coverage_bias
  ) |>
  left_join(
    coverage |> select(area, name),
    by = "area"
  )

residual_lad |>
  select(
    area,
    name,
    population,
    user_count,
    baseline_coverage_score,
    coverage_score,
    coverage_bias,
    residual_bias
  ) |>
  head()
```

The largest positive residuals are LADs where digital trace coverage is lower than expected from the national baseline.

```{r}
residual_lad |>
  arrange(desc(residual_bias)) |>
  select(area, name, population, user_count, coverage_score, coverage_bias, residual_bias) |>
  head(10)
```

The most negative residuals are LADs where coverage is higher than expected.

```{r}
residual_lad |>
  arrange(residual_bias) |>
  select(area, name, population, user_count, coverage_score, coverage_bias, residual_bias) |>
  head(10)
```

```{r}
ggplot(residual_lad, aes(x = residual_bias)) +
  geom_vline(xintercept = 0, colour = "grey40") +
  geom_histogram(bins = 30, colour = "white", fill = "#6B5B95") +
  labs(
    x = expression("Bias residual relative to proportional baseline, " * r[i]),
    y = "Number of LADs",
    title = "Some LADs depart from the constant-coverage baseline"
  ) +
  theme_minimal()
```

## Interpretation

We have seen that population coverage bias tells us how much of the benchmark population is visible in the digital trace data. But, is this coverage roughly constant across places, or are some LADs more visible in the data than others?

In practice, we find that coverage tends to be uneven. Some LADs are overcovered or undercovered relative to the national average coverage rate. These departures raise a representativeness question: why is a particular place more or less visible than expected? One possibility is that the area contains a larger share of population groups (e.g. people aged 20-29) that are more or less likely to appear in the digital trace data.

The key message is that high overall coverage is not the same as representativeness. A data source can capture many people in total while still representing some places and population groups better than others. So, our next step is to ask whether residual coverage bias is associated with area characteristics such as age structure, socioeconomic composition or rurality.
