---
title: "Adjusting biases in origin-destination flows"
vignette: >
  %\VignetteIndexEntry{Adjusting biases in origin-destination flows}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Aim

This vignette describes the set of six adjustment methods included in
`debiasR`, to correct mobile-phone-derived origin-destination mobility
counts. `debiasR` implements five existing methods and a novel Bayesian
multilevel modelling approach. The Bayesian approach comprises a set of
functional form and model options which we explain in this and next
vignette.

This vignette describes how each adjustment method works, associated
functions and parameters, and how to implement it. The vignette focuses
on how to implement the various functions using an small sample of the
dataset introduced in the [Getting set up](v03-getting-set-up.html).
This means that the results do not reflect the real performance of the
methods. The next [Validation](v07-validation.html) vignette illustrates
the comparative performance of the various methods. All the methods in
`debiasR` use mobile-phone-derived flows as input data and return an
adjusted flow `flow_adj`.

However, the methods also differ in fundamental ways, in terms of data
input requirements, how they derive adjustment weights and the extent
they can adjust biases in the data. These differences are described in
more detail below. *A key difference* is that most existing approaches
are trained on a target benchmark set of flows. The DEBIAS Bayesian
approach does not impose this constraint as input data and this is its
biggest advantage. The lack of timely and granular benchmark flow data
from traditional data systems is normally the primary reason to use
flows from digital sources. Hence, methods relying on target benchmark
data are of limited use in live scenarios of emergency, humanitarian or
anticipatory action response.

As discussed below, the practical choice of which methods to use should
be guided by what data are available and validation against a credible
benchmark.

## Intuition

Let us quickly recall the challenge we face when working with
mobile-phone-derived mobility flows. These flows are typically represent
an small proportion of the resident population, hence we end up
under-counting these populations. Though, we can also over-count
populations if individuals have multiple mobile devices and they are
recorded in the data. So the idea of the methods included in `debiasR`
is to adjust these counts to make these counts more representative of
the actual observed flow. Formally, let $F^{mpd}_{ij}$ denote the
observed mobile-phone-derived flow from origin $i$ to destination $j$,
and let $F^{adj}_{ij}$ denote the adjusted flow. If a method treats the
observed MPD flow as under-counted, it increases $F^{adj}_{ij}$. If it
treats the observed MPD flow as over-counted, it reduces $F^{adj}_{ij}$.

Where, let us recall:

| Symbol           | Description                                             |
|:-----------------------------------|:-----------------------------------|
| $F^{mpd}_{ij}$   | Observed MPD flow from origin $i$ to destination $j$.   |
| $F^{bench}_{ij}$ | Benchmark flow from origin $i$ to destination $j$.      |
| $F^{adj}_{ij}$   | Adjusted flow returned by an adjustment method.         |
| $P_i, P_j$       | Population at origin and destination.                   |
| $U_i, U_j$       | Active-user count at origin and destination.            |
| $c_i = U_i/P_i$  | Active-user coverage rate.                              |
| $b_i = 1 - c_i$  | Active-user coverage bias.                              |
| $d_{ij}$         | Distance between origin $i$ and destination $j$.        |
| $X_i, X_j$       | Origin and destination covariates, such as rural share. |

## Methods in `debiasR`

This vignette covers the six bias adjustment methods currently available
in `debiasR`. The table below illustrates how they map onto `debiasR`
functions and relevant references that have inspire our thinking. Below
we explain how each of these methods adjust biases and how you can
implement them via `debiasR`.

| Method | Function | Reference |
|:-----------------------|:-----------------------|:-----------------------|
| Inverse Penetration Rate Weights | `adjust_inverse_penetration()` | Chi et al. (2025) |
| Selection Rate | `adjust_selection_rate()` | Chi et al. (2025) |
| Selection Rate II | `adjust_selection_rate2()` | Zagheni and Weber (2012) |
| Raking Ratio (or Iterative proportional fitting) | `adjust_raking_ratio()` | Chi et al. (2025) |
| Coefficient regression modelling | `adjust_coefficient()` | Chi et al. (2025) |
| Bayesian multilevel modelling | `adjust_multilevel_bayes()` | Rowe and Cabrera (in progress) |

## Data

We use the same empirical local authority district (LAD) travel-to-work
example introduced in earlier vignettes. The observed origin-destination
flow data comes from Locomizer-derived mobile phone data and the
benchmark origin-destination flows from the 2021 UK Census. The data are
provided through the `debiasRdata` companion package and described in
more detail in the Data vignette.

The README shows how to install `debiasR` and`debiasRdata` package from
GitHub. We start by loading those packages.

```{r load-packages}
library(debiasR)
library(debiasRdata)
library(dplyr)
```

```{r load-adjustment-helpers, include=FALSE}
helper_candidates <- c(
  "vignettes/_common.R",
  "_common.R",
  "../vignettes/_common.R",
  "../../vignettes/_common.R"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]
if (is.na(helper_file)) {
  stop(
    "Could not find vignettes/_common.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
source(helper_file)
```

First, we identify the main LAD data objects directly from
`debiasRdata`.

```{r}
OD_travel2work <- debiasRdata::lad_OD_travel2work
census_OD_travel2work <- debiasRdata::census_lad_OD_travel2work
coverage <- debiasRdata::coverage_lad |>
  dplyr::rename(area = code)
covariates <- debiasRdata::lad_covariates
centroids <- debiasRdata::lad_centroids
```

```{r}
head(OD_travel2work)
head(census_OD_travel2work)
head(coverage)
```

For the adjustment examples, we use `debiasR_example_data()` to
prepare a compact, aligned version of these data. It creates the
complete-grid OD structure and distance table used by some methods
below, which keeps the focus on how the `debiasR` functions work.

```{r load-adjustment-data}
example_data <- debiasR_example_data(
  n_areas = 25,
  complete_grid = TRUE
)

mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od
coverage_adjustment <- example_data$coverage
covariates_adjustment <- example_data$covariates
distance <- example_data$distance
```

We structure the outputs in a standard format so they are easier to
interpret and compare:

- `flow_mpd_raw`: column reporting raw, unadjusted MPD flows used as
  input.
- `flow_mpd_adjusted`: column reporting adjusted MPD-derived estimates
  returned by the adjustment method.
- `flow_benchmark`: column reporting observed Census benchmark flows
  used as a validation and input by some deterministic methods.

```{r}
benchmark_flow_lookup <- benchmark_od |>
  dplyr::select(
    origin,
    destination,
    flow_benchmark = flow
  )

show_adjustment_comparison <- function(adjusted_df,
                                       extra_cols = character(),
                                       n = 5) {
  adjusted_df |>
    dplyr::left_join(
      benchmark_flow_lookup,
      by = c("origin", "destination")
    ) |>
    dplyr::select(
      origin,
      destination,
      flow_mpd_raw = flow,
      flow_mpd_adjusted = flow_adj,
      flow_benchmark,
      dplyr::any_of(extra_cols)
    ) |>
    dplyr::slice_head(n = n)
}
```

## Inverse penetration rate

The inverse penetration rate adjusts origin-destination flows by scaling
observed mobile-phone-derived flows according to how well the population
is represented in the mobile-phone data. For each area, we calculate the
active-user coverage rate ($c_i$) as the number of observed active users
($U_i$) divided by the benchmark population ($P_i$). The inverse of this
rate becomes the adjustment weight ($w_i$). Areas with lower coverage
receive larger weights because their flows are assumed to be more
undercounted. For example, if only 5% of residents in an origin area
appear in the mobile-phone data, flows from that origin can be scaled up
by a factor of 20; that is $1 / 0.05 = 20$, so each observed flow is
treated as representing about twenty times as many people in the
benchmark population. The adjustment can use origin coverage,
destination coverage or a combination of both. This makes the method
simple and transparent, but it assumes that coverage bias is mainly
proportional to population coverage and does not directly account for
differences in who is represented among mobile-phone users. Formally:

*Penetration rate and inverse penetration rate weight*:

$$c_i = \frac{U_i}{P_i}, \qquad
w_i = \frac{1}{c_i} = \frac{P_i}{U_i}$$

*Origin-weighted adjustment*:

$$F^{adj}_{ij} = w_i F^{mpd}_{ij}$$

*Destination-weighted adjustment*:

$$F^{adj}_{ij} = w_j F^{mpd}_{ij}$$

*Both-side adjustment*:

$$F^{adj}_{ij} = \sqrt{w_iw_j}F^{mpd}_{ij}$$

### Implementation

To use `adjust_inverse_penetration()`, we input our observed
mobile-phone-derived origin-destination flows via `mpd_od_df` and
coverage data to `coverage_df` in tabular format. The coverage table
should contain the benchmark population and active-user counts needed to
calculate how strongly each area is underrepresented. The `weight_by`
argument lets you choose where the correction is applied: "origin"
adjusts flows using coverage at the origin, "destination" uses coverage
at the destination and "both" combines both sides. The result is a data
frame that preserves the original observed flow in `flow` and adds
`flow_adj`, the adjusted flow after inverse penetration weighting.

If we implement inverse penetration weighting adjusting by origins only,
we set `weight_by` to "origin" as done below. The resulting table
compares the original observed MPD flow, the
inverse-penetration-adjusted estimate and the Census benchmark flow. The
difference between the adjusted and benchmark columns gives a quick
descriptive view of the adjustment, though we assess it formally in the
validation vignette.

```{r}
adj_inverse_penetration <- adjust_inverse_penetration(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  weight_by = "origin"
)

show_adjustment_comparison(adj_inverse_penetration)
```

This method is a useful first adjustment when the main known source of
bias is uneven population coverage across areas. It is especially
helpful when we have reliable benchmark population counts, active-user
counts and benchmark origin-destination flows. In that setting, inverse
penetration weighting provides a transparent way to rescale observed MPD
flows so that areas with lower mobile-phone coverage contribute more
strongly to the adjusted flow estimates. Its main limitation is that it
treats coverage as the correction mechanism. It assumes that the
observed users in an area are broadly representative of the residents
who are missing from the data. However, it does not adjust for
representativeness bias. It cannot directly account for social,
demographic, behavioural or spatial differences in who appears in the
MPD source. If mobile-phone users are systematically different from
non-users, or if the relationship between coverage and mobility varies
across places, this method may reduce undercounting while still leaving
important selection bias in the adjusted flows.

## Selection rate

Selection rate adjusts origin-destination flows by allowing the
adjustment weight to depend on both population coverage and a covariate
that helps explain who appears in the mobile-phone-derived data. The
intuition is that two areas can have the same active-user coverage rate
but still differ in the kinds of people represented in the MPD source.
The selection-rate method therefore adds an area characteristic, such as
rural share, to the coverage correction. It can also use benchmark
origin-destination flows to calibrate the selection parameter, so that
the adjusted flows are closer to a trusted benchmark when one is
available. This makes the method more flexible than inverse penetration
weighting, but it also makes the result more dependent on the quality of
the chosen covariate and the calibration choice. Formally:

*Origin-side selection weight*:

$$w_i^{(O)}(r_t) =
\frac{1}{I_i c_i^{(O)} + (1 - I_i) r_t}$$

*Destination-side selection weight*:

$$w_j^{(D)}(r_t) =
\frac{1}{I_j c_j^{(D)} + (1 - I_j) r_t}$$

*Adjusted flow*:

$$F^{adj}_{ij} = F^{mpd}_{ij}w_{ij}$$

where $I_i$ and $I_j$ are covariate values; $r_t$ is the selection-rate
parameter; and, $w_{ij}$ is the origin, destination or both-side
correction.

### Implementation

To use `adjust_selection_rate()`, we again input the observed MPD
origin-destination flows through `mpd_od_df` and the population coverage
table through `coverage_df`. The `covariates_df` and `covariate_col`
arguments tell the function which area-level characteristic to use when
estimating differential selection. In this example, we use `rural_pct`,
the rural land-share covariate from `debiasRdata`. While all other
parameters in `adjust_selection_rate()` have defaults, we recommend
users to specify `benchmark_od_df` for a meaningful selection-rate
adjustment. This data frame corresponds to a benchmark OD table used to
calibrate the selection rate parameter `r` in the equations above.

::: {.callout-note title="What about if you do not have benchmark flow?"}
`adjust_selection_rate()` has two no-benchmark routes:

(1) you can provide `r_global` directly based on prior evidence,
    substantive knowledge or a sensitivity-analysis design.

(2) if both `benchmark_od_df` and `r_global` are omitted, `debiasR` uses
    a fallback based on the overall active-user coverage rate `U / P`.
    This makes the method executable, but the result should be
    interpreted as an exploratory adjustment rather than a calibrated
    estimate.

*Our recommendation* is to calibrate with a credible `benchmark_od_df`
if available. If no benchmark exists, we recommend treating `r_global`
as a sensitivity parameter. Report the fallback value, test a small set
of plausible alternatives and validate the adjusted flows against any
aggregate or available external evidence.
:::

To calibrate MPD flows, `calibration_aggregate` is set to origin. This
means that MDP flows are calibrated *against* origin-level benchmark
totals. This function tries each value in a grid of `r` values
(`r_grid`), adjusts the MPD flows, sums adjusted and benchmark flows by
origin, and chooses the `r_t` that minimises the absolute difference in
origin totals. `r_grid` is the set of candidate values that
`adjust_selection_rate()` tries when it needs to calibrate the global
selection parameter `r_t`. If `calibration_aggregate` is set to `od`, it
skips that origin aggregation. Instead, for each candidate `r_t`, it
compares each adjusted OD cell directly with the benchmark OD cell (as
described in the note below), and picks the `r_t` with the smallest
total OD-level error. The `weight_by` argument controls if the
correction is applied using origin coverage, destination coverage or
both.

```{r}
adj_selection_rate <- adjust_selection_rate(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  covariates_df = covariates_adjustment,
  covariate_col = "rural_pct",
  weight_by = "origin",
  benchmark_od_df = benchmark_od,
  calibration_aggregate = "origin"
)

show_adjustment_comparison(adj_selection_rate)

tibble::tibble(r_global = attr(adj_selection_rate,
                               "r_global"))
```

The table below reports the original MPD flow, selection-rate adjusted
estimate and Census benchmark flow. Because this example uses
`weight_by = "origin"`, all flows leaving a given origin are adjusted
using that origin's coverage and selected covariate value. The reported
`r_global` attribute is the calibrated selection-rate parameter chosen
by the function. This method is useful when coverage alone is too simple
and we believe that the data representativeness varies prominently with
an area characteristic. Its main limitation is that the chosen covariate
must be meaningful. If the covariate is weakly related to selection into
the MPD source, the adjustment may add complexity without reducing
coverage or selection bias. The `weight_missing` column in the
`adj_selection_rate` is a diagnostic label. If origin coverage data are
missing and/or invalid, this leads to missing and/or invalid weight
inputs.

## Selection rate II

Selection rate II adjusts origin-destination flows by applying a compact
coverage correction controlled by a single curvature parameter, $k$. The
method keeps the focus on population coverage. Instead of using the
simple inverse of the coverage rate, it transforms the coverage rate
through a nonlinear correction factor. This gives users a middle ground
between inverse penetration weighting and the fuller covariate-based
selection-rate model. When benchmark origin-destination flows are
available, `debiasR` can calibrate $k$ so that the adjusted flows better
match a trusted comparison. Formally:

*Coverage correction factor*:

$$CF(c; k) =
c\frac{\exp(-k) - 1}{\exp(-kc) - 1}$$

where $CF(c; k)$ is the coverage correction factor, $c$ is the
active-user coverage rate for the selected origin or destination area,
and $k$ is the curvature parameter that controls how strongly coverage
is transformed.

*Adjusted flow*:

$$F^{adj}_{ij} = F^{mpd}_{ij}CF(c; k)$$

The same correction can be applied by origin, destination or both,
depending on `weight_by`.

### Implementation

To use `adjust_selection_rate2()`, the function requires MPD flows and
coverage data as inputs. Input the observed MPD flows to `mpd_od_df` and
coverage data to `coverage_df`. The `weight_by` argument controls
whether the correction uses origin coverage, destination coverage or
both. You can set `k` directly if you have a chosen correction
curvature, or leave it as `NULL` and provide `benchmark_od_df` so the
function calibrates `k` over `k_grid`. As above,
`calibration_aggregate = "origin"` calibrates against origin-level
benchmark totals. The parameter `k` is the analog to `r` in
`adjust_selection_rate()`.

```{r}
adj_selection_rate2 <- adjust_selection_rate2(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  weight_by = "origin",
  benchmark_od_df = benchmark_od,
  calibration_aggregate = "origin"
)

show_adjustment_comparison(adj_selection_rate2)

tibble::tibble(k = attr(adj_selection_rate2, "k"))
```

The output table reports the observed MPD count, adjusted flow and
Census benchmark flow. In this example, the correction applied to a flow
depends on the origin area's coverage rate and calibrated $k$ value. The
reported `k` attribute shows the curvature parameter used by the
adjustment. This method is useful when coverage is the dominant concern
and a compact correction is adequate. Its main limitation is that one
parameter may be too simple when selection bias varies strongly across
places or population groups, and no benchmark data exist.

## Raking ratio

Raking ratio is also known as iterative proportional fitting (IPF).
Raking ratio adjusts origin-destination flows by iteratively rescaling
the observed MPD matrix so that its origin and destination totals match
trusted benchmark margins. The intuition is different from coverage
weighting. Instead of calculating a weight from active-user coverage,
the method asks the adjusted origin-destination matrix to satisfy known
aggregate constraints. If we trust total outflows from each origin and
total inflows to each destination, IPF updates the OD cells until the
row and column totals agree with those benchmarks. Formally:

*Margin constraints*:

$$\sum_j F^{adj}_{ij} = O_i^{bench},
\qquad
\sum_i F^{adj}_{ij} = D_j^{bench}$$

where $O_i^{bench}$ and $D_j^{bench}$ are benchmark outflow and inflow
totals.

### Implementation

To use `adjust_raking_ratio()`, the function requires MDP flows and set
of constraints as inputs. Feed the observed MPD flows to `mpd_od_df` and
either provide `benchmark_od_df` or supply explicit `origin_targets` and
`destination_targets`. We use `benchmark_od_df`, so `debiasR` derives
the target origin and destination margins from the benchmark OD table.

The `max_iter` and `tol` arguments control the iterative fitting
process. `max_iter` sets the maximum number of update cycles and `tol`
sets the convergence tolerance.

```{r}
adj_raking_ratio <- adjust_raking_ratio(
  mpd_od_df = mpd_od,
  benchmark_od_df = benchmark_od
)

show_adjustment_comparison(adj_raking_ratio)

adj_raking_ratio |>
  dplyr::summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_weight = median(weight_ipf, na.rm = TRUE),
    ipf_converged = attr(adj_raking_ratio, "ipf_converged"),
    ipf_iterations = attr(adj_raking_ratio, "ipf_iterations")
  )
```

The output table shows how individual observed MPD flows have been
converted into adjusted flows and how those adjusted flows compare with
the Census benchmark. The summary reports whether the IPF routine
converged, how many iterations it used, and how much the total flow
changed after adjustment. Raking is useful when trusted marginal totals
are available and consistency with those totals is a priority. Its main
limitation is that it requires observed benchmark origin and destination
flow data. A second limitation is that matching origin and destination
margins does not guarantee that the internal allocation between specific
origin-destination pairs is correct, so it should still be validated at
the OD-pair level.

## Coefficient regression modelling

Coefficient regression modelling adjusts origin-destination flows by
learning a direct empirical relationship between observed MPD flows and
benchmark flows. The method is useful when a trusted benchmark OD table
is available for the same origin-destination pairs. Instead of using
population coverage or margins, it fits a regression model that
estimates how MPD counts map onto benchmark counts, then applies that
fitted relationship to the MPD table. In its simplest form, this is a
single multiplicative coefficient. In count-model versions, the
relationship is estimated through a log link. Flow measured as
proportions or probability as, for example, the number of outflows over
the population at risk could be used as input, as well. Formally:

*Core proportional form*:

$$E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}$$

*Count-model form*:

$$\log(\mu_{ij}) = \alpha + \log(F^{mpd}_{ij})$$

so that:

$$\mu_{ij} = e^{\alpha}F^{mpd}_{ij},
\qquad
\beta = e^{\alpha}$$

### Implementation

To use `adjust_coefficient()`, the essential input requirements
are the MPD and benchmark flows. Input the observed MPD flows to
`mpd_od_df` and benchmark OD flows to `benchmark_od_df`. The
`model_family` argument controls the regression family used to learn the
MPD-to-benchmark relationship. This example uses `"ols"` for a simple
proportional calibration. The `level` argument controls whether the
coefficient is estimated at the OD, origin or destination level.
Additional options, such as `fit_intercept` and `by_source`, let users
decide whether to estimate an intercept or source-specific coefficients.

```{r}
adj_coefficient <- adjust_coefficient(
  mpd_od_df = mpd_od,
  benchmark_od_df = benchmark_od,
  model_family = "ols",
  level = "od"
)

show_adjustment_comparison(adj_coefficient)

attr(adj_coefficient, "coef")
```

The output table shows the original observed MPD flow, regression
adjusted estimate and Census benchmark flow. The coefficient attribute
reports the fitted scaling relationship used to convert MPD flows into
benchmark-like flows. This method is easy to validate because the
benchmark is part of the model-fitting process. Its main limitation is
that it requires benchmark flow data, and it does not adjust for
representativeness bias. A simple coefficient can miss local
heterogeneity unless the chosen model family and estimation level
capture the relevant structure.

## Bayesian multilevel modelling

Bayesian multilevel modelling in `debiasR` is a flexible modelling
framework comprising three observation-model variants. The function
`adjust_multilevel_bayes()` fits structured count models to observed MPD
flows. The meaning of the adjusted flow depends on the selected model
variant.

This section focuses on the default coverage-offset model variant. In
this model, the observed MPD flow is treated as a coverage-scaled
observation of an underlying true flow. It adjusts origin-destination
flows by fitting a structured count model to the observed MPD flows and
estimating the underlying true-flow scale. Active-user coverage enters
the model as a fixed observation offset, so the fitted model asks: what
true OD flow would have generated the MPD count we observed, given the
measured coverage? The multilevel structure can borrow information
across origins, destinations and corridors when those data structures
are available.

The function provides three observation-model variants developed within
the DEBIAS project: `coverage_offset`, `latent_two_level` and
`reduced_form`. It also provides a frequentist option for checking
formulas, covariates and source/time structures before running the
Bayesian fit. This vignette explains the model variants intuitively,
then implements only the coverage-offset Bayesian model
variant. The [advanced Bayesian adjustment
vignette](v09-advanced-bayesian-adjustment.html) explains the remaining
model variants and wider set of parameters in more detail.

The code below shows the model call and the main returned columns for the
coverage-offset specification. The function also provides a frequentist
option, `model_engine = "frequentist"`. Use this option to test different
model specifications, covariates, source/time structures, random effects
and distributional families before Bayesian sampling. It is useful for
specification checks and method comparison, but it does not replace the
Bayesian fit because it does not provide posterior uncertainty. For
applied Bayesian analysis, run enough iterations and chains for your data
and check the sampler diagnostics before interpreting posterior
summaries.

::: {.callout-tip title="How to make the Bayesian comparison fair"}
The proposed Bayesian models should not be assessed only as an in-sample
alternative to methods that rely on benchmark OD flows during fitting.
Raking ratio and coefficient regression can perform very well when the
validation benchmark is also part of the adjustment information, and
provide deterministic flow estimates ignoring their uncertainty.
However, benchmark flow data are rarely available when natural hazards,
armed conflict or health emergencies strike, and flow estimates from
these methods are imprecise. Yet lack of uncertainty measures prevent
from evaluating the extent of their precision.

The proposed Bayesian variants are specifically designed to be deployed
in these contexts. They do not require benchmark flow data for training
and enable measuring the uncertainty of estimates. They fit MPD flows,
active-user coverage, covariates and distance. Benchmark OD flows can be
used for external validation but, again, not required for training. The
performance of the Bayesian variants should not only be assessed on
better error performance over benchmark-calibrated methods. Additional
dimensions of assessment are needed, such as model convergence, flow
estimate improvement over raw MPD flows, measures of uncertainty and
flexibility to accommodate multiple MPD data structures from different
sources, time granularity and frequency.
:::

A practical guide:

```{r echo=FALSE}
bayesian_option_guide <- tibble::tribble(
  ~`Model variant`, ~`Use it when`, ~`Returned scale`, ~`Intuition`,
  "`observation_model = \"coverage_offset\"`",
  "Estimate real population flows from phone counts, accounting for how well each area is represented in the mobile-phone data.",
  "True-flow scale",
  "The model estimates Census-scale true flows while treating source/time coverage as a fixed observation offset.",
  "`observation_model = \"reduced_form\"`",
  "Adjust the observed phone counts directly, instead of using coverage rates to estimate how many people those phone counts represent.",
  "MPD-scale counterfactual",
  "The model fits observed MPD flows directly and predicts a counterfactual MPD flow with the fitted bias term set to zero.",
  "`observation_model = \"latent_two_level\"`",
  "Estimate shared population flows when the same OD pairs are observed more than once, such as across different data sources or time periods.",
  "Latent true-flow scale",
  "The model estimates source-invariant latent OD or OD-time true-flow states and treats observed source/time rows as noisy coverage-scaled measurements.",
  "`model_engine = \"frequentist\"`",
  "Run the frequentist option to check that the data, formulas and assumptions work before running the full Bayesian model.",
  "Frequentist option",
  "This option uses the same non-latent input structure for rapid specification checks and method comparison."
)

knitr::kable(bayesian_option_guide)
```

For routine adjustment, the practical default is:

| Target            | Recommended setting                     |
|:------------------|:----------------------------------------|
| Adjustment scale  | `target_scale = "true_flow"`            |
| Observation model | `observation_model = "coverage_offset"` |
| Coverage offset   | `coverage_scale = "origin"`             |

DEBIAS' Bayesian approach is a structured count multilevel model. The
default model option is the coverage-offset model variant. It can be
understood as two linked parts: an **MPD observation model** and a
**true-flow model**. The observation model describes why the MPD count
is only a partial view of mobility, reflecting that only a fraction of
the population appears in the MPD data. The true-flow model describes
the underlying population mobility pattern we want to estimate. This is
the part that can be specified in a multilevel structure to borrow
information across origins, destinations, OD corridors, sources or time
periods. DEBIAS offers two other model variants `reduced_form` and
`latent_two_level`. They use different observation structures and are
explained in the advanced Bayesian vignette.

*Equation 1: MPD observation model.*

$$F^{mpd}_{ij} \sim \text{Poisson}(q_{ij} F^{true}_{ij})$$

$$\log E(F^{mpd}_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}$$

where $F^{mpd}{ij}$ is the observed MPD flow from origin $i$ to
destination $j$, and $q{ij}$ is the observation probability derived from
active-user coverage. With coverage_scale = "origin",
$q_{ij} = c_i = U_i/P_i$. The term $\log(q_{ij})$ enters as an offset.
This means coverage is treated as known information about the
observation process. It is a parameter to be estimated. Intuitively, if
active-user coverage in an origin is 20%, we do not expect the MPD data
to observe all trips from that origin. Before allowing for random count
variation, the MPD count is expected to be about one fifth of the
underlying population flow. The offset tells the model how much of the
true flow the MPD system is expected to see.

*Equation 2: True-flow model.*

$$\eta^{true}_{ij}
= \alpha + \beta_o X_i + \beta_d X_j + \gamma \log d_{ij}
+ u_i + v_j + w_{ij} + \xi_{ij}$$

where $X_i$ and $X_j$ are origin and destination covariates, $d_{ij}$ is
the origin-destination distance, and $u_i$, $v_j$ and $w_{ij}$ capture
origin, destination and origin-destination structure. The true-flow
model seeks to capture the mobility patterns behind the observed MPD
counts. Flows may also share persistent structure. They may be higher
from some types of origins, lower to some types of destinations, and
lower over longer distances.

This model incorporates the multilevel structure. It enables pooling
information across related places. An origin effect lets flows from the
same origin share information. A destination effect does the same for
flows arriving at the same destination. A corridor effect captures
persistent structure between a specific origin and destination. These
terms are useful when the data contain enough repeated information for
their estimation.

The estimated true flow $F^{true}_{ij}$ is thus the prediction from the
true-flow equation. Random effects can be included and also contribute
to the prediction. In the S1 example below, we set
`random_intercept = "none"` because each corridor appears once, so the
example is a fixed-effect coverage-offset fit. The same model variant
becomes explicitly multilevel when the formula or arguments add origin,
destination, OD, source or time pooling structure. We illustrate
examples of more complex structures, including random intercepts, in the
next vignette.

The Bayesian model adjusts MPD flows by producing predictions on two
scales. The first is the **MPD-observed scale**, which keeps the
coverage offset:

$$F^{mpd,pred}_{ij} = \exp(\log(q_{ij}) + \eta^{true}_{ij})$$

This is the modelled MPD count which is the number of MPD observations
we would expect to see for an OD flow at the observed coverage level. It
remains on the device or active-user scale. This scale is useful to
assess if the model can reproduce the MPD data, but it is not the
adjusted population estimate.

The second is the **population or true-flow scale**, which removes the
coverage offset:

$$F^{adj}_{ij} = F^{true,pred}_{ij} = \exp(\eta^{true}_{ij})$$

This is the adjusted flow returned by the model when
`target_scale = "true_flow"`. It addresses the question: what
population-level flow is consistent with the observed MPD count,
active-user coverage rate, covariates and model structure? The model
moves from MPD observations to people by accounting for the fraction of
the population visible in the MPD data:

$$\text{population flow} \approx
\frac{\text{MPD-observed flow}}{\text{active-user coverage rate}}$$

The full Bayesian model combines the observed MPD count, coverage rate,
covariates, distance and any multilevel terms to estimate a plausible
population flow for each OD pair. This stabilises the adjustment when
some OD flows are small or noisy.

Both predictions are computed for every posterior draw and then
summarised with `flow_adj_summary`, the posterior median or posterior
mean. The column `flow_mpd_pred` reports the posterior summary on the
MPD-observed scale. The column `flow_adj` is the main adjusted estimate.
With `target_scale = "true_flow"`, it reports the population-scale
prediction.

::: {.callout-note title="Coverage-offset is not coefficient regression"}
The coefficient regression and DEBIAS coverage-offset model can both
return Census-scale adjusted flows, but they use different information
to get there. Coefficient regression learns a direct calibration from an
observed benchmark relationship. A simple version is:

$$E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}$$

or, on a log scale:

$$\log E(F^{bench}_{ij}) = \alpha + \log(F^{mpd}_{ij})$$

The benchmark flow is therefore part of the fitting process. The
estimated coefficient tells the model how to transform MPD flows so they
look more like the benchmark flows used for calibration.

The coverage-offset model instead starts from the observation process.
It says that the MPD count is a partial observation of an unobserved
true flow:

$$F^{mpd}_{ij} \sim \mathrm{Poisson}(q_{ij}F^{true}_{ij})$$

Here $q_{ij}$ is fixed from active-user coverage. The model estimates
$F^{true}_{ij}$ by fitting the MPD observation model and then removing
the coverage offset. A benchmark OD table is useful for validation, but
it is not required to fit the coverage-offset model.
:::

### Implementation

To use `adjust_multilevel_bayes()`, input the observed MPD flows to
`mpd_od_df`, population coverage to `coverage_df`, area-level covariates
to `covariates_df` and OD distances to `distance_df`. The example below
uses a single source and a single observation period. The
`prediction_scope` argument controls if the method adjusts only observed
MPD pairs or predicts over a complete OD matrix.

```{r echo=FALSE}
bayesian_input_guide <- tibble::tribble(
  ~Input, ~What_it_supplies,
  "`mpd_od_df`", "Observed MPD origin-destination counts.",
  "`coverage_df`", "Active-user coverage rates used to build the observation offset.",
  "`covariates_df`", "Origin and destination predictors for the true-flow model.",
  "`distance_df`", "OD distances or distance-derived predictors."
)

knitr::kable(
  bayesian_input_guide,
  col.names = c("Input", "What it supplies")
)
```

For the coverage-offset model variant, most users only need a small set
of core arguments.

```{r echo=FALSE}
bayesian_parameter_guide <- tibble::tribble(
  ~Argument, ~Description,
  "`mobility_formula`", "Defines the true-flow predictor, such as origin covariates, destination covariates, distance and optional multilevel terms.",
  "`target_scale = \"true_flow\"`", "Returns adjusted flows on the estimated true-flow scale rather than on the MPD observation scale.",
  "`observation_model = \"coverage_offset\"`", "Treats MPD flows as coverage-scaled observations of true flows.",
  "`coverage_scale`", "Chooses whether the fixed coverage offset uses origin coverage, destination coverage or both.",
  "`model_engine = \"bayesian\"`", "Fits the posterior model. Use the frequentist option, `model_engine = \"frequentist\"`, first when you want a specification check.",
  "`scenario`, `source_col`, `time_col`", "Declare the source/time data structure. The example below uses S1: one source and one time period.",
  "`prediction_scope`", "Chooses whether predictions are returned only for observed MPD rows or for a supplied complete OD matrix.",
  "`random_intercept`", "Adds or omits origin, destination, OD, source or time pooling structure when the data support it.",
  "`model_family`", "Chooses the count likelihood family, such as Poisson or negative binomial.",
  "`iter`, `chains`, `seed`", "Control Bayesian sampling and reproducibility."
)

knitr::kable(bayesian_parameter_guide)
```

As stated above, the coverage-offset model equations translate into two
components. `mobility_formula` defines the true-flow model. Here we use
`rural_pct_o`, `rural_pct_d` and `log_distance` to represent origin
covariates, destination covariates and distance. This example uses a
fixed-effect intercept, but the model can accommodate multilevel terms
such as `(1 | origin)` or `(1 | destination)`. The recommended mode for
the observation model is `target_scale = "true_flow"` and
`observation_model = "coverage_offset"`. The `coverage_scale` selects if
the offset uses origin, destination or geometric-mean coverage.

The wider parameter set, including `latent_flow_unit`, latent-prior
controls and the `reduced_form` compatibility model variant, is covered
in the advanced Bayesian vignette. Use these model structures when you
would like to model datasets with repeated source/time observations, or
compare model variants.

The example below uses a Bayesian fixed-effect true-flow model with an
origin coverage offset.

```{r eval=FALSE}
mpd_s1 <- mpd_od |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

coverage_s1 <- coverage_adjustment |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

adj_multilevel <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates_adjustment,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median",
  include_flow_adj_draws = TRUE,
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```

The table below reports the main outputs from
`adjust_multilevel_bayes()`, along with `flow_benchmark` reporting the
benchmark flows for comparison.

```{r echo=FALSE}
bayesian_example_file <- system.file(
  "extdata",
  "v06-bayesian-s1-comparison.csv",
  package = "debiasR"
)

if (identical(bayesian_example_file, "")) {
  stop(
    "Missing Bayesian example output."
  )
}

bayesian_example <- tibble::as_tibble(
  utils::read.csv(bayesian_example_file, check.names = FALSE)
)

knitr::kable(bayesian_example, digits = 2)
```

In the output summary, `flow` is the observed MPD count and `flow_adj`
is the main adjusted flow estimate. The function call sets
`flow_adj_summary = "median"`, so `flow_adj` equals `flow_adj_median` in
this example. The columns `flow_adj_median` and `flow_adj_mean` report
two posterior summaries of the estimated population flow. The column
`flow_mpd_pred` reports the fitted value on the MPD-observed scale,
after applying the coverage offset. The full returned object also
includes scale-specific and interval columns such as `flow_true_pred`,
`flow_mpd_pred_median`, `flow_mpd_pred_mean`, `flow_adj_q2.5`,
`flow_adj_q97.5`, `flow_mpd_pred_q2.5` and `flow_mpd_pred_q97.5`. In
this example, `flow_true_pred` is equal to `flow_adj`, so the displayed
table keeps `flow_adj` as the single main population-scale adjusted
estimate.

The example uses moderate sampler settings for illustrative purposes.
For applied analysis, increase the number of iterations and chains.

*Additional outputs and diagnostics*

You can inspect the metadata and terms to confirm the scenario and
fitted formula. Inspect sampler diagnostics before interpreting the
posterior summaries.

```{r eval=FALSE}
attr(adj_multilevel, "result_metadata")
attr(adj_multilevel, "model_terms")
attr(adj_multilevel, "diagnostics")
```

```{r echo=FALSE}
bayesian_metadata_file <- system.file(
  "extdata",
  "v06-bayesian-s1-metadata.csv",
  package = "debiasR"
)

bayesian_terms_file <- system.file(
  "extdata",
  "v06-bayesian-s1-model-terms.csv",
  package = "debiasR"
)

if (identical(bayesian_metadata_file, "") || identical(bayesian_terms_file, "")) {
  stop(
    "Missing Bayesian example metadata."
  )
}

bayesian_metadata <- tibble::as_tibble(
  utils::read.csv(bayesian_metadata_file, check.names = FALSE)
)

bayesian_terms <- tibble::as_tibble(
  utils::read.csv(bayesian_terms_file, check.names = FALSE)
)

knitr::kable(bayesian_metadata)
knitr::kable(bayesian_terms)
```

The metadata report the engine, target scale, observation model,
coverage offset, resolved scenario, repeated-observation structure,
prediction scope and model terms used to fit the adjustment model. They
show the data source and temporal structure used (S1) i.e. one MPD
source and one observation period. In the full R object, `flow` stores
the observed MPD count and `flow_mpd_pred` stores the fitted MPD-scale
prediction. The practical adjusted-flow column is `flow_adj`. When
`target_scale = "true_flow"`, `flow_adj` is the estimated population
flow. The object also keeps `flow_true_pred` as a scale-specific
Bayesian prediction column, but in this setting it is equal to
`flow_adj`.

*More complex Bayesian structures*

The function `adjust_multilevel_bayes()` can accommodate more complex
S2-S4 source/time structures through `scenario`, `source_col`,
`time_col` and `repeated_observation`. For repeated source/time data,
users can select `observation_model = "latent_two_level"`. This model
structure estimates one shared latent true-flow state for an OD pair or
OD-time group and treats each source/time MPD row as an observation of
that state. This is most useful for S3 and S4 structures, where multiple
MPD sources observe the same underlying mobility state. Source and time
effects live in the observation model, so `flow_true_pred` is
source-invariant within a latent state, while `flow_mpd_pred` can vary
by source and time. See the [advanced Bayesian adjustment
vignette](v09-advanced-bayesian-adjustment.html) for more details.

## Fit all methods

We can also run all six adjustment methods at once using
`adjust_all_methods()`. The function uses the frequentist option for the
multilevel method by default, which compares the main adjustment methods
without posterior sampling. You can switch `multilevel_engine` to
`"bayesian"` for more complex structures.

```{r}
method_results <- adjust_all_methods(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  benchmark_od_df = benchmark_od,
  covariates_df = covariates_adjustment,
  distance_df = distance,
  covariate_col = "rural_pct",
  multilevel_engine = "frequentist"
)
```

```{r echo=FALSE}
method_engine <- function(x) {
  engine <- attr(x, "model_engine", exact = TRUE)
  if (is.null(engine)) "deterministic" else engine
}

tibble::tibble(
  method = names(method_results),
  engine = vapply(method_results, method_engine, character(1)),
  rows = vapply(method_results, nrow, integer(1)),
  adjusted_total = vapply(
    method_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

## Method summary

As described above, each of the adjustment methods implemented via
`debiasR` has different assumptions, input data requirements, strengths
and weaknesses. The table below offers a summary which can be used as a
conceptual guide helping users identify which method is more appropriate
given their purpose and constraints.

```{r results='asis', echo=FALSE}
eq_stack <- function(...) {
  paste0(
    "<div class='eq-stack'>",
    paste0("<div class='eq-line'>", c(...), "</div>", collapse = ""),
    "</div>"
  )
}

method_summary <- tibble::tribble(
  ~Method, ~`Correction<br>level`, ~`How the correction<br>is applied`, ~Rationale, ~Equation, ~`Benchmark<br>population`, ~`Active<br>users`, ~`Benchmark<br>OD flows`, ~Covariates, ~Strengths, ~Limitations,
  "Inverse Penetration Rate Weights",
  "Area-specific coverage correction",
  "Applies origin, destination, or both-side coverage weights using <code>weight_by</code>.",
  "Inflates observed flows using population-to-user ratios where coverage is thin.",
  eq_stack(
    "<i>w</i><sub>i</sub> = P<sub>i</sub> / U<sub>i</sub>",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>i</sub>, <i>w</i><sub>j</sub>, or &radic;(<i>w</i><sub>i</sub><i>w</i><sub>j</sub>)"
  ),
  "Yes", "Yes", "No", "No",
  "Simple, transparent, fast, and does not require benchmark OD flows.",
  "Can overinflate sparse areas and misses compositional selection bias.",

  "Selection Rate",
  "Area-specific selection correction",
  "Applies origin, destination, or both-side selection weights controlled by a global <i>r</i><sub>t</sub> parameter.",
  "Uses coverage, covariates, and optional benchmark calibration to model differential selection.",
  eq_stack(
    "<i>w</i><sub>i</sub>(<i>r</i><sub>t</sub>) = 1 / (I<sub>i</sub><i>c</i><sub>i</sub> + (1 - I<sub>i</sub>)<i>r</i><sub>t</sub>)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>"
  ),
  "Yes", "Yes", "Optional", "Yes",
  "More flexible than inverse penetration and can use benchmark calibration.",
  "Depends on covariate quality, calibration choices, and a meaningful selection parameter.",

  "Selection Rate II",
  "Area-specific nonlinear coverage correction",
  "Applies origin, destination, or both-side coverage correction controlled by a global <code>k</code> parameter.",
  "Applies a compact coverage correction with optional <code>k</code> calibration.",
  eq_stack(
    "CF(<i>c</i>; <i>k</i>) = <i>c</i> &times; (exp(-<i>k</i>) - 1) / (exp(-<i>k</i><i>c</i>) - 1)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; CF(...)"
  ),
  "Yes", "Yes", "Optional", "No",
  "Parsimonious, interpretable, and lighter than a covariate-rich adjustment.",
  "Less flexible when bias varies strongly across areas.",

  "Raking Ratio (or Iterative proportional fitting)",
  "Margin-constrained correction",
  "Iteratively rescales OD cells so adjusted origin and/or destination totals match supplied margins.",
  "Uses iterative proportional fitting so adjusted flows match trusted origin and destination margins.",
  eq_stack(
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>",
    "&Sigma;<sub>j</sub><i>F</i><sup>adj</sup><sub>ij</sub> = O<sub>i</sub> and &Sigma;<sub>i</sub><i>F</i><sup>adj</sup><sub>ij</sub> = D<sub>j</sub>"
  ),
  "Optional as targets", "No", "Yes or user targets", "No",
  "Guarantees consistency with supplied marginal totals.",
  "Matching margins does not guarantee correct destination allocation within each origin.",

  "Coefficient regression modelling",
  "Global coefficient correction",
  "Estimates a multiplicative coefficient from MPD-benchmark overlap and applies it to the MPD table.",
  "Learns a direct mapping from MPD flows to benchmark flows and applies it to the MPD table.",
  eq_stack(
    "E[<i>F</i><sup>bench</sup><sub>ij</sub>] = &beta; <i>F</i><sup>mpd</sup><sub>ij</sub>",
    "log(&mu;<sub>ij</sub>) = &alpha; + log(<i>F</i><sup>mpd</sup><sub>ij</sub>)"
  ),
  "No", "No", "Yes", "No",
  "Direct benchmark-driven calibration with several model families.",
  "Benchmark quality is critical and a global coefficient can miss local heterogeneity.",

  "Bayesian multilevel modelling",
  "Model-based corridor correction",
  "Predicts OD-specific adjusted flows from a true-flow formula and a coverage-offset observation model.",
  "Fits a count model where active-user coverage enters as a fixed observation-process offset, then returns the estimated true-flow scale.",
  eq_stack(
    "log E(<i>F</i><sup>mpd</sup><sub>ij</sub>) = log(<i>q</i><sub>ij</sub>) + &eta;<sup>true</sup><sub>ij</sub>",
    "&eta;<sup>true</sup><sub>ij</sub> = &alpha; + X<sub>i</sub>&beta;<sub>o</sub> + X<sub>j</sub>&beta;<sub>d</sub> + &gamma;log(<i>d</i><sub>ij</sub>)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = exp(&eta;<sup>true</sup><sub>ij</sub>)"
  ),
  "Yes", "Yes", "No", "Yes",
  "Richer structure, no benchmark OD needed for fitting, posterior uncertainty, and a path to source/time-aware partial pooling.",
  "Heavier runtime and dependencies; posterior fitting requires sampler diagnostics and should be compared externally against benchmark-calibrated methods."
)

table_html <- knitr::kable(
  method_summary,
  format = "html",
  escape = FALSE,
  table.attr = "class='table method-summary-table'"
)

cat("<div class='method-table-wrap'>", table_html, "</div>")
```

```{=html}
<style>
main table {
  font-size: calc(0.9rem - 1.5px);
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}

.method-table-wrap {
  overflow-x: auto;
  margin: 1.25rem 0 1.75rem 0;
  padding: 0.6rem 0.8rem;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  background: #fff;
}

.method-summary-table {
  min-width: 1650px;
}

.method-summary-table th,
.method-summary-table td {
  padding: 0.75rem;
  white-space: normal;
}

.eq-stack {
  font-family: "Cambria Math", "STIX Two Text", "Times New Roman", serif;
  line-height: 1.45;
}

.eq-line {
  display: block;
}

.eq-line + .eq-line {
  margin-top: 0.25rem;
}
</style>
```
