---
title: "Validation"
description: >
  Compare adjusted mobile-phone-derived origin-destination flows with benchmark
  flows across overall, marginal, OD-pair, distributional allocation, and
  residual-structure diagnostics.
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
.template-quarto table {
  font-size: 14.5px;
}
.validation-table-small,
.validation-table-small table,
table.validation-table-small,
.template-quarto .validation-table-small {
  font-size: 13px;
}
.validation-table-compact,
.validation-table-compact table,
table.validation-table-compact,
.template-quarto .validation-table-compact {
  font-size: 13.5px;
}
</style>
```

## Aim

This vignette describes how to validate adjusted mobile-phone-derived
origin-destination (OD) flows against a benchmark OD table. The aim is simple:
after an adjustment method has been applied, we want to know whether the
adjusted flows are closer to the benchmark flows than the raw mobile-phone
flows were.

The key comparison is adjusted *versus* benchmark flows. This tells us how close
the final adjusted output is to the target benchmark. We also keep two secondary
comparisons in view:

- raw mobile-phone-derived *versus* benchmark flows, to show the starting level
  of bias before adjustment.
- raw mobile-phone-derived *versus* adjusted flows, to show how much each method
  moved the original data.

A central requirement in this vignette is fairness. All methods are assessed
using the same validation data: the same raw OD table, the same benchmark OD
table, and the same full LAD origin-destination matrix. When a method needs
additional fitting inputs, such as coverage rates, distance or covariates,
those inputs are taken from the same prepared example data.
LAD means local authority district, a local government geography used in the
example data.

The example compares deterministic adjustment methods that can be fitted live
on the full LAD matrix during vignette rendering. These include simple coverage
weighting, benchmark-assisted calibration and benchmark-cell calibration. Some
methods use benchmark totals or benchmark OD cells during fitting, so the
validation result should be read alongside the fitting inputs table rather than
as a single winner-takes-all ranking.

## Intuition

A practical validation strategy is to move from broad summaries to more local
diagnostics. This helps us avoid relying on one number. A method can look good
overall and still perform poorly for some origins, destinations or individual
OD pairs.

This vignette uses five levels.

**Level 1** summarises performance across the whole OD system. This tells us
whether a method is broadly close to the benchmark.

**Level 2** compares origin and destination totals. This checks whether adjusted
flows reproduce total outflows from each origin and total inflows to each
destination.

**Level 3** compares individual OD pairs. This shows where large local errors
remain after adjustment.

**Level 4** assesses whether each origin allocates its flows across destinations
in a similar way to the benchmark. This is about destination shares within
origins, not only total scale.

**Level 5** checks whether remaining residuals still show structure after
adjustment. For example, residuals may be larger for large benchmark flows,
clustered across neighbouring areas, or associated with an area covariate.

At each level, it is useful to separate error metrics from correlation metrics.
Error metrics tell us whether the numbers are close in size. Correlation
metrics tell us whether the broad pattern is similar. Both matter. A method can
have a high correlation with the benchmark and still be too high or too low in
absolute terms.

## Validation measures

Let $F^{adj}_{ij}$ be the adjusted flow from origin $i$ to destination $j$, and
let $F^{bench}_{ij}$ be the benchmark flow for the same OD pair. In the
adjusted-versus-benchmark comparison used by `debiasR`, the signed residual is
defined as:

$$
e_{ij} = F^{bench}_{ij} - F^{adj}_{ij}.
$$

Positive residuals mean the benchmark flow is larger than the adjusted flow.
Negative residuals mean the adjusted flow is larger than the benchmark. The
same sign convention is used in the residual tables below.[^validation-measure-reuse]

::: {.validation-table-small}
| Metric | Description | Equation |
|:--|:--|:--|
| Mean error | Average signed error; shows whether flows are underestimated or overestimated on average. | $\frac{1}{n}\sum_{ij} e_{ij}$ |
| Mean absolute error | Average absolute error; shows the typical size of the error, ignoring sign. | $\frac{1}{n}\sum_{ij} \lvert e_{ij} \rvert$ |
| Root mean squared error | Error metric that gives more weight to large errors. | $\sqrt{\frac{1}{n}\sum_{ij} e_{ij}^{2}}$ |
| Mean absolute percentage error | Average absolute error as a percentage of the benchmark flow. | $\frac{100}{n}\sum_{ij}\lvert \frac{e_{ij}}{F^{bench}_{ij}} \rvert$ |
| Pearson's r | Linear correlation; shows whether large benchmark flows also tend to be large adjusted flows. | $\mathrm{cor}(F^{adj}_{ij}, F^{bench}_{ij})$ |
| Spearman's rho | Rank correlation; shows whether the ordering of flows is similar. | $\mathrm{cor}(\mathrm{rank}(F^{adj}_{ij}), \mathrm{rank}(F^{bench}_{ij}))$ |
| Jensen-Shannon divergence | Level 4 metric. Symmetric difference between origin-conditioned destination-share distributions. Lower values mean closer allocation. | $\frac{1}{2} KL(p^{ref}\mid\mid m) + \frac{1}{2} KL(p^{comp}\mid\mid m)$ |
| Moran's I | Level 5 diagnostic. Spatial autocorrelation in area-level residuals using neighbour links $w_{ik}$. Positive values suggest neighbouring areas have similar residuals. | $\frac{n}{W}\frac{\sum_i\sum_k w_{ik}(\bar e_i-\bar e)(\bar e_k-\bar e)}{\sum_i(\bar e_i-\bar e)^2}$ |
:::

[^validation-measure-reuse]: The equations are written for
    adjusted-versus-benchmark flows. The same formulas can be reused for
    raw-versus-benchmark or raw-versus-adjusted comparisons by replacing
    $F^{adj}_{ij}$ and $F^{bench}_{ij}$ with the two flow columns being
    compared. For example, raw-versus-benchmark uses raw MPD flows and
    benchmark flows.

## Getting ready

We start by loading the packages used in this vignette.

```{r load-packages}
library(debiasR)
library(debiasRdata)
library(dplyr)
library(ggplot2)
```

```{r load-validation-helpers, include=FALSE}
helper_candidates <- c(
  "vignettes/_validation_helpers.R",
  "_validation_helpers.R",
  "../vignettes/_validation_helpers.R",
  "../../vignettes/_validation_helpers.R"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]
if (is.na(helper_file)) {
  stop(
    "Could not find vignettes/_validation_helpers.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
source(helper_file)
```

We use the same empirical LAD travel-to-work example introduced in the earlier
vignettes. `debiasR_example_data()` prepares an aligned full LAD
origin-destination matrix. It returns raw MPD flows, benchmark flows, coverage
rates, covariates and distances for the same LAD areas.

```{r load-validation-data}
example_data <- debiasR_example_data(
  n_areas = Inf,
  complete_grid = TRUE,
  geography = "lad"
)

mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od
coverage_validation <- example_data$coverage
covariates_validation <- example_data$covariates
distance_validation <- example_data$distance

area_names <- covariates_validation |>
  dplyr::select(area, name)
```

The validation data contain `r example_data$metadata$n_areas` LADs and
`r example_data$metadata$n_mpd_od` OD rows.

The table below states which inputs each method uses. This matters because
methods do not all learn from the same information. We still validate all
outputs on the same benchmark OD table and the same OD rows.

```{r method-inputs, echo=FALSE}
validation_display_method_inputs()
```

The comparison uses deterministic adjustment methods that can be fitted during
the live vignette render. Each output uses the full LAD support prepared above.

```{r build-method-results}
deterministic_results <- list(
  unadjusted = mpd_od |>
    dplyr::mutate(flow_adj = flow),
  inverse_penetration = adjust_inverse_penetration(
    mpd_od_df = mpd_od,
    coverage_df = coverage_validation,
    weight_by = "both"
  ),
  selection_rate = adjust_selection_rate(
    mpd_od_df = mpd_od,
    coverage_df = coverage_validation,
    covariates_df = covariates_validation,
    covariate_col = "per_level4",
    weight_by = "origin",
    benchmark_od_df = benchmark_od,
    calibration_aggregate = "origin"
  ),
  selection_rate2 = adjust_selection_rate2(
    mpd_od_df = mpd_od,
    coverage_df = coverage_validation,
    weight_by = "origin",
    benchmark_od_df = benchmark_od,
    calibration_aggregate = "origin"
  ),
  raking_ratio = adjust_raking_ratio(
    mpd_od_df = mpd_od,
    benchmark_od_df = benchmark_od
  ),
  coefficient = adjust_coefficient(
    mpd_od_df = mpd_od,
    benchmark_od_df = benchmark_od,
    model_family = "ols",
    level = "od"
  )
)

all_method_results <- deterministic_results |>
  validation_standardize_adjusted_outputs(mpd_df = mpd_od)

method_results <- all_method_results[validation_teaching_methods]
broad_method_results <- all_method_results[validation_broad_methods]
```

The advanced Bayesian vignette covers Bayesian coverage-offset models. Those
models are not fitted here because full-LAD Bayesian runs are heavier than the
deterministic examples used in this live validation vignette. When full-LAD
Bayesian output files are regenerated for the same `r example_data$metadata$n_areas`
LAD support, they can be added back to this comparison without changing the
validation levels below.

Before validating, we check that every method output uses the same OD pairs and
the same raw MPD flow column. This is the main comparability check in the
vignette.

```{r validation-row-audit, echo=FALSE}
row_audit <- validation_audit_shared_validation_rows(
  adjusted_outputs = all_method_results,
  mpd_df = mpd_od,
  benchmark_df = benchmark_od
)
validation_assert_row_audit(row_audit)

validation_display_row_audit(row_audit)
```

The audit should read as a simple confirmation: all methods have the same
`r nrow(mpd_od)` OD pairs, no missing or extra pairs, and the raw MPD flow
column matches the same input data. If this check failed, the vignette would
stop rather than report metrics that mix adjustment performance with data
construction differences.

The support check covers the `r nrow(mpd_od)` OD rows used for validation.
Overall metrics below keep zero-flow rows for MAE, RMSE and correlation so that
methods are compared on the same support. MAPE is interpreted separately
because percentage errors cannot use zero benchmark flows as denominators.

## Broader method comparison

Before moving through the five validation levels, it is useful to look at a
compact broader comparison. This table includes the unadjusted MPD baseline
and five deterministic adjustment methods. It uses only the
adjusted-versus-benchmark comparison, because that is the main question: how
close is each final output to the benchmark OD table?

The column called "Benchmark used to fit" is important. Some deterministic
methods use benchmark totals or benchmark OD cells during calibration.

```{r broad-method-comparison, echo=FALSE}
broad_overall_results <- validation_overall_results(
  adjusted_outputs = broad_method_results,
  benchmark_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)

validation_display_compact_overall_metrics(
  broad_overall_results,
  methods = validation_broad_methods,
  include_benchmark_used = TRUE
)
```

```{r broad-method-plot}
plot_validation_metrics(
  broad_overall_results,
  error_measures = c("mae", "rmse"),
  comparisons = "adjusted_vs_benchmark",
  methods = validation_broad_methods,
  method_labels = validation_plot_method_labels
)
```

This broader table is deliberately compact. It is meant to show the range of
available choices without making every later section too crowded. The rest of
the vignette keeps a focused teaching set: Inverse penetration, Selection rate
II, Raking ratio and Coefficient adjustment. The advanced Bayesian vignette
explains the Bayesian model variants, parameters and returned columns in more
detail.

## Level 1: Overall validation metrics

Level 1 compares whole OD tables. The table below focuses on the main
adjusted-versus-benchmark comparison for the teaching methods. The secondary
comparisons are still computed for the plot and for interpretation, but they
are not all printed in the table.

```{r level1-overall-metrics, echo=FALSE}
overall_results <- validation_overall_results(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  comparisons = "all"
)

validation_display_compact_overall_metrics(
  overall_results,
  methods = validation_teaching_methods
)
```

Start with the error columns. Lower mean absolute error and RMSE mean the
compared flows are closer in size. The broader validation object also includes
signed mean error and MAPE, but the printed table keeps the count-scale error
metrics that are easiest to compare at this level.

Then look at the correlation columns. A high Pearson correlation means the
method keeps the same broad pattern of large and small flows as the benchmark.
A high Spearman correlation means the ranking of flows is similar. A method can
have a high correlation and still have large errors if it gets the pattern
right but the scale wrong.

```{r level1-metric-plot}
plot_validation_metrics(
  overall_results,
  error_measures = c("mae", "rmse"),
  comparisons = validation_benchmark_comparisons,
  method_labels = validation_plot_method_labels
)
```

The plot focuses on MAE and RMSE because they are on the flow-count scale.
MAPE is useful as a percentage-based check, but it is less helpful beside
count-scale errors because it behaves differently when benchmark flows are
small.

```{r level1-inference-values, include=FALSE}
level1_best_overall <- validation_best_method(
  overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "mae"
)
level1_best_cor <- validation_best_method(
  overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "pearson_r",
  largest = TRUE
)
```

In this example, `r level1_best_overall` has the lowest overall mean absolute
error against the benchmark. `r level1_best_cor` has the strongest Pearson
correlation. These are related but not identical statements: one is about
distance from the benchmark, and the other is about similarity in the overall
pattern of flows.

## Level 2: Origin and destination marginals

Level 2 compares geographical area totals. An **origin marginal** adds together
all flows leaving each origin. A **destination marginal** adds together all
flows arriving in each destination. This section focuses on four readable
examples from the comparison set.

This level is useful because an adjustment can look reasonable overall while
still overestimating or underestimating the flows of particular areas.

```{r level2-origin-marginal-metrics, echo=FALSE}
origin_margins <- validation_build_marginal_tables(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  role = "origin"
)

origin_margin_metrics <- validation_overall_results(
  adjusted_outputs = origin_margins$adjusted_outputs,
  benchmark_df = origin_margins$benchmark,
  comparisons = "all"
)

validation_display_compact_overall_metrics(
  origin_margin_metrics,
  methods = validation_plot_methods
)
```

```{r level2-destination-marginal-metrics, echo=FALSE}
destination_margins <- validation_build_marginal_tables(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  role = "destination"
)

destination_margin_metrics <- validation_overall_results(
  adjusted_outputs = destination_margins$adjusted_outputs,
  benchmark_df = destination_margins$benchmark,
  comparisons = "all"
)

validation_display_compact_overall_metrics(
  destination_margin_metrics,
  methods = validation_plot_methods
)
```

The scatterplots below show the data behind the marginal metrics. Each point is
an area total. Points near the dashed line have similar values in the two
datasets being compared. The colour shows the signed difference, so strong
colours flag areas where the gap is larger.

```{r level2-origin-scatter, fig.width=11, fig.height=8}
origin_margin_residuals <- validation_residual_results(
  adjusted_outputs = origin_margins$adjusted_outputs,
  benchmark_df = origin_margins$benchmark,
  top_n = 8
)

plot_validation_scatter(
  origin_margin_residuals,
  comparisons = validation_benchmark_comparisons,
  methods = validation_plot_methods,
  method_labels = validation_plot_method_labels,
  point_size = 2.2,
  point_alpha = 0.8
)
```

```{r level2-destination-scatter, fig.width=11, fig.height=8}
destination_margin_residuals <- validation_residual_results(
  adjusted_outputs = destination_margins$adjusted_outputs,
  benchmark_df = destination_margins$benchmark,
  top_n = 8
)

plot_validation_scatter(
  destination_margin_residuals,
  comparisons = validation_benchmark_comparisons,
  methods = validation_plot_methods,
  method_labels = validation_plot_method_labels,
  point_size = 2.2,
  point_alpha = 0.8
)
```

```{r level2-inference-values, include=FALSE}
level2_origin_best <- validation_best_method(
  origin_margin_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae"
)
level2_destination_best <- validation_best_method(
  destination_margin_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae"
)
```

In this example, `r level2_origin_best` is closest to the benchmark origin
totals by mean absolute error, while `r level2_destination_best` is closest to
the benchmark destination totals. If the best method differs between origins
and destinations, the choice depends on the purpose of the analysis. Commuting
outflows, service catchments and destination demand do not always place the
same weight on the two margins.

## Level 3: Individual OD-flow validation

Level 3 compares individual origin-destination pairs. This is the most granular
validation level and is the best place to inspect large local errors. This
section uses the same four-method subset as the plots so the summary, scatter
and outlier review line up.

```{r level3-residual-summary, echo=FALSE}
flow_residuals <- validation_residual_results(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  top_n = 12
)

validation_display_residual_summary(
  flow_residuals,
  methods = validation_plot_methods
)
```

The next scatterplot compares individual OD pairs. Each point is one OD pair.
The point colour is the signed difference between the two flows in the panel.
For adjusted-versus-benchmark panels, this is benchmark minus adjusted flow.
This makes outliers easier to see: points far from the dashed line and with
strong colour deserve attention.

```{r level3-flow-scatter, fig.width=11, fig.height=8}
plot_validation_scatter(
  flow_residuals,
  comparisons = validation_benchmark_comparisons,
  methods = validation_plot_methods,
  method_labels = validation_plot_method_labels,
  point_alpha = 0.55,
  point_size = 1.2
)
```

The residual-band plot summarises how much of the OD table falls into small or
large residual bands. The residual used here is benchmark minus adjusted flow.
Lower shares in the high-residual bands indicate fewer large remaining
OD-pair errors.

```{r level3-residual-bands}
plot_validation_residual_bands(
  flow_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = validation_teaching_methods,
  method_labels = validation_plot_method_labels,
  band_method = "sd",
  sd_reference = "adjusted",
  orientation = "horizontal",
  label_min_share = 4
)
```

The table below lists the largest remaining OD-pair residuals for the plot
subset. Positive values mean the benchmark flow is larger than the adjusted
flow. Negative values mean the adjusted flow is larger than the benchmark.

```{r level3-largest-residuals, echo=FALSE}
validation_display_top_residuals(
  flow_residuals,
  area_names = area_names,
  methods = validation_plot_methods,
  n = 3
)
```

```{r level3-inference-values, include=FALSE}
level3_metrics <- validation_overall_results(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)
level3_best <- validation_best_method(
  level3_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae"
)
```

At the individual OD-pair level, `r level3_best` has the lowest mean absolute
error in this example. The outlier tables and residual bands are useful because
the average can hide a small number of very large local errors.

## Level 4: Distributional allocation validation

The previous levels compare scale, residuals and outliers. Level 4 asks a
different question: whether each origin sends its flows to destinations in
similar proportions across the raw, adjusted and benchmark OD matrices.

This is an origin-conditioned allocation check. For each origin, flows are
converted into destination shares. The method is then assessed on whether those
shares resemble the benchmark shares. This matters because a method can improve
the total scale of flows while still allocating an origin's flows to the wrong
destinations.

```{r level4-distribution-summary, echo=FALSE}
distribution_results <- validation_distribution_results(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  comparisons = "all"
)

validation_display_distribution_summary(
  distribution_results,
  comparison = "adjusted_vs_benchmark",
  methods = validation_plot_methods
)
```

The plot below focuses on weighted Jensen-Shannon divergence. Lower values mean
the destination-share distribution is closer to the reference distribution.
The weighted version gives more influence to origins with larger benchmark
outflows.

```{r level4-distribution-plot}
plot_validation_distribution(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = validation_benchmark_comparisons,
  methods = validation_plot_methods,
  method_labels = validation_plot_method_labels
)
```

The comparison plot below shows the same idea from another angle: which pairs
of flow distributions are closest after converting flows into
origin-conditioned destination shares.

```{r level4-distribution-pairwise, fig.width=9, fig.height=6}
plot_validation_distribution_pairwise(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = "all",
  methods = validation_plot_methods,
  method_labels = validation_plot_method_labels,
  plot_type = "comparison",
  sort = "ascending"
)
```

```{r level4-inference-values, include=FALSE}
distribution_summary_tbl <- dplyr::bind_rows(lapply(names(distribution_results), function(method_id) {
  out <- distribution_results[[method_id]]$summary
  out$method <- method_id
  out
}))
level4_best <- distribution_summary_tbl |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::slice_min(jsd_weighted_mean, n = 1, with_ties = FALSE) |>
  dplyr::pull(method) |>
  validation_method_label()
```

In this example, `r level4_best` has the lowest benchmark-weighted
Jensen-Shannon divergence in the adjusted-versus-benchmark comparison. This is
not the same as having the lowest OD-flow error. It means the method is closest
to the benchmark in terms of the destination shares sent from each origin.

## Level 5: Spatial and residual-structure diagnostics

The previous levels can show that an adjustment improves average error, area
totals, individual OD pairs and destination-share allocation. Level 5 asks
whether the remaining residuals still look systematic.

Here we check three simple structures. First, are residuals associated with the
size of benchmark flows? Second, are origin-level residuals spatially clustered
across neighbouring areas? Third, are origin-level residuals associated with a
known area covariate? In this example, the covariate is the percentage of
residents with Level 4 qualifications or above (`per_level4`).

This section uses the same focused teaching set as the earlier plots. The
optional Moran's I and LISA diagnostics need a neighbour-link table. The Local
Moran statistic itself uses a reproducible four-nearest-neighbour table derived
from the real LAD centroid distances returned by `debiasR_example_data()`.
The LISA map uses a cached, simplified ONS 2021 LAD boundary file when `sf` and
network access are available. Readers can also set `DEBIAS_LAD_BOUNDARY_PATH`
to a local `sf`-readable boundary file. The diagnostic should be interpreted
as distance-neighbour rather than contiguity-neighbour LISA.

```{r prepare-residual-structure-inputs}
validation_lisa_inputs <- validation_prepare_lad_lisa_inputs(
  areas = covariates_validation$area,
  distance_df = distance_validation,
  distance_k = 4
)
structure_neighbors <- validation_lisa_inputs$neighbors
validation_lisa_boundaries <- validation_lisa_inputs$boundaries
validation_lisa_map_note <- validation_lisa_inputs$map_note

structure_covariates <- covariates_validation |>
  dplyr::select(area, per_level4)

structure_methods <- validation_plot_methods
```

```{r level5-neighbor-note, echo=FALSE}
cat(validation_lisa_inputs$note, "\n")
```

```{r level5-residual-structure, echo=FALSE}
residual_structure <- validation_structure_results(
  adjusted_outputs = method_results,
  benchmark_df = benchmark_od,
  area_neighbors = structure_neighbors,
  covariate_df = structure_covariates,
  covariate_col = "per_level4",
  methods = structure_methods,
  local_moran_nsim = 199,
  local_moran_seed = 20260613
)

validation_display_residual_structure(residual_structure)
```

Values close to zero are usually preferable for the residual-flow and
residual-covariate correlations, because they suggest less obvious residual
structure. For Moran's I, values close to zero suggest less spatial clustering
under the supplied neighbour definition. These are diagnostic checks, not final
proof that residuals are random.

```{r level5-structure-plot}
plot_validation_structure(
  residual_structure,
  methods = structure_methods,
  method_labels = validation_plot_method_labels
)
```

Local Moran's I provides a local version of the spatial clustering check. The
LISA class labels combine the sign of an area's centred residual with the
weighted residual pattern in its neighbours. The p-values are permutation
pseudo p-values from the supplied neighbour links, so they are best used as
exploratory flags for local review.

The LISA map below uses the same Local Moran output. Areas with raw permutation
pseudo p-values above 0.05 are shown as not significant. The table reports both
raw and adjusted p-values, so the map should be read as an exploratory local
screen rather than a multiple-testing-adjusted decision rule. Boundary polygons
come from a cached public ONS LAD boundary download, unless a local path is
supplied with `DEBIAS_LAD_BOUNDARY_PATH`. LADs that are present in the boundary
file but outside the validation support are retained in the same grey used for
"not significant" areas rather than dropped from the map. Scottish boundary
polygons are omitted because the validation example focuses on the England and
Wales support used by the LAD data.

```{r level5-local-moran, echo=FALSE}
validation_display_local_moran(
  residual_structure,
  area_names = area_names,
  methods = structure_methods,
  n = 8
)
```

```{r level5-lisa-map, fig.width=10, fig.height=7}
if (is.null(validation_lisa_boundaries)) {
  cat(validation_lisa_map_note, "\n")
} else {
  plot_validation_lisa_map(
    residual_structure,
    boundaries = validation_lisa_boundaries,
    boundary_area_col = "area",
    comparisons = "adjusted_vs_benchmark",
    methods = structure_methods,
    method_labels = validation_plot_method_labels,
    cluster_col = "lisa_quadrant",
    p_value_col = "p_value",
    p_value_threshold = 0.05,
    missing_fill = "#D1D5DB",
    boundary_colour = NA,
    boundary_linewidth = 0
  )
}
```

We can also inspect the origins with the largest remaining mean residuals. This
helps identify whether the main remaining problems are concentrated in a small
number of areas.

```{r level5-largest-area-residuals, echo=FALSE}
validation_display_residual_structure_areas(
  residual_structure,
  area_names = area_names,
  methods = structure_methods,
  n = 5
)
```

## Summary

The main lesson is to validate in levels. Start with the overall metrics, then
check origin and destination totals, then inspect individual OD-pair residuals,
then assess origin-conditioned destination allocation, and finally look for
spatial or covariate-linked residual structure.

The example also shows why the same validation data matter. All methods were
assessed on the same OD pairs and the same benchmark table. This makes the
method comparison interpretable. If methods were assessed on different rows,
different time periods or different benchmark definitions, the validation
results would mix adjustment performance with data construction differences.

The evidence from this vignette should therefore be read in a balanced way. In
some applications, a benchmark-assisted deterministic method may have the
lowest in-sample error, especially when the benchmark is trusted and matches
the period being estimated. In other applications, model-based approaches may
be preferable because they can bring in extra information and more flexible
structures. Those extensions need their own runtime and diagnostic checks.

For adoption, the practical question is not whether one method is always best.
The question is whether the selected method is close enough to the benchmark
for the intended use, whether remaining errors are understandable, and whether
the data used to fit the method are available and relevant for the period being
studied.
