---
title: "Validation"
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette describes the range of validation methods included in `debiasR`, to assess the extent of difference between our adjusted mobile-phone-derived flow estimates and target benchmark flows.

The methods assess differences between:

- adjusted *versus* benchmark flows, to evaluate the final adjustment.
- raw mobile-phone-derived *versus* benchmark flows, to evaluate the
  extent of bias before adjustment.
- raw *versus* adjusted flows, to evaluate the extent of adjustment made to the raw data.

The validation methods included in `debiasR` allows the assessment of adjustment outputs at three levels: overall, OD marginal and individual OD-pair adjustment. Tree type of assessment tools are included:
 error metrics, correlation coefficients and visual tools. These diagnostics provide flexibility to assess particular dimensions of the data which may vary depending on the target use of the adjusted flows.

## Intuition

A practical way to think of assessing the output of the various adjustment methods is to adopt a overall system-wide perspective of the flow system, move to an intermediate and then more granular origin-destination pair level. We use three levels.

The **Level 1** summarises the performance of flow adjustment methods across the whole origin-destination system.

The **Level 2** assesses the performance of flow adjustment methods by origin and destination totals.

The **Level 3** provides the performance of flow adjustment methods between individual origin-destination pairs.

Correlation coefficients offer a measure of correspondence between, for example, adjusted and benchmark flows, and they are better interpreted as a general measure. Error measures provide a measure of the level of difference between these sets of flows. They provide a general measure to assess the precision of the flow adjustment across the whole network of flows, but they also offer a way to measure and assess the precision of the flow adjustment for individual origin-destination pairs.

## Validation measures

Let $F^{adj}_{ij}$ be the adjusted flow from origin $i$ to destination
$j$, and let $F^{bench}_{ij}$ be the benchmark flow for the same OD pair.
Define the error as
$e_{ij} = F^{adj}_{ij} - F^{bench}_{ij}$. Positive errors mean the adjusted
flow is larger than the benchmark. Negative errors mean it is smaller. The
table uses adjusted-versus-benchmark notation.[^validation-measure-reuse]

| Metric | Description | Equation |
|:--|:--|:--|
| Mean error | Average signed error; shows whether flows are overestimated or underestimated on average. | $\frac{1}{n}\sum_{ij} e_{ij}$ |
| Mean absolute error | Average absolute error; shows the typical size of the error, ignoring sign. | $\frac{1}{n}\sum_{ij} \lvert e_{ij} \rvert$ |
| Root mean squared error | Error metric that gives more weight to large errors. | $\sqrt{\frac{1}{n}\sum_{ij} e_{ij}^{2}}$ |
| Median absolute error | Middle absolute error; less affected by a small number of very large errors. | $\mathrm{median}(\lvert e_{ij} \rvert)$ |
| Mean absolute percentage error | Average absolute error as a percentage of the benchmark flow. | $\frac{100}{n}\sum_{ij}\lvert \frac{e_{ij}}{F^{bench}_{ij}} \rvert$ |
| Pearson's r | Linear correlation; shows whether large benchmark flows also tend to be large adjusted flows. | $\mathrm{cor}(F^{adj}_{ij}, F^{bench}_{ij})$ |
| Spearman's rho | Rank correlation; shows whether the ordering of flows is similar. | $\mathrm{cor}(\mathrm{rank}(F^{adj}_{ij}), \mathrm{rank}(F^{bench}_{ij}))$ |

[^validation-measure-reuse]: The equations are written for
    adjusted-versus-benchmark flows. The same formulas can be reused for
    raw-versus-benchmark or raw-versus-adjusted comparisons by replacing
    $F^{adj}_{ij}$ and $F^{bench}_{ij}$ with the two flow columns being
    compared.

## Getting ready

We now get ready reading helper files, data and visualisation
tools.

```{r setup}
input_file <- knitr::current_input(dir = TRUE)
input_dir <- if (is.null(input_file)) {
  candidates <- c(".", "vignettes", "../vignettes", "../../vignettes")
  candidates[file.exists(file.path(candidates, "_common.R"))][1]
} else {
  dirname(input_file)
}
if (is.na(input_dir)) {
  stop(
    "Could not find vignettes/_common.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
input_dir <- normalizePath(input_dir, mustWork = TRUE)
debiasr_pkg_root <- normalizePath(file.path(input_dir, ".."), mustWork = FALSE)
source(file.path(input_dir, "_common.R"))
source(file.path(input_dir, "_validation_helpers.R"))
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
library(ggplot2)
```

In this vignette, we will illustrate the validation methods in `debiasR` comparing the two adjustment methods. We compare the inverse penetration and selection rate II methods. First,
we separately organise the raw mobile-phone-derived data and benchmark data. We then apply inverse penetration and selection rate II, and
join the raw, adjusted and benchmark flows into one data frame for comparison.

```{r}
mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od

method_results <- list(
  inverse_penetration = debiasR::adjust_inverse_penetration(
    mpd_od_df = mpd_od,
    coverage_df = example_data$coverage,
    weight_by = "both"
  ),
  selection_rate2 = debiasR::adjust_selection_rate2(
    mpd_od_df = mpd_od,
    coverage_df = example_data$coverage,
    weight_by = "origin",
    benchmark_od_df = benchmark_od,
    calibration_aggregate = "origin"
  )
)

flow_comparison <- validation_build_flow_comparison(
  method_results = method_results,
  mpd_df = mpd_od,
  benchmark_df = benchmark_od
)
```

## Level 1: Overall validation

This level provides a single number summary metric for the whole system of origin-destination flows. Each row in the table below summarises all
OD pairs for one method and one comparison. The two methods are **Inverse
penetration** and **Selection rate II**.

The main rows are the adjusted-versus-benchmark rows. These rows tell us
whether the adjusted outputs are close to the Census benchmark. The
raw-versus-benchmark rows show the starting point before adjustment. The
raw-versus-adjusted rows show how much the method moved the original data.

```{r}
overall_metrics <- validation_metrics_by_method(
  data = flow_comparison,
  level = "overall"
)

overall_metrics |>
  dplyr::arrange(comparison, mae) |>
  validation_display_metrics()
```

Start with the error columns. Lower mean absolute error, RMSE and median
absolute error mean the compared flows are closer in size. The sign of mean
error tells us direction: a positive value means overestimation on average, and
a negative value means underestimation on average.

Then look at the correlation columns. A high Pearson correlation means the
method keeps the same broad pattern of large and small flows as the benchmark.
A high Spearman correlation means the ranking of flows is similar. A method can
have a high correlation and still have large errors if it gets the pattern
right but the scale wrong.

```{r}
level1_primary <- overall_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark")

level1_lowest_mae <- level1_primary |>
  dplyr::slice_min(mae, n = 1, with_ties = FALSE) |>
  dplyr::pull(method_label)

level1_highest_r <- level1_primary |>
  dplyr::slice_max(pearson_r, n = 1, with_ties = FALSE) |>
  dplyr::pull(method_label)
```

In this example, `r level1_lowest_mae` has the lower mean absolute error in the
adjusted-versus-benchmark comparison. `r level1_highest_r` has the stronger
Pearson correlation. These two statements answer different questions: one is
about distance from the benchmark, and the other is about similarity in the
overall pattern of flows.

## Level 2: Origin and destination marginals

This level compares area totals. An **origin marginal** adds together all flows
leaving each area. It asks whether each area has the right total outflow. A
**destination marginal** adds together all flows arriving in each area. It asks
whether each area has the right total inflow.

We again compare **Inverse penetration** and **Selection rate II**. This level
is useful because a method can look reasonable overall while still putting too
many flows into some areas and too few into others.

```{r}
marginal_comparison <- validation_build_margins(flow_comparison)
marginal_metrics <- validation_metrics_by_method_level(marginal_comparison)

marginal_metrics |>
  dplyr::arrange(level, comparison, mae) |>
  validation_display_metrics()
```

The error metrics now refer to area totals rather than individual OD pairs.
For example, a lower origin-marginal MAE means the method is closer to the
benchmark total outflow for the typical origin area. A lower
destination-marginal MAE means it is closer to the benchmark total inflow for
the typical destination area. As before, the adjusted-versus-benchmark rows are
the main validation rows. The raw-versus-benchmark rows show the starting
bias in area totals, and the raw-versus-adjusted rows show how much each method
changed those totals.

The scatterplots below show the same idea visually for the three comparisons.
Each point is an area total. Points near the dashed line have similar values in
the two flow tables being compared. Points above the line are larger in the
compared table on the vertical axis. Points below the line are smaller in that
compared table.

```{r}
validation_plot_margin_scatter(marginal_comparison)
```

The table below lists the largest remaining marginal errors. It helps identify
whether the remaining error is concentrated in a small number of areas.

```{r}
area_names <- example_data$covariates |>
  dplyr::select(area, name)

marginal_patterns <- marginal_comparison |>
  dplyr::left_join(area_names, by = "area") |>
  dplyr::mutate(
    raw_residual = flow_mpd - flow_bench,
    adjusted_residual = flow_adj - flow_bench,
    abs_adjusted_residual = abs(adjusted_residual),
    abs_error_reduction = abs(raw_residual) - abs(adjusted_residual)
  )

marginal_patterns |>
  dplyr::group_by(method_label, level) |>
  dplyr::slice_max(order_by = abs_adjusted_residual, n = 3, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(method_label, level, dplyr::desc(abs_adjusted_residual)) |>
  dplyr::select(
    method = method_label,
    level,
    area,
    name,
    flow_adj,
    flow_bench,
    adjusted_residual,
    abs_error_reduction
  )
```

```{r}
level2_best <- marginal_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::group_by(level) |>
  dplyr::slice_min(mae, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

level2_origin_best <- level2_best |>
  dplyr::filter(level == "origin_marginal") |>
  dplyr::pull(method_label)

level2_destination_best <- level2_best |>
  dplyr::filter(level == "destination_marginal") |>
  dplyr::pull(method_label)
```

In this example, `r level2_origin_best` is closer to benchmark origin totals by
mean absolute error, while `r level2_destination_best` is closer to benchmark
destination totals. If the same method is closer at both margins, the marginal
evidence points in one direction. If the two margins favour different methods,
the choice depends on whether origin outflows or destination inflows matter
more for the analysis.

## Level 3: Individual OD-flow validation

This level compares each origin-destination pair directly. The methods are
again **Inverse penetration** and **Selection rate II**. This is the most
detailed level, and it is where large local errors become visible.

```{r}
flow_level_metrics <- validation_metrics_by_method(
  data = flow_comparison,
  level = "individual_flow"
)

flow_level_metrics |>
  dplyr::arrange(comparison, mae) |>
  validation_display_metrics()
```

The scatterplot below repeats the three comparisons for individual OD pairs.
Each point is one origin-destination flow. Points close to the dashed line are
similar in the two flow tables being compared. Points far from the line are OD
flows where the compared value differs strongly from the reference value.

```{r}
validation_plot_od_scatter(flow_comparison)
```

The next scatterplot highlights outliers by colouring each point according to
the signed adjusted-minus-benchmark difference. Blue points are adjusted flows
below the benchmark. Red points are adjusted flows above the benchmark. Darker
points have larger residuals.

```{r}
validation_plot_residual_outliers(flow_comparison)
```

The table below ranks individual OD pairs by the largest remaining
adjusted-minus-benchmark residuals. Positive values mean the adjusted estimate
is larger than the benchmark. Negative values mean it is smaller.

```{r}
validation_rank_largest_residuals(flow_comparison, n = 12)
```

The compact heatmap-style table below groups OD-flow residuals by standard
deviation bands. The residual for each adjustment method is the
adjusted-minus-benchmark difference, $F^{adj}_{ij} - F^{bench}_{ij}$. The SD
bands are defined from the pooled adjusted-minus-benchmark residuals across
the two adjustment methods, so a band means the same thing in each column.

Each cell is the percentage of OD pairs in that residual band, rounded to two
decimal places. For example, a value of 0.16 means that 0.16% of OD pairs in
that column fall in the residual-size band named on the row. Lower percentages
in the high-SD rows indicate fewer large outliers.

```{r}
residual_heatmap <- validation_build_residual_heatmap(flow_comparison)
validation_plot_residual_heatmap(residual_heatmap)
```

The heatmap also includes the unadjusted raw MPD baseline. This baseline uses
the raw-minus-benchmark difference, $F^{mpd}_{ij} - F^{bench}_{ij}$, but it is
classified using the same adjusted-residual SD bands. This makes the visual
comparison direct: if an adjustment reduces the share of flows above 2, 3 or 4
standard deviations, it has reduced the number of large OD-flow errors in this
example.

```{r}
level3_best <- flow_level_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::slice_min(mae, n = 1, with_ties = FALSE) |>
  dplyr::pull(method_label)

large_outlier_summary <- residual_heatmap |>
  dplyr::filter(residual_band %in% c("Greater than 4.0 SD", "3.0 to 4.0 SD")) |>
  dplyr::group_by(method_label) |>
  dplyr::summarise(large_outlier_share = sum(share), .groups = "drop") |>
  dplyr::filter(method_label != "Unadjusted") |>
  dplyr::slice_min(large_outlier_share, n = 1, with_ties = FALSE)

level3_outlier_best <- large_outlier_summary |>
  dplyr::pull(method_label)
```

At the individual-flow level, `r level3_best` has the lower mean absolute error
against the benchmark in this example. Looking only at the largest residual
bands, `r level3_outlier_best` has the smaller share of OD pairs above three
standard deviations. These results should be read together: average error tells
us about typical performance, while the outlier table shows whether a method
still leaves a few very large local errors.

## Summary

Validation works best as a sequence. First, check the overall metrics to see
whether adjusted flows are broadly closer to the benchmark. Then check origin
and destination marginals to see whether area totals are plausible. Finally,
inspect individual OD pairs to find large local errors.

In this vignette, the two methods answer slightly different needs. Inverse
penetration is simple and transparent. Selection rate II adds a calibrated
nonlinear correction. The validation tables and figures show which method is
closer to the benchmark for this example, but the right choice in another
application should depend on the available data, the validation target, and the
flows that matter most for the analysis.
