---
title: "Visualising origin-destination flows"
description: >
  Prepare origin-destination flows and area centroids for exploratory static
  and interactive flow visualisation.
vignette: >
  %\VignetteIndexEntry{Visualising origin-destination flows}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette shows how origin-destination (OD) flows can be prepared for
visual inspection. The aim is not to replace the formal validation diagnostics
in `debiasR`, but to give a quick way to see which origin-destination corridors
carry the largest flows and where those corridors are located.

The examples use benchmark Census travel-to-work flows between Local Authority
Districts (LADs) from `debiasRdata`, together with LAD centroid coordinates.
The same structure can be used for observed mobile-phone-derived flows or
adjusted flows returned by `debiasR`.

## Getting ready

The README shows how to install `debiasR` and the companion `debiasRdata`
package from GitHub. This vignette also uses `dplyr`, `ggplot2` and `scales`
for data preparation and plotting.

```{r load-packages}
library(debiasR)
library(debiasRdata)
```

We start with the LAD benchmark OD-flow table and the LAD centroid table. The
OD-flow table stores one row per origin-destination pair. The centroid table
stores one coordinate pair per LAD.

```{r load-flow-data}
benchmark_flows <- debiasRdata::census_lad_OD_travel2work
centroids <- debiasRdata::lad_centroids
```

## Identify the largest flows

Flow maps can become difficult to read if every OD pair is plotted. A practical
first step is to focus on the largest flows. This keeps the visualisation
readable and makes the plotted corridors easier to interpret.

```{r largest-flows}
largest_flows <- benchmark_flows |>
  dplyr::filter(origin != destination) |>
  dplyr::slice_max(order_by = flow, n = 40, with_ties = FALSE)

largest_flows |>
  dplyr::slice_head(n = 10)
```

The table above shows the largest between-area flows. Within-area flows are
excluded here because they have the same origin and destination centroid, so
they do not produce visible lines on a centroid-to-centroid plot.

## Attach coordinates

To draw a line between each origin and destination, the flow table needs origin
coordinates and destination coordinates. We create two coordinate lookup tables
from the same LAD centroid data and join them to the selected flows.

```{r attach-coordinates}
origin_locations <- centroids |>
  dplyr::select(
    origin = area,
    origin_name = name,
    origin_lon = longitude,
    origin_lat = latitude
  )

destination_locations <- centroids |>
  dplyr::select(
    destination = area,
    destination_name = name,
    destination_lon = longitude,
    destination_lat = latitude
  )

flow_lines <- largest_flows |>
  dplyr::left_join(origin_locations, by = "origin") |>
  dplyr::left_join(destination_locations, by = "destination") |>
  dplyr::filter(
    !is.na(origin_lon),
    !is.na(origin_lat),
    !is.na(destination_lon),
    !is.na(destination_lat)
  )

flow_lines |>
  dplyr::select(
    origin_name,
    destination_name,
    flow,
    origin_lon,
    origin_lat,
    destination_lon,
    destination_lat
  ) |>
  dplyr::slice_head(n = 10)
```

This `flow_lines` table is the main structure needed by most flow visualisation
tools: each row has an origin identifier, a destination identifier, a flow value
and coordinates for both ends of the OD pair.

## Static flow plot

A simple static plot is useful for reports, teaching material and package
documentation because it does not depend on web map tiles or experimental
interactive mapping packages. The plot below uses LAD centroids and straight
line segments. It is therefore an exploratory diagram of the OD network, not a
route map.

```{r static-flow-plot, fig.width=7, fig.height=7}
ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = flow_lines,
    ggplot2::aes(
      x = origin_lon,
      y = origin_lat,
      xend = destination_lon,
      yend = destination_lat,
      linewidth = flow,
      alpha = flow
    ),
    colour = "#246A73",
    lineend = "round"
  ) +
  ggplot2::geom_point(
    data = centroids,
    ggplot2::aes(x = longitude, y = latitude),
    size = 0.7,
    colour = "grey35",
    alpha = 0.6
  ) +
  ggplot2::scale_linewidth_continuous(
    range = c(0.2, 2.8),
    labels = function(x) format(x, big.mark = ",", scientific = FALSE)
  ) +
  ggplot2::scale_alpha_continuous(range = c(0.25, 0.85), guide = "none") +
  ggplot2::coord_quickmap() +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    linewidth = "Census flow",
    title = "Largest Census travel-to-work flows between LADs"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    legend.position = "bottom"
  )
```

The figure highlights the largest benchmark travel-to-work corridors among the
selected LADs. Larger and darker lines represent larger flows. This kind of
plot is useful for checking whether a small number of corridors dominate the
OD system, identifying flows that may deserve closer inspection and comparing
raw, adjusted or benchmark flow patterns.

## Visualise other flow columns

The same workflow can be used for observed mobile-phone-derived flows or
adjusted flows. The only requirement is that the table has an origin column, a
destination column and a numeric flow column. For example, after using an
adjustment method from `debiasR`, replace `benchmark_flows` with the adjusted
table and select `flow_adj` as the plotted value.

```{r eval=FALSE}
adjusted_flows <- debiasR::adjust_inverse_penetration(
  mpd_od_df = debiasRdata::lad_OD_travel2work,
  coverage_df = debiasRdata::coverage_lad,
  weight_by = "origin"
)

largest_adjusted_flows <- adjusted_flows |>
  dplyr::filter(origin != destination) |>
  dplyr::slice_max(order_by = flow_adj, n = 40, with_ties = FALSE)
```

In a validation workflow, visualisation is most useful when it is paired with
formal metrics. A map can show where large flows are located, but it does not
quantify whether adjusted flows are closer to a benchmark. For that, use the
validation methods in the validation vignette.

## Optional interactive flow maps

Interactive flow maps can be useful when there are many OD pairs and users need
to zoom, pan or inspect corridors. Issue #27 suggested the experimental
flow-map support in `mapgl`. Because that functionality depends on a development
branch, the example below is not evaluated when this vignette is rendered.

```{r eval=FALSE}
# install.packages("pak")
pak::pak("e-kotov/mapgl@flowmap")
pak::pak("de-bias/debiasRdata")

locations <- data.frame(
  id = debiasRdata::lad_centroids$area,
  lon = debiasRdata::lad_centroids$longitude,
  lat = debiasRdata::lad_centroids$latitude,
  stringsAsFactors = FALSE
)

flows <- data.frame(
  origin = debiasRdata::census_lad_OD_travel2work$origin,
  dest = debiasRdata::census_lad_OD_travel2work$destination,
  count = debiasRdata::census_lad_OD_travel2work$flow,
  stringsAsFactors = FALSE
)

mapgl::maplibre(
  style = mapgl::carto_style("dark-matter"),
  center = c(-2.0, 54.0),
  zoom = 5
) |>
  mapgl::add_flowmap(
    id = "census-lad-flows",
    locations = locations,
    flows = flows,
    flow_color_scheme = "Teal",
    flow_dark_mode = TRUE,
    flow_max_top_flows_display_num = 10000
  )
```

Other packages can also be used for flow visualisation. `flowmapper` is a
possible route for static flow maps, while `flowmapblue` provides interactive
flow maps but is less current than the emerging `mapgl` route.
