---
title: "Advanced Bayesian multilevel adjustment"
vignette: >
  %\VignetteIndexEntry{Advanced Bayesian multilevel adjustment}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
main table {
  font-size: 0.9rem;
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}
</style>
```

## Aim

This vignette explains the advanced implementation routes in
`adjust_multilevel_bayes()`. The main adjustment vignette introduces the
recommended default, `observation_model = "coverage_offset"`, because it is the
most practical starting point for routine Bayesian adjustment. Start with
[Adjusting origin-destination flows](v06-adjusting-biases.html) if you want the
default workflow first. Here we compare the DEBIAS observation-model choices,
show how the fast deterministic engine can be used for specification checks, and
explain when the approved advanced latent two-level backend is useful.

The guiding principle is progressive complexity: use the simplest model that
matches the data structure and inferential goal.

## Why more than one implementation route?

`adjust_multilevel_bayes()` supports three Bayesian observation-model ideas and
one fast non-posterior engine:

```{r echo=FALSE}
bayesian_modes <- tibble::tribble(
  ~Route, ~Status, ~Main_use, ~Core_idea,
  "`coverage_offset`",
  "Recommended default",
  "Routine true-flow adjustment and teaching examples",
  "Treat MPD flows as coverage-scaled observations of true flows, with coverage fixed as an offset.",
  "`latent_two_level`",
  "Experimental advanced backend",
  "Repeated source/time data where multiple MPD observations can share a hidden OD or OD-time state",
  "Create a latent-flow state and treat repeated MPD rows as observations of that state.",
  "`reduced_form`",
  "Legacy/backward-compatible",
  "Reproducing earlier analyses or sensitivity checks on the MPD observation scale",
  "Fit observed MPD flows directly and adjust by setting fitted bias terms to zero.",
  "`model_engine = \"frequentist\"`",
  "Fast deterministic scaffold",
  "Testing formulas, covariates, source/time structures and count families before Bayesian sampling",
  "Use the same non-latent data contract without posterior sampling."
)

knitr::kable(bayesian_modes)
```

The default coverage-offset model and the latent two-level model both return
adjusted flows on a true-flow scale. They differ in what is estimated inside the
model. The coverage-offset model estimates the true-flow scale by removing a
fixed coverage offset from an observed-flow likelihood. The latent backend
estimates a shared latent-flow intensity directly, so it is most useful when the
same underlying OD or OD-time flow is observed by multiple sources. The
frequentist engine is not a Bayesian model. It is a fast way to check whether
the data contract, formulas and distributional choices are sensible before
running a slower Bayesian implementation.

## Choosing a Bayesian model

```{r echo=FALSE}
decision_table <- tibble::tribble(
  ~Data_or_goal, ~Recommended_mode, ~Reason,
  "One source and one time period, or a small teaching example",
  "`coverage_offset`",
  "The latent split is weakly identified when each OD pair is observed once.",
  "Complete-grid prediction over supplied OD rows",
  "`coverage_offset`",
  "The model can predict true-flow scale for supplied rows using the coverage offset and mobility formula.",
  "Multiple sources observing the same OD pairs",
  "`latent_two_level`",
  "Repeated source observations can share an OD-level latent state.",
  "Multiple sources and multiple time periods",
  "`latent_two_level` with care",
  "Repeated source observations can share OD-time states, but time and source effects need careful interpretation.",
  "Fast design, tests or method comparison",
  "`model_engine = \"frequentist\"` with non-latent modes",
  "The frequentist engine preserves the S1-S4 data contract but does not fit the latent Bayesian backend.",
  "Backward compatibility with earlier reduced-form analyses",
  "`reduced_form`",
  "This preserves the older MPD-scale counterfactual behaviour."
)

knitr::kable(decision_table)
```

As a rule of thumb, start with `coverage_offset`. Move to
`latent_two_level` only when the repeated-observation structure gives the model
real information about the hidden flow state.

## How the model variants differ

The three Bayesian observation-model variants answer related but different
questions. They should not all be read through the two-equation
coverage-offset explanation used in the main adjustment vignette.

```{r echo=FALSE}
variant_structure_table <- tibble::tribble(
  ~Variant, ~Main_question, ~Adjustment_target, ~What_is_not_estimated,
  "`coverage_offset`",
  "What true OD flow would have produced the observed MPD flow, given known coverage?",
  "Census-scale true-flow prediction",
  "Coverage is fixed as an offset; there is no separate source-invariant latent OD state.",
  "`reduced_form`",
  "What would the fitted MPD flow look like if the fitted bias term were neutral?",
  "MPD-scale counterfactual prediction",
  "It does not model MPD as coverage-scaled observations of Census-scale true flows.",
  "`latent_two_level`",
  "What shared hidden OD or OD-time flow state is implied by repeated source/time observations?",
  "Latent true-flow prediction",
  "The latent state is not source-specific; source/time variation belongs to the observation layer."
)

knitr::kable(variant_structure_table)
```

### Coverage-offset true-flow route

The coverage-offset route treats the MPD count as a partial observation of an
underlying true flow. Its observation equation is:

$$
Y^{mpd}_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) = \log(q_{ijst}) + \eta^{true}_{ijst}
$$

where $Y^{mpd}_{ijst}$ is the observed MPD flow for origin $i$, destination
$j$, source $s$ and time $t$; $q_{ijst}$ is the coverage-derived observation
probability; and $\eta^{true}_{ijst}$ is the true-flow linear predictor. In a
simple S1 setting, the source and time subscripts can be dropped:

$$
\log(\mu_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}
$$

The true-flow predictor can include origin covariates, destination covariates,
distance, and optional random-effect terms:

$$
\eta^{true}_{ij}
= \alpha + X_i\beta_o + X_j\beta_d + \gamma \log(d_{ij}) + u_i + v_j + w_{ij}
$$

The adjusted flow is the prediction after removing the fixed coverage offset:

$$
F^{adj}_{ij} = \exp(\eta^{true}_{ij})
$$

This route is not a latent-state model in the strict sense. The true-flow scale
is recovered by fitting a count model with a fixed observation offset and then
predicting with that offset set to zero. If source or time terms are included in
the formula, they are fitted as predictors in this non-latent model; the route
does not force one shared source-invariant OD state unless the formula and data
structure imply that restriction.

### Reduced-form compatibility route

The reduced-form route preserves the earlier observed-flow correction logic. It
fits the MPD observations directly:

$$
Y^{mpd}_{ijst} \sim \mathrm{Poisson}(\mu^{mpd}_{ijst})
$$

$$
\log(\mu^{mpd}_{ijst}) =
\eta^{mobility}_{ijst} + \eta^{bias}_{ijst}
$$

The mobility part can contain OD covariates, distance and source/time terms.
The bias part contains fitted bias covariates, such as an origin coverage-bias
term:

$$
\eta^{bias}_{ijst} = \delta b_i
$$

The adjustment is a counterfactual prediction in which the fitted bias term is
set to a neutral value:

$$
F^{adj}_{ij} =
\exp(\eta^{mobility}_{ij} + \eta^{bias}_{ij}\mid b_i = 0)
$$

This means the reduced-form route returns an MPD-scale counterfactual, not a
Census-scale true-flow state identified through a coverage offset. It is useful
for reproducing earlier analyses and for sensitivity checks, but it should not
be described as the same observation-process model as `coverage_offset`.

### Latent two-level route

The latent two-level route is the advanced model that most directly matches
the phrase "two-level" in a statistical sense. It introduces a shared hidden
true-flow state and treats repeated source/time MPD rows as observations of
that state:

$$
Y^{mpd}_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) =
\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma
$$

where $\theta_{g(i,j,t)} = \log(\lambda^{true}_{g(i,j,t)})$ is the log latent
true-flow intensity for an OD or OD-time group, and $z_{ijst}\gamma$ captures
observation-layer structure such as source and time effects. The latent group is
controlled by `latent_flow_unit`: `latent_flow_unit = "od"` shares one state
for each OD pair, while `latent_flow_unit = "od_time"` shares one state for
each OD-time pair.

The adjusted true-flow prediction is:

$$
F^{adj}_{g} = \lambda^{true}_{g} = \exp(\theta_g)
$$

The fitted MPD observation-scale prediction keeps the source/time observation
layer:

$$
F^{mpd,pred}_{ijst} =
\exp(\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma)
$$

This route is most informative when repeated observations make the latent state
empirically meaningful. In S3, multiple sources observe the same OD state. In
S4, multiple source/time rows can share OD-time states. In S1, each state is
observed only once, so the latent split is weakly identified and the
coverage-offset route is usually clearer.

## Simulated source-time example

The examples below use a small simulated dataset so the code can be read and
adapted without requiring external empirical files. The simulation creates five
areas, a hidden true OD-flow surface, two MPD sources and two time periods.
Coverage differs by area, source and time. The observed MPD rows are generated
as coverage-scaled measurements of the hidden true flows.

```{r simulated-source-time-data}
set.seed(20260625)

areas <- paste0("area_", LETTERS[1:5])
area_profile <- data.frame(
  area = areas,
  area_index = seq_along(areas),
  population = c(900, 1200, 760, 1500, 1080),
  rural_pct = c(0.10, 0.25, 0.55, 0.40, 0.70),
  stringsAsFactors = FALSE
)

od_grid <- expand.grid(
  origin = areas,
  destination = areas,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

origin_profile <- area_profile
names(origin_profile) <- paste0(names(origin_profile), "_o")
names(origin_profile)[names(origin_profile) == "area_o"] <- "origin"

destination_profile <- area_profile
names(destination_profile) <- paste0(names(destination_profile), "_d")
names(destination_profile)[names(destination_profile) == "area_d"] <- "destination"

od_truth <- merge(od_grid, origin_profile, by = "origin")
od_truth <- merge(od_truth, destination_profile, by = "destination")
od_truth$distance_km <- abs(od_truth$area_index_o - od_truth$area_index_d) + 1
od_truth$log_distance <- log(od_truth$distance_km)
od_truth$self_flow <- as.numeric(od_truth$origin == od_truth$destination)
od_truth$eta_true <- 5.6 +
  0.45 * od_truth$rural_pct_o -
  0.25 * od_truth$rural_pct_d -
  0.70 * od_truth$log_distance +
  0.90 * od_truth$self_flow
od_truth$flow_true <- round(exp(od_truth$eta_true))

source_time_grid <- expand.grid(
  mpd_source = c("operator_a", "operator_b"),
  mpd_time = c("2021_q1", "2021_q2"),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

coverage_repeated <- merge(
  area_profile[c("area", "population")],
  source_time_grid
)
names(coverage_repeated)[names(coverage_repeated) == "area"] <- "origin"
coverage_repeated$coverage_rate <- with(
  coverage_repeated,
  ifelse(mpd_source == "operator_a", 0.09, 0.13) +
    ifelse(mpd_time == "2021_q2", 0.015, 0) +
    c(0.005, -0.004, 0.002, -0.003, 0.004)[match(origin, areas)]
)
coverage_repeated$user_count <- pmax(
  1,
  round(coverage_repeated$population * coverage_repeated$coverage_rate)
)

mpd_repeated <- merge(
  od_truth[c("origin", "destination", "flow_true")],
  source_time_grid
)
mpd_repeated <- merge(
  mpd_repeated,
  coverage_repeated[c("origin", "mpd_source", "mpd_time", "coverage_rate")],
  by = c("origin", "mpd_source", "mpd_time")
)
mpd_repeated$source_time_effect <- with(
  mpd_repeated,
  ifelse(mpd_source == "operator_b", 1.12, 0.95) *
    ifelse(mpd_time == "2021_q2", 1.04, 1.00)
)
mpd_repeated$flow <- stats::rpois(
  nrow(mpd_repeated),
  lambda = pmax(
    0.1,
    mpd_repeated$coverage_rate *
      mpd_repeated$flow_true *
      mpd_repeated$source_time_effect
  )
)
mpd_repeated <- mpd_repeated[
  c("origin", "destination", "flow", "mpd_source", "mpd_time")
]

coverage <- coverage_repeated[
  c("origin", "population", "user_count", "mpd_source", "mpd_time")
]
covariates <- area_profile[c("area", "rural_pct", "population")]
distance <- od_truth[c("origin", "destination", "distance_km")]

mpd_s1 <- subset(
  mpd_repeated,
  mpd_source == "operator_a" & mpd_time == "2021_q1"
)
coverage_s1 <- subset(
  coverage,
  mpd_source == "operator_a" & mpd_time == "2021_q1"
)

mpd_s3 <- subset(mpd_repeated, mpd_time == "2021_q1")
coverage_s3 <- subset(coverage, mpd_time == "2021_q1")

mpd_s4 <- mpd_repeated
coverage_s4 <- coverage

knitr::kable(head(mpd_s4, 8))
```

The S1 object contains one source and one time period. The S3 object contains
two sources for one time period. The S4 object contains two sources and two time
periods. These small objects are intentionally simple, but they exercise the
same parameter contracts used by empirical source/time flow data.

The Bayesian model-fitting chunks below are shown but not evaluated during
ordinary vignette rendering. Run them manually in RStudio when you want posterior
summaries, and inspect `attr(result, "diagnostics")` before interpreting the
adjusted flows. The deterministic frequentist check in this vignette is
evaluated because it is fast and is useful for confirming that the simulated
source/time structure has been declared correctly.

## Default coverage-offset model

The default true-flow model is:

$$
F^{mpd}_{ij} \sim \mathrm{Poisson}(q_{ij}F^{true}_{ij})
$$

$$
\log E(F^{mpd}_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}
$$

where $q_{ij}$ is derived from active-user coverage. With
`coverage_scale = "origin"`, $q_{ij} = c_i = U_i / P_i$. The returned
`flow_adj` is a posterior summary of:

$$
F^{true,pred}_{ij} = \exp(\eta^{true}_{ij})
$$

This is the model explained in the main adjustment vignette. It is usually the
best option for S1 examples and for users who want readable, reproducible
Bayesian adjustment without modelling source-specific observation layers.

```{r eval=FALSE}
adj_default <- debiasR::adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "observed",
  random_intercept = "none",
  model_family = "poisson",
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```

With richer data, the coverage-offset model can include multilevel structure in
the true-flow formula. Random-effect terms represent origin, destination and
corridor structure: `(1 | origin)` for an origin intercept, `(1 | destination)`
for a destination intercept, and `(1 | od_id)` for an OD-corridor intercept.
Use these terms when the data contain enough repeated information to estimate
the pooling structure.

```{r eval=FALSE}
adj_default_multilevel <- debiasR::adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance +
    (1 | origin) + (1 | destination) + (1 | od_id),
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

## Fast deterministic specification check

Before running the Bayesian examples, it is often useful to check the same data
contract with the fast deterministic engine. The code below uses the S4
simulated data, declares source and time columns, and fits the coverage-offset
route without posterior sampling. This is a practical way to test formulas,
covariates, count families and source/time declarations before paying the MCMC
runtime cost.

```{r frequentist-s4-check}
adj_fast_s4 <- debiasR::adjust_multilevel_bayes(
  mpd_od_df = mpd_s4,
  coverage_df = coverage_s4,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance +
    mpd_source + mpd_time,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "frequentist",
  scenario = "s4",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source_time",
  prediction_scope = "observed",
  random_intercept = "none",
  model_family = "poisson"
)

knitr::kable(head(adj_fast_s4, 8), digits = 2)
```

The frequentist result should be interpreted as a fast design check. It tells
you whether the scenario, coverage joins, formulas and prediction scope are
working, but it is not posterior inference.

## Latent two-level backend

The latent two-level backend is for repeated source/time settings. It keeps
the true-flow target but introduces a shared latent-flow state. Conceptually:

$$
Y_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) =
\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma
$$

where:

- $Y_{ijst}$ is the observed MPD flow for origin $i$, destination $j$, source
  $s$ and time $t$.
- $q_{ijst}$ is the coverage-derived observation probability.
- $\theta_{g(i,j,t)} = \log(\lambda^{true}_{g(i,j,t)})$ is the log shared
  latent true-flow intensity for an OD or OD-time group.
- $z_{ijst}\gamma$ represents observation-layer structure such as coverage,
  source and time effects.

In the approved custom Stan backend, `adjust_multilevel_bayes()` represents
this shared state through `latent_flow_id` and estimates source-invariant
latent effects with an identified sum-to-zero parameterization. The returned
`flow_adj` and `flow_true_pred` summarize posterior draws of the latent
true-flow intensity, while `flow_mpd_pred` summarizes the source/time-specific
MPD observation mean.

```{r eval=FALSE}
adj_latent <- debiasR::adjust_multilevel_bayes(
  mpd_od_df = mpd_s3,
  coverage_df = coverage_s3,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "latent_two_level",
  backend = "auto",
  coverage_scale = "origin",
  latent_flow_unit = "auto",
  model_engine = "bayesian",
  scenario = "s3",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

The returned table includes `latent_flow_id`, `latent_flow_unit`,
`flow_adj_mean`, `flow_adj_median`, and 95% interval columns for the true-flow
and MPD observation scales. The result metadata also records the custom backend,
the number of latent states, the true-flow and observation formulas, latent
prior/sampler controls, ignored formula random-effect terms, and identifiability
notes.
Because this mode relies on repeated observations, it is most informative for
S3 and S4 designs. In S1 and S2, use it mainly as a sensitivity or compatibility
check.

## Source-time structures

The same S1-S4 scenario language is used across the Bayesian and frequentist
engines.

```{r echo=FALSE}
source_time_table <- tibble::tribble(
  ~Structure, ~Scenario, ~Latent_state, ~Bayesian_guidance,
  "One source, one time",
  "`s1`",
  "OD state, but each state is observed once",
  "Prefer `coverage_offset`; latent mode is weakly identified.",
  "One source, multiple times",
  "`s2`",
  "OD or OD-time state, depending on the research question",
  "Use care: time variation can reflect true mobility change rather than observation bias.",
  "Multiple sources, one time",
  "`s3`",
  "OD state shared across sources",
  "Best first use case for the approved advanced `latent_two_level` backend.",
  "Multiple sources, multiple times",
  "`s4`",
  "OD-time state shared across sources within each time period",
  "Useful but more demanding; inspect source and time diagnostics carefully."
)

knitr::kable(source_time_table)
```

To move the latent example from S3 to S4, use the full repeated source/time
table and share latent states within OD-time groups.

```{r eval=FALSE}
adj_latent_s4 <- debiasR::adjust_multilevel_bayes(
  mpd_od_df = mpd_s4,
  coverage_df = coverage_s4,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "latent_two_level",
  backend = "auto",
  coverage_scale = "origin",
  latent_flow_unit = "od_time",
  model_engine = "bayesian",
  scenario = "s4",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source_time",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

## Reduced-form compatibility mode

The reduced-form mode preserves the earlier observed-flow workflow:

```{r eval=FALSE}
adj_reduced_form <- debiasR::adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "mpd_counterfactual",
  observation_model = "reduced_form",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "observed",
  random_intercept = "none",
  model_family = "poisson",
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```

Use this mode only when you need to reproduce older analyses or compare the
newer true-flow scale against the earlier MPD-scale counterfactual. For most
new analyses, prefer `coverage_offset`.

## Diagnostics

For any Bayesian fit, inspect:

- sampler diagnostics from `attr(result, "diagnostics")`;
- model metadata from `attr(result, "result_metadata")`;
- model terms from `attr(result, "model_terms")`;
- posterior summaries and, when requested, draw-level outputs;
- whether the source/time structure supplies enough repeated observations for
  the chosen observation model.

```{r eval=FALSE}
attr(adj_latent, "result_metadata")
attr(adj_latent, "diagnostics")
attr(adj_latent, "model_terms")
```

For the custom latent backend, `attr(adj_latent, "diagnostics")$convergence`
reports divergence and treedepth rates, acceptance summaries, effective sample
size, and R-hat for core model parameters. The `latent_*_prior_scale`,
`latent_phi_prior_rate`, `latent_adapt_delta`, `latent_max_treedepth`, and
`latent_rng_eta_max` arguments are available for sensitivity checks when the
default weakly regularizing settings are not appropriate.

The latent backend should still be treated as an advanced model: use it when
the repeated-source structure is substantively important, keep prior checks in
the workflow, and inspect posterior predictive diagnostics before relying on
the fitted latent state.

## Current validation status

The default `coverage_offset` route is the approved Bayesian implementation for
observed-flow LAD workflows. In the empirical S4 LAD validation run, fixed and
origin random-intercept coverage-offset models completed on the full selected
LAD source/time table with acceptable sampler diagnostics. The Bayesian fits
were competitive with benchmark-calibrated deterministic methods on MAE and
strong on RMSE, while retaining the main Bayesian advantage: the benchmark OD
table was used for validation rather than fitting.

This approval is intentionally scoped. It supports `coverage_offset` as the
default empirical Bayesian route for S1-S4 observed-flow workflows. It also
approves `latent_two_level` as an advanced repeated-source S3/S4 route after
real MPD/Census empirical runs, real centroid distances, prior-sensitivity
checks, and confirmatory sampler diagnostics. The `reduced_form` route remains
useful for compatibility and sensitivity checks, but it is not the recommended
implementation for new Bayesian adjustment work.

## Practical recommendation

Use `coverage_offset` as the default Bayesian implementation. It is readable,
validated for the default empirical LAD workflow, and directly returns a
true-flow scale from a coverage-aware observation model.

Use `latent_two_level` when the data structure justifies it: especially
multiple MPD sources observing the same OD or OD-time state. Keep the older
`reduced_form` mode for reproducibility and backward compatibility.
