---
title: "Advanced Bayesian multilevel adjustment"
vignette: >
  %\VignetteIndexEntry{Advanced Bayesian multilevel adjustment}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
main table {
  font-size: 0.9rem;
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}
</style>
```

```{r setup, echo=FALSE}
library(debiasR)
```

## Aim

This vignette explains the advanced model variants in
`adjust_multilevel_bayes()`. The main adjustment vignette introduces the
coverage-offset model variant first because it is the most practical starting
point: take the observed MPD flow, use measured coverage as an observation
offset, and predict the underlying true-flow scale. Start with
[Adjusting origin-destination flows](v06-adjusting-biases.html) if you want that
default workflow first.

Here we slow down and look across the model variants. The useful question is
what you think the MPD data are measuring. Sometimes you have one source and one
time period, and the coverage-offset model variant is the clearest choice.
Sometimes more than one source observes the same OD or OD-time pattern, and it
becomes possible to estimate a shared hidden flow behind those observations.
Sometimes you need to reproduce the earlier MPD-scale correction logic, where
the model asks what the fitted MPD flow would look like if the fitted bias term
were neutral.

The guiding principle is progressive complexity: use the simplest model that
matches the data structure and inferential goal.

## Why more than one model variant?

`adjust_multilevel_bayes()` provides three observation-model variants that
connect observed MPD flows to adjusted flows. They are related, but they do not
answer exactly the same question.

The coverage-offset and latent two-level model variants both return a
true-flow scale, but they get there differently. The coverage-offset model
variant uses coverage as a known adjustment in the observation equation. The
latent model variant estimates a hidden state and is only really interesting
when repeated MPD rows give the model something to learn about that state.

The reduced-form model variant is different. It stays on the MPD observation
scale and asks for a counterfactual fitted MPD flow. That makes it valuable for
continuity with earlier analyses, but it should not be read as the same
true-flow model as the coverage-offset model variant.

There is also a frequentist option, `model_engine = "frequentist"`, for
non-latent models. It is not a Bayesian model. It is a practical preflight
check: use it to test joins, formulas, source/time declarations and count
families before paying the runtime cost of posterior sampling.

## Choosing a Bayesian model

```{r echo=FALSE}
decision_table <- tibble::tribble(
  ~If_your_data_or_goal_looks_like_this, ~Start_with, ~Why,
  "One source and one time period, or a small teaching example",
  "`coverage_offset`",
  "The latent split is weakly identified when each OD pair is observed once.",
  "Complete-grid prediction over supplied OD rows",
  "`coverage_offset`",
  "The model can predict true-flow scale for supplied rows using the coverage offset and mobility formula.",
  "Multiple sources observing the same OD pairs",
  "`latent_two_level`",
  "Repeated source observations can share an OD-level latent state.",
  "Multiple sources and multiple time periods",
  "`latent_two_level` with care",
  "Repeated source observations can share OD-time states, but time and source effects need careful interpretation.",
  "Backward compatibility with earlier reduced-form analyses",
  "`reduced_form`",
  "This preserves the older MPD-scale counterfactual behaviour."
)

knitr::kable(decision_table)
```

As a rule of thumb, start with `coverage_offset`. Move to
`latent_two_level` only when the repeated-observation structure gives the model
real information about the hidden flow state.

## Source-time structures

The same S1-S4 language is used throughout the function. It is simply a compact
way to say how much repeated observation structure the MPD data contain.

```{r echo=FALSE}
source_time_table <- tibble::tribble(
  ~Structure, ~Scenario, ~What_the_model_sees, ~Practical_guidance,
  "One source, one time",
  "`s1`",
  "One MPD row for each observed OD pair.",
  "Use `coverage_offset` unless you have a specific reason to run a sensitivity check.",
  "One source, multiple times",
  "`s2`",
  "The same source is observed across periods.",
  "Be careful: time variation may be real mobility change, not only observation bias.",
  "Multiple sources, one time",
  "`s3`",
  "More than one source observes the same OD pairs.",
  "`latent_two_level` becomes useful because sources can share an OD state.",
  "Multiple sources, multiple times",
  "`s4`",
  "Source-time rows can observe the same OD or OD-time structure.",
  "Useful for latent modelling, but source and time effects need careful diagnostics."
)

knitr::kable(source_time_table)
```

## How the model variants differ

The three Bayesian observation-model variants answer related but different
questions. They should not all be read through the two-equation
coverage-offset explanation used in the main adjustment vignette.

The coverage-offset model variant asks: what true OD flow would have produced the MPD
flow we observed, given measured coverage? The reduced-form model variant asks: what
would the fitted MPD flow look like if the fitted bias term were neutral? The
latent model variant asks: what shared hidden OD or OD-time flow state is implied by
repeated source/time observations?

### Coverage-offset true-flow model variant

The coverage-offset model variant treats the MPD count as a partial observation of an
underlying true flow. Its observation equation is:

$$
Y^{mpd}_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) = \log(q_{ijst}) + \eta^{true}_{ijst}
$$

where $Y^{mpd}_{ijst}$ is the observed MPD flow for origin $i$, destination
$j$, source $s$ and time $t$; $q_{ijst}$ is the coverage-derived observation
probability; and $\eta^{true}_{ijst}$ is the true-flow linear predictor. In a
simple S1 setting, the source and time subscripts can be dropped:

$$
\log(\mu_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}
$$

The true-flow predictor can include origin covariates, destination covariates,
distance, and optional random-effect terms:

$$
\eta^{true}_{ij}
= \alpha + X_i\beta_o + X_j\beta_d + \gamma \log(d_{ij}) + u_i + v_j + w_{ij}
$$

The adjusted flow is the prediction after removing the fixed coverage offset:

$$
F^{adj}_{ij} = \exp(\eta^{true}_{ij})
$$

This model variant is not a latent-state model in the strict sense. The
true-flow scale is recovered by fitting a count model with a fixed observation
offset and then predicting with that offset set to zero. If source or time terms
are included in the formula, they are fitted as predictors in this non-latent
model; the variant does not force one shared source-invariant OD state unless
the formula and data structure imply that restriction.

### Reduced-form compatibility model variant

The reduced-form model variant preserves the earlier observed-flow correction logic. It
fits the MPD observations directly:

$$
Y^{mpd}_{ijst} \sim \mathrm{Poisson}(\mu^{mpd}_{ijst})
$$

$$
\log(\mu^{mpd}_{ijst}) =
\eta^{mobility}_{ijst} + \eta^{bias}_{ijst}
$$

The mobility part can contain OD covariates, distance and source/time terms.
The bias part contains fitted bias covariates, such as an origin coverage-bias
term:

$$
\eta^{bias}_{ijst} = \delta b_i
$$

The adjustment is a counterfactual prediction in which the fitted bias term is
set to a neutral value:

$$
F^{adj}_{ij} =
\exp(\eta^{mobility}_{ij} + \eta^{bias}_{ij}\mid b_i = 0)
$$

This means the reduced-form model variant returns an MPD-scale counterfactual, not a
true-flow state identified through a coverage offset. It is useful
for reproducing earlier analyses and for sensitivity checks, but it should not
be described as the same observation-process model as `coverage_offset`.

### Latent two-level model variant

The latent two-level model variant is the advanced model that most directly matches
the phrase "two-level" in a statistical sense. It introduces a shared hidden
true-flow state and treats repeated source/time MPD rows as observations of
that state:

$$
Y^{mpd}_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) =
\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma
$$

where $\theta_{g(i,j,t)} = \log(\lambda^{true}_{g(i,j,t)})$ is the log latent
true-flow intensity for an OD or OD-time group, and $z_{ijst}\gamma$ captures
observation-layer structure such as source and time effects. The latent group is
controlled by `latent_flow_unit`: `latent_flow_unit = "od"` shares one state
for each OD pair, while `latent_flow_unit = "od_time"` shares one state for
each OD-time pair.

The adjusted true-flow prediction is:

$$
F^{adj}_{g} = \lambda^{true}_{g} = \exp(\theta_g)
$$

The fitted MPD observation-scale prediction keeps the source/time observation
layer:

$$
F^{mpd,pred}_{ijst} =
\exp(\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma)
$$

This model variant is most informative when repeated observations make the latent state
empirically meaningful. In S3, multiple sources observe the same OD state. In
S4, multiple source/time rows can share OD-time states. In S1, each state is
observed only once, so the latent split is weakly identified and the
coverage-offset model variant is usually clearer.

## Simulated source-time example

The examples below use a small simulated dataset so the calls can be read and
adapted without requiring external empirical files. The simulation creates five
areas, a true OD-flow surface, two MPD sources and two time periods. Coverage
differs by area, source and time. The observed MPD rows are generated as
coverage-scaled measurements of the true flows.

```{r simulated-source-time-data, echo=FALSE}
set.seed(20260625)

areas <- paste0("area_", LETTERS[1:5])
area_profile <- data.frame(
  area = areas,
  area_index = seq_along(areas),
  population = c(900, 1200, 760, 1500, 1080),
  rural_pct = c(0.10, 0.25, 0.55, 0.40, 0.70),
  stringsAsFactors = FALSE
)

od_grid <- expand.grid(
  origin = areas,
  destination = areas,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

origin_profile <- area_profile
names(origin_profile) <- paste0(names(origin_profile), "_o")
names(origin_profile)[names(origin_profile) == "area_o"] <- "origin"

destination_profile <- area_profile
names(destination_profile) <- paste0(names(destination_profile), "_d")
names(destination_profile)[names(destination_profile) == "area_d"] <- "destination"

od_truth <- merge(od_grid, origin_profile, by = "origin")
od_truth <- merge(od_truth, destination_profile, by = "destination")
od_truth$distance_km <- abs(od_truth$area_index_o - od_truth$area_index_d) + 1
od_truth$log_distance <- log(od_truth$distance_km)
od_truth$self_flow <- as.numeric(od_truth$origin == od_truth$destination)
od_truth$eta_true <- 5.6 +
  0.45 * od_truth$rural_pct_o -
  0.25 * od_truth$rural_pct_d -
  0.70 * od_truth$log_distance +
  0.90 * od_truth$self_flow
od_truth$flow_true <- round(exp(od_truth$eta_true))

source_time_grid <- expand.grid(
  mpd_source = c("operator_a", "operator_b"),
  mpd_time = c("2021_q1", "2021_q2"),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

coverage_repeated <- merge(
  area_profile[c("area", "population")],
  source_time_grid
)
names(coverage_repeated)[names(coverage_repeated) == "area"] <- "origin"
coverage_repeated$coverage_rate <- with(
  coverage_repeated,
  ifelse(mpd_source == "operator_a", 0.09, 0.13) +
    ifelse(mpd_time == "2021_q2", 0.015, 0) +
    c(0.005, -0.004, 0.002, -0.003, 0.004)[match(origin, areas)]
)
coverage_repeated$user_count <- pmax(
  1,
  round(coverage_repeated$population * coverage_repeated$coverage_rate)
)

mpd_repeated <- merge(
  od_truth[c("origin", "destination", "flow_true")],
  source_time_grid
)
mpd_repeated <- merge(
  mpd_repeated,
  coverage_repeated[c("origin", "mpd_source", "mpd_time", "coverage_rate")],
  by = c("origin", "mpd_source", "mpd_time")
)
mpd_repeated$source_time_effect <- with(
  mpd_repeated,
  ifelse(mpd_source == "operator_b", 1.12, 0.95) *
    ifelse(mpd_time == "2021_q2", 1.04, 1.00)
)
mpd_repeated$flow <- stats::rpois(
  nrow(mpd_repeated),
  lambda = pmax(
    0.1,
    mpd_repeated$coverage_rate *
      mpd_repeated$flow_true *
      mpd_repeated$source_time_effect
  )
)
mpd_repeated <- mpd_repeated[
  c("origin", "destination", "flow", "mpd_source", "mpd_time")
]

coverage <- coverage_repeated[
  c("origin", "population", "user_count", "mpd_source", "mpd_time")
]
covariates <- area_profile[c("area", "rural_pct", "population")]
distance <- od_truth[c("origin", "destination", "distance_km")]

mpd_s1 <- subset(
  mpd_repeated,
  mpd_source == "operator_a" & mpd_time == "2021_q1"
)
coverage_s1 <- subset(
  coverage,
  mpd_source == "operator_a" & mpd_time == "2021_q1"
)

mpd_s3 <- subset(mpd_repeated, mpd_time == "2021_q1")
coverage_s3 <- subset(coverage, mpd_time == "2021_q1")

mpd_s4 <- mpd_repeated
coverage_s4 <- coverage

knitr::kable(head(mpd_s4, 8))
```

The S1 object contains one source and one time period. The S3 object contains
two sources for one time period. The S4 object contains two sources and two time
periods. These small objects are intentionally simple, but they exercise the
same parameter contracts used by empirical source/time flow data.

```{r echo=FALSE}
example_objects <- tibble::tribble(
  ~Object, ~Scenario, ~Description,
  "`mpd_s1`, `coverage_s1`",
  "`s1`",
  "One source and one time period.",
  "`mpd_s3`, `coverage_s3`",
  "`s3`",
  "Two sources observing the same OD pairs in one period.",
  "`mpd_s4`, `coverage_s4`",
  "`s4`",
  "Two sources observed across two periods."
)

knitr::kable(example_objects)
```

The Bayesian model-fitting chunks below are useful to run interactively when
you want posterior summaries. They are not evaluated when the vignette is
rendered, and you should inspect `attr(result, "diagnostics")` before
interpreting the adjusted flows. The frequentist option is evaluated in this
vignette because it is fast and is useful for confirming that the simulated
source/time structure has been declared correctly.

## Default coverage-offset model

The default true-flow model is:

$$
F^{mpd}_{ij} \sim \mathrm{Poisson}(q_{ij}F^{true}_{ij})
$$

$$
\log E(F^{mpd}_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}
$$

where $q_{ij}$ is derived from active-user coverage. With
`coverage_scale = "origin"`, $q_{ij} = c_i = U_i / P_i$. The returned
`flow_adj` is a posterior summary of:

$$
F^{true,pred}_{ij} = \exp(\eta^{true}_{ij})
$$

This is the model explained in the main adjustment vignette. It is usually the
best option for S1 examples and for users who want readable, reproducible
Bayesian adjustment without modelling source-specific observation layers.

```{r eval=FALSE}
adj_default <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "observed",
  random_intercept = "none",
  model_family = "poisson",
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```

With richer data, the coverage-offset model can include multilevel structure in
the true-flow formula. In an S1 setting, that usually means origin and
destination pooling: `(1 | origin)` for an origin intercept and
`(1 | destination)` for a destination intercept. Use these terms when the data
contain enough repeated information to estimate the pooling structure.

```{r eval=FALSE}
adj_default_multilevel <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance +
    (1 | origin) + (1 | destination),
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

## Frequentist option for specification checks

Before running the Bayesian examples, it is often useful to check the same data
contract with the frequentist option. The code below uses the S4 simulated
data, declares source and time columns, and fits the coverage-offset model
variant without posterior sampling. This is a practical way to test formulas,
covariates, count families and source/time declarations before paying the MCMC
runtime cost.

```{r frequentist-s4-check}
adj_fast_s4 <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s4,
  coverage_df = coverage_s4,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance +
    mpd_source + mpd_time,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "frequentist",
  scenario = "s4",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source_time",
  prediction_scope = "observed",
  random_intercept = "none",
  model_family = "poisson"
)

knitr::kable(head(adj_fast_s4, 8), digits = 2)
```

The frequentist result should be interpreted as a fast design check. It tells
you whether the scenario, coverage joins, formulas and prediction scope are
working, but it is not posterior inference.

Once that preflight passes, the same S4 structure can be fitted with the
Bayesian engine. This chunk is useful to run interactively because MCMC runtime
depends on the data size and local sampler setup.

```{r eval=FALSE}
adj_bayes_s4 <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s4,
  coverage_df = coverage_s4,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance +
    mpd_source + mpd_time + (1 | origin),
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s4",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source_time",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

## Latent two-level backend

The latent two-level backend is for repeated source/time settings. It keeps
the true-flow target but introduces a shared latent-flow state. Conceptually:

$$
Y_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) =
\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma
$$

where:

- $Y_{ijst}$ is the observed MPD flow for origin $i$, destination $j$, source
  $s$ and time $t$.
- $q_{ijst}$ is the coverage-derived observation probability.
- $\theta_{g(i,j,t)} = \log(\lambda^{true}_{g(i,j,t)})$ is the log shared
  latent true-flow intensity for an OD or OD-time group.
- $z_{ijst}\gamma$ represents observation-layer structure such as source and
  time effects.

In this backend, `adjust_multilevel_bayes()` represents the shared state through
`latent_flow_id` and estimates source-invariant latent effects with an
identified sum-to-zero parameterization. The returned `flow_adj` and
`flow_true_pred` summarize posterior draws of the latent true-flow intensity,
while `flow_mpd_pred` summarizes the source/time-specific MPD observation mean.

```{r eval=FALSE}
adj_latent <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s3,
  coverage_df = coverage_s3,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "latent_two_level",
  backend = "auto",
  coverage_scale = "origin",
  latent_flow_unit = "auto",
  model_engine = "bayesian",
  scenario = "s3",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

The returned table includes `latent_flow_id`, `latent_flow_unit`,
`flow_adj_mean`, `flow_adj_median`, and 95% interval columns for the true-flow
and MPD observation scales. For most users, these are the first things to
check: which rows share a latent state, what true-flow summary the model
returns, and how the source/time-specific MPD prediction compares with the
observed MPD flow.
Because this mode relies on repeated observations, it is most informative for
S3 and S4 designs. In S1, the latent split is usually too weak. In S2, use it
only when treating repeated time periods as observations of a stable underlying
OD state is substantively defensible.

To move the latent example from S3 to S4, use the full repeated source/time
table and share latent states within OD-time groups.

```{r eval=FALSE}
adj_latent_s4 <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s4,
  coverage_df = coverage_s4,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "latent_two_level",
  backend = "auto",
  coverage_scale = "origin",
  latent_flow_unit = "od_time",
  model_engine = "bayesian",
  scenario = "s4",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source_time",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

## Reduced-form compatibility mode

The reduced-form mode preserves the earlier observed-flow workflow:

```{r eval=FALSE}
adj_reduced_form <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "mpd_counterfactual",
  observation_model = "reduced_form",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "observed",
  random_intercept = "none",
  model_family = "poisson",
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```

Use this mode when you need to reproduce older analyses or compare the newer
true-flow scale against the earlier MPD-scale counterfactual. For most new
analyses, the coverage-offset model variant is the clearer starting point.

## What the returned columns mean

The returned object is a tibble with the original OD identifiers, the observed
MPD flow, adjusted-flow summaries, and model metadata attached as attributes.
The same column name can carry slightly different meaning across model variants,
so it is worth reading the output through the model you fitted.

```{r echo=FALSE}
output_columns <- tibble::tribble(
  ~Column_or_attribute, ~Coverage_offset, ~Latent_two_level, ~Reduced_form,
  "`flow_adj`",
  "Posterior summary of the coverage-adjusted true-flow prediction.",
  "Posterior summary of the latent true-flow state.",
  "Posterior summary of the MPD-scale counterfactual with fitted bias neutralised.",
  "`flow_true_pred`",
  "Same true-flow-scale quantity as `flow_adj`.",
  "Same latent true-flow-scale quantity as `flow_adj`.",
  "`NA`, because this model variant is not a true-flow model.",
  "`flow_mpd_pred`",
  "Posterior prediction on the MPD observation scale, with the coverage offset included.",
  "Source/time-specific MPD observation mean implied by the latent state and observation layer.",
  "`NA` for the reduced-form counterfactual model variant.",
  "`latent_flow_id`",
  "Not used.",
  "Identifier for the shared OD or OD-time latent state.",
  "Not used.",
  "`attr(result, \"diagnostics\")`",
  "Sampler diagnostics for the fitted Bayesian count model.",
  "Sampler diagnostics for the custom latent Stan fit, including convergence summaries.",
  "Sampler diagnostics for the fitted Bayesian count model."
)

knitr::kable(output_columns)
```

## Diagnostics

For any Bayesian fit, inspect:

- sampler diagnostics from `attr(result, "diagnostics")`;
- model metadata from `attr(result, "result_metadata")`;
- model terms from `attr(result, "model_terms")`;
- posterior summaries and, when requested, draw-level outputs;
- whether the source/time structure supplies enough repeated observations for
  the chosen observation model.

```{r eval=FALSE}
attr(adj_latent, "result_metadata")
attr(adj_latent, "diagnostics")
attr(adj_latent, "model_terms")
```

For the custom latent backend, `attr(adj_latent, "diagnostics")$convergence`
reports divergence and treedepth rates, acceptance summaries, effective sample
size, and R-hat for core model parameters. The `latent_*_prior_scale`,
`latent_phi_prior_rate`, `latent_adapt_delta`, `latent_max_treedepth`, and
`latent_rng_eta_max` arguments are available for sensitivity checks when the
default weakly regularizing settings are not appropriate.

The latent backend should still be treated as an advanced model: use it when
the repeated-source structure is substantively important, keep prior checks in
the workflow, and inspect sampler diagnostics and prediction summaries before
relying on the fitted latent state.

## Reading empirical comparisons

When you compare Bayesian adjustment with deterministic methods, it helps to
remember what information each method is allowed to use. Raking,
selection-rate calibration and coefficient calibration can use benchmark
margins or benchmark OD targets while they are being fitted. The
coverage-offset Bayesian model variant does not need benchmark OD flows for
fitting, so the benchmark can be held back for validation.

That changes the interpretation of the comparison. The aim is not simply to win
the lowest MAE in every table. If a coverage-offset Bayesian model gets close
to benchmark-trained methods while fitting without the benchmark OD targets,
that is already a useful result. It means the coverage-aware observation model
is recovering much of the same flow structure from MPD counts, coverage and
covariates alone.

The latent model variant should be read in the same spirit, but with one extra
question: do the repeated source/time observations give the model enough
information to separate the hidden flow state from the observation layer? In S3
and S4 data, that question is empirical. In S1 data, the answer is usually no,
because each OD flow is only observed once.

## Putting the model variants together

In most applications, begin with `coverage_offset`. It gives a direct
true-flow-scale adjustment and keeps the role of active-user coverage explicit.

Move to `latent_two_level` when the data structure earns the extra complexity:
more than one MPD source, or source-time observations that can plausibly share
an underlying OD or OD-time flow.

Keep `reduced_form` for continuity with earlier analyses and for sensitivity
checks. It is still useful, but its adjusted value is an MPD-scale
counterfactual, so it should not be interpreted as the same true-flow quantity
returned by the coverage-offset and latent model variants.
