---
title: "Advanced Bayesian multilevel adjustment"
vignette: >
  %\VignetteIndexEntry{Advanced Bayesian multilevel adjustment}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

```{=html}
<style>
main table {
  font-size: 0.9rem;
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}
</style>
```

```{r setup, echo=FALSE}
library(debiasR)
```

```{r load-lad-bayesian-results, include=FALSE}
v09_find_extdata_file <- function(filename) {
  package_path <- system.file("extdata", filename, package = "debiasR")
  if (!identical(package_path, "")) {
    return(package_path)
  }

  candidates <- c(
    file.path("inst", "extdata", filename),
    file.path("..", "inst", "extdata", filename),
    file.path("..", "..", "inst", "extdata", filename)
  )
  candidates <- candidates[file.exists(candidates)]
  if (length(candidates) == 0L) {
    stop("Could not find the v09 LAD Bayesian result file: ", filename, ".")
  }
  candidates[1]
}

v09_bayesian_output <- tibble::as_tibble(
  utils::read.csv(
    v09_find_extdata_file("v09-bayesian-real-data-output.csv"),
    check.names = FALSE
  )
)

v09_model_summary <- tibble::as_tibble(
  utils::read.csv(
    v09_find_extdata_file("v09-bayesian-real-data-summary.csv"),
    check.names = FALSE
  )
)

v09_diagnostics <- tibble::as_tibble(
  utils::read.csv(
    v09_find_extdata_file("v09-bayesian-real-data-diagnostics.csv"),
    check.names = FALSE
  )
)

v09_kable <- function(x, digits = NULL) {
  args <- list(
    x = x,
    format = "html",
    row.names = FALSE,
    table.attr = 'class="table table-sm"'
  )
  if (!is.null(digits)) {
    args$digits <- digits
  }
  do.call(knitr::kable, args)
}
```

## Aim

The previous adjustment vignette introduced Bayesian multilevel
modelling as one of the adjustment methods implemented in `debiasR`.
This vignette describes the Bayesian model options in more detail. The
aim is to explain how `adjust_multilevel_bayes()` can be used to adjust
mobile-phone-derived origin-destination (OD) flows, what data inputs the
function expects, which parameters define each model variant and how to
interpret the outputs returned by `debiasR`.

The vignette focuses on the three `observation_model` models:
`coverage_offset`, `latent_two_level` and `reduced_form`. These models
share a common purpose. They use observed MPD flows, coverage, area
characteristics and OD distance to produce adjusted flows from MPD. The
three models differ in how they produce adjusted flows. The vignette
also explains `model_engine = "frequentist"` as a deterministic way to
test data structures, functional forms and covariates before fitting a
Bayesian model. The examples use the local authority district (LAD) data
supplied via `debiasR_example_data()`.

## LAD data inputs

The Bayesian adjustment function needs an OD table with observed MPD
flows. Depending on the model variant, it can also use coverage,
covariates and distances. The LAD example prepares a complete OD grid
so every origin-destination pair has a row.

```{r lad-inputs}
lad_data <- debiasR_example_data(
  n_areas = Inf,
  complete_grid = TRUE,
  geography = "lad"
)

mpd_od_df <- lad_data$mpd_od
mpd_od_df$mpd_time <- "single_period"

coverage_df <- lad_data$coverage
coverage_df$mpd_time <- "single_period"

covariates_df <- lad_data$covariates
distance_df <- lad_data$distance
benchmark_od_df <- lad_data$benchmark_od
```

`mpd_od_df` contains the observed MPD flows. `flow` is the observed MPD
count for an origin-destination pair. `mpd_source` and `mpd_time`
identify the digital source and period represented by the row.

```{r lad-mpd-input}
head(mpd_od_df)
```

`coverage_df` contains the estimated MPD coverage for each area, source
and period. In this example, `coverage_rate` is the share of the
benchmark population observed in the MPD source.

```{r lad-coverage-input}
head(coverage_df)
```

`covariates_df` provides origin and destination characteristics that can
enter the mobility model. `distance_df` supplies the OD distance used in
the gravity component of the model.

```{r lad-covariate-distance-input}
head(covariates_df)
head(distance_df)
```

`benchmark_od_df` is not used to fit the Bayesian adjustment model. It
is used after adjustment to validate the adjusted flows against an
external benchmark.

```{r lad-benchmark-input}
head(benchmark_od_df)
```

## Bayesian options

`adjust_multilevel_bayes()` exposes three Bayesian model variants
through the `observation_model` argument.

```{r bayesian-options-table, echo=FALSE}
bayesian_options <- tibble::tribble(
  ~`Function argument`, ~`Main question`, ~`Adjusted flow scale`, ~`When to start here`,
  '`observation_model = "coverage_offset"`',
  "What true-flow pattern is most consistent with the observed MPD counts, after accounting for coverage?",
  "`flow_adj` and `flow_true_pred` are on the true-flow scale.",
  "Use this as the main starting point when coverage rates are available.",
  '`observation_model = "latent_two_level"`',
  "What shared hidden OD or OD-time flow is consistent with repeated source/time observations?",
  "`flow_adj` and `flow_true_pred` summarize the latent true-flow intensity.",
  "Use this when the same OD or OD-time flow is observed through real repeated source or time rows.",
  '`observation_model = "reduced_form"`',
  "What fitted MPD flow would be expected if the fitted bias term were neutral?",
  "`flow_adj` is an MPD-scale counterfactual.",
  "Use this for compatibility checks or sensitivity analyses on the MPD scale."
)

v09_kable(bayesian_options)
```

The function also offers a deterministic fitting option:
`model_engine = "frequentist"`. This is not a Bayesian model. It is a
useful way to test joins, formulas, functional forms, covariates and
count-family choices before fitting a Bayesian model. The frequentist
option is available for `coverage_offset` and `reduced_form`;
`latent_two_level` requires `model_engine = "bayesian"`.

## Model equations

This section gives the general equations behind the three model
variants. The notation is deliberately close to the `debiasR` inputs and
outputs:

- \(Y_{ij}\) is the observed MPD `flow` from origin \(i\) to
  destination \(j\).
- \(Y_{ijst}\) is the observed MPD `flow` for origin \(i\),
  destination \(j\), source \(s\) and time period \(t\).
- \(q\) is the `observation_probability` derived from the selected
  `coverage_scale`.
- \(F^{true}\) is the true-flow intensity targeted by the true-flow model
  variants.
- `flow_mpd_pred` is the fitted MPD-scale prediction.
- `flow_true_pred` is the fitted true-flow-scale prediction where that
  scale is defined by the selected model variant.
- `flow_adj` is the adjusted flow returned by
  `adjust_multilevel_bayes()`.

The equations are there to clarify what each function option is doing.
The key idea is simple: each model starts with an observed MPD count,
then decides whether to treat that count as a coverage-scaled
observation of a true flow, a repeated observation of a shared latent
flow, or an MPD-scale counterfactual.

### `coverage_offset`

The `coverage_offset` model treats the observed MPD flow as a
coverage-scaled observation of a true OD flow:

\[ Y_{ij} \sim \operatorname{Poisson}(q_{ij} F^{true}_{ij}) \]

\[ \log{E(Y_{ij})} = \log(q_{ij}) + \eta^{true}_{ij} \]

\[ \eta^{true}_{ij} = \alpha + X_i \beta_o + X_j \beta_d +
\gamma \log(d_{ij}) + u_i \]

Here, \(X_i\) and \(X_j\) are origin and destination covariates,
\(d_{ij}\) is distance and \(u_i\) is an optional origin random
intercept. The offset \(\log(q_{ij})\) enters the observation equation,
so the model estimates
a true-flow intensity and then maps it back to the MPD observation
scale.

For this model variant, `flow_adj` is the adjusted true-flow prediction.
It is reported on the same scale as `flow_true_pred`. `flow_mpd_pred`
keeps the coverage-adjusted fitted value on the MPD observation scale.

### `latent_two_level`

The `latent_two_level` model is designed for repeated observations. For
example, the same OD pair may be observed by more than one MPD source,
or the same OD pair may be observed across several time periods. The
model separates a shared hidden flow from the source/time observation
process:

\[ Y_{ijst} \sim \operatorname{Poisson}(\mu_{ijst}) \]

\[ \log(\mu_{ijst}) = \log(q_{ijst}) + \theta_{g(i,j,t)} +
z_{ijst}\gamma \]

\[ \theta_{g(i,j,t)} \sim
\operatorname{Normal}(\eta^{true}_{g(i,j,t)}, \sigma_\theta) \]

The term \(\theta_{g(i,j,t)}\) is the latent true-flow state shared by
the rows that refer to the same OD or OD-time unit. The argument
`latent_flow_unit` controls whether that state is defined at the OD
level or the OD-time level. The returned column `latent_flow_id`
identifies the shared state.

For this model variant, `flow_adj` and `flow_true_pred` summarize the
latent true-flow intensity. `flow_mpd_pred` remains source/time-specific
because it includes the observation layer.

### `reduced_form`

The `reduced_form` model stays on the MPD observation scale. It models
the observed MPD flow directly:

\[ Y_{ij} \sim \operatorname{Poisson}(\lambda^{mpd}_{ij}) \]

\[ \log(\lambda^{mpd}_{ij}) = \alpha + X_i \beta_o + X_j \beta_d +
\gamma \log(d_{ij}) + B_{ij}\delta \]

The term \(B_{ij}\delta\) represents the fitted bias component supplied
through `bias_formula`. The adjusted value is a counterfactual MPD
prediction in which the fitted bias component is set to its neutral
value.

For this model variant, `flow_adj` is an MPD-scale counterfactual. It
should not be interpreted as the same true-flow quantity returned by
`coverage_offset` or `latent_two_level`.

## Source and time structures

Origin-destination flows can be produced by different digital sources
and reported in different time units. One provider may supply daily OD
flows, another may supply weekly flows, and a research project may
aggregate either of them to a monthly or study-period total.
`adjust_multilevel_bayes()` uses `source_col`, `time_col`, `scenario`
and `repeated_observation` to describe this structure.

```{r source-time-table, echo=FALSE}
source_time_table <- tibble::tribble(
  ~`Data structure`, ~`Function arguments`, ~Interpretation,
  "One source, one time period",
  '`scenario = "s1"`, `repeated_observation = "none"`',
  "Each OD pair appears once. This is the LAD structure used below.",
  "One source, multiple time periods",
  '`scenario = "s2"`, `repeated_observation = "time"`',
  "The same source observes OD flows repeatedly over time.",
  "Multiple sources, one time period",
  '`scenario = "s3"`, `repeated_observation = "source"`',
  "Several sources observe the same OD flow for the same period.",
  "Multiple sources, multiple time periods",
  '`scenario = "s4"`, `repeated_observation = "source_time"`',
  "Several sources observe OD flows across several periods."
)

v09_kable(source_time_table)
```

The source/time structure matters most for `latent_two_level`. A latent
model needs repeated rows that genuinely observe the same underlying OD
or OD-time flow. In contrast, `coverage_offset` can be fitted to a
single source and a single period, as in the LAD example below.

## Implementation on LAD data

This section demonstrates how the model variants are called and how the
returned objects should be read. The implementation uses the LAD inputs
created above.

### Check the data and formula first

A useful first step is to fit the `coverage_offset` model with
`model_engine = "frequentist"`. This tests the data joins, formula and
output structure before the Bayesian fit.

```{r coverage-offset-frequentist-check}
adj_frequentist <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od_df,
  coverage_df = coverage_df,
  covariates_df = covariates_df,
  distance_df = distance_df,
  mobility_formula = ~ log_pop_o + log_pop_d + log_distance +
    rural_pct_o + rural_pct_d + per_level4_o + per_level4_d,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "frequentist",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson"
)

head(
  adj_frequentist[
    c(
      "origin", "destination", "flow", "flow_adj",
      "flow_mpd_pred", "flow_true_pred", "observation_probability"
    )
  ]
)
```

The returned columns are the same column names users see from
`adjust_multilevel_bayes()`. In this example, `flow` is the observed MPD
count, `observation_probability` is the coverage-derived observation
probability, `flow_mpd_pred` is the fitted MPD-scale value, and
`flow_adj` is the adjusted true-flow prediction.

### `coverage_offset` with `model_engine = "bayesian"`

The Bayesian `coverage_offset` call uses the same data and formula, but
returns posterior summaries for the adjusted flows. The origin random
intercept `(1 | origin)` allows the model to partially pool origins with
similar evidence.

```{r coverage-offset-bayesian-call, eval=FALSE}
adj_coverage_offset <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od_df,
  coverage_df = coverage_df,
  covariates_df = covariates_df,
  distance_df = distance_df,
  mobility_formula = ~ log_pop_o + log_pop_d + log_distance +
    rural_pct_o + rural_pct_d + per_level4_o + per_level4_d +
    (1 | origin),
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "origin",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 20260630,
  refresh = 0
)
```

The output below shows the column names returned by `debiasR` for the
LAD `coverage_offset` fit.

```{r coverage-offset-bayesian-output, echo=FALSE}
coverage_output_columns <- c(
  "origin", "destination", "flow", "flow_adj", "flow_mpd_pred",
  "flow_true_pred", "flow_adj_mean", "flow_adj_median",
  "flow_adj_q2.5", "flow_adj_q97.5", "observation_probability"
)

v09_kable(v09_bayesian_output[coverage_output_columns], digits = 2)
```

```{r coverage-offset-first-row, include=FALSE}
coverage_first_row <- v09_bayesian_output[1, ]
```

The first displayed row is the LAD flow from
`r coverage_first_row$origin` to `r coverage_first_row$destination`.
The observed MPD `flow` is
`r formatC(coverage_first_row$flow, digits = 0, format = "f", big.mark = ",")`.
The model estimates a larger `flow_adj` because the
`observation_probability` is about
`r formatC(coverage_first_row$observation_probability, digits = 3, format = "f")`:
only a share of the benchmark population is observed in the MPD source.
The posterior interval is reported through `flow_adj_q2.5` and
`flow_adj_q97.5`.

The fit also carries metadata and diagnostics as attributes:

```{r coverage-offset-diagnostics, echo=FALSE}
coverage_diagnostics <- v09_diagnostics[
  v09_diagnostics$object == "coverage_offset_full_lad",
  c("attribute", "field", "value")
]

v09_kable(coverage_diagnostics)
```

These rows correspond to the information users access with:

```{r coverage-offset-attributes, eval=FALSE}
attr(adj_coverage_offset, "result_metadata")
attr(adj_coverage_offset, "diagnostics")
attr(adj_coverage_offset, "model_terms")
```

The same adjusted object can be compared with the benchmark OD flows:

```{r coverage-offset-validation, eval=FALSE}
validate_flow_overall(adj_coverage_offset, benchmark_od_df)
```

### `latent_two_level`

Use `latent_two_level` when the input data contain repeated observations
of the same underlying OD or OD-time flow. The LAD input above is an S1
data structure: one source and one period. That structure is appropriate
for `coverage_offset`, but it does not give a latent two-level model
repeated source/time evidence to separate a shared latent flow from
observation-specific variation.

For data with repeated source/time rows, the call has the same shape but
uses `observation_model = "latent_two_level"` and a repeated-observation
setting that matches the data.

```{r latent-two-level-call, eval=FALSE}
adj_latent_two_level <- adjust_multilevel_bayes(
  mpd_od_df = mpd_source_time_df,
  coverage_df = coverage_source_time_df,
  covariates_df = covariates_df,
  distance_df = distance_df,
  mobility_formula = ~ log_pop_o + log_pop_d + log_distance +
    rural_pct_o + rural_pct_d + per_level4_o + per_level4_d,
  bias_formula = ~ 0,
  target_scale = "true_flow",
  observation_model = "latent_two_level",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s4",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source_time",
  prediction_scope = "observed",
  latent_flow_unit = "od_time",
  model_family = "poisson",
  iter = 1000,
  chains = 4,
  seed = 20260630,
  refresh = 0
)
```

The key additional returned column is `latent_flow_id`. It identifies
the shared OD or OD-time state used by the latent model. Rows that share
a `latent_flow_id` share the same latent true-flow component, while
`flow_mpd_pred` can still differ across source/time rows because the
observation layer includes source/time-specific information.

```{r latent-two-level-diagnostics, echo=FALSE}
latent_diagnostics <- v09_diagnostics[
  v09_diagnostics$object == "latent_two_level_full_s3_s4_approval",
  c("attribute", "field", "value")
]

v09_kable(latent_diagnostics)
```

Users should inspect `attr(adj_latent_two_level, "diagnostics")`
carefully. For latent models, important fields include divergent
transitions, treedepth hits, effective sample sizes and (\widehat{R}).

### `reduced_form`

The `reduced_form` model is useful when users want continuity with
earlier MPD-scale analyses. It uses `bias_formula` to model the fitted
bias component and then returns an MPD-scale counterfactual through
`flow_adj`.

```{r reduced-form-call, eval=FALSE}
adj_reduced_form <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od_df,
  coverage_df = coverage_df,
  covariates_df = covariates_df,
  distance_df = distance_df,
  mobility_formula = ~ log_pop_o + log_pop_d + log_distance +
    rural_pct_o + rural_pct_d + per_level4_o + per_level4_d,
  bias_formula = ~ bias_e_origin,
  target_scale = "mpd_counterfactual",
  observation_model = "reduced_form",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 20260630,
  refresh = 0
)
```

A deterministic `reduced_form` check can be fitted on the same LAD
inputs.

```{r reduced-form-frequentist-check}
adj_reduced_form_check <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od_df,
  coverage_df = coverage_df,
  covariates_df = covariates_df,
  distance_df = distance_df,
  mobility_formula = ~ log_pop_o + log_pop_d + log_distance +
    rural_pct_o + rural_pct_d + per_level4_o + per_level4_d,
  bias_formula = ~ bias_e_origin,
  target_scale = "mpd_counterfactual",
  observation_model = "reduced_form",
  model_engine = "frequentist",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson"
)

head(
  adj_reduced_form_check[
    c("origin", "destination", "flow", "flow_adj", "flow_mpd_pred")
  ]
)
```

For `reduced_form`, `flow_adj` is on the MPD scale. It is not the same
target quantity as `flow_adj` from `coverage_offset` or
`latent_two_level`, where the adjusted flow is on a true-flow scale.

## Reading `adjust_multilevel_bayes()` results

The returned object is a data frame with OD identifiers, fitted values
and model-specific columns. Some columns appear for every model variant;
others are specific to the observation model or engine.

```{r returned-columns-table, echo=FALSE}
returned_columns <- tibble::tribble(
  ~Name, ~Where_it_appears, ~How_to_read_it,
  "`flow`",
  "All model variants",
  "Observed MPD flow from `mpd_od_df`.",
  "`flow_adj`",
  "All model variants",
  "Main adjusted flow returned by `debiasR`; its scale depends on `observation_model`.",
  "`flow_mpd_pred`",
  "All model variants",
  "Fitted prediction on the MPD observation scale.",
  "`flow_true_pred`",
  "`coverage_offset` and `latent_two_level`; present for some compatibility paths",
  "Fitted true-flow-scale prediction where the selected model defines a true-flow target.",
  "`flow_adj_mean`",
  "Bayesian fits when posterior flow summaries are available",
  "Posterior mean of the adjusted flow.",
  "`flow_adj_median`",
  "Bayesian fits when posterior flow summaries are available",
  "Posterior median of the adjusted flow.",
  "`flow_adj_q2.5`",
  "Bayesian fits when posterior flow summaries are available",
  "Lower 2.5 percent posterior quantile for the adjusted flow.",
  "`flow_adj_q97.5`",
  "Bayesian fits when posterior flow summaries are available",
  "Upper 97.5 percent posterior quantile for the adjusted flow.",
  "`observation_probability`",
  "`coverage_offset` and `latent_two_level`",
  "Coverage-derived probability used in the observation equation.",
  "`latent_flow_id`",
  "`latent_two_level`",
  "Identifier for the shared OD or OD-time latent state.",
  '`attr(result, "result_metadata")`',
  "All model variants",
  "Model choices, data structure and prediction scope recorded with the result.",
  '`attr(result, "diagnostics")`',
  "All model variants",
  "Fit diagnostics, including convergence information when available.",
  '`attr(result, "model_terms")`',
  "All model variants",
  "Resolved model terms used by the fitted model."
)

names(returned_columns) <- c("Name", "Where it appears", "How to read it")

v09_kable(returned_columns)
```

The most important interpretation rule is to read `flow_adj` together
with `observation_model`:

```{r model-summary-table, echo=FALSE}
model_summary_display <- v09_model_summary
names(model_summary_display) <- c(
  "observation_model",
  "What the LAD application shows",
  "How to read the output"
)

v09_kable(model_summary_display)
```

## Choosing a variant in practice

Start with `coverage_offset` when you have coverage rates and want
adjusted flows on a true-flow scale. Use `model_engine = "frequentist"`
first when you want to test formulas and covariates, then fit
`model_engine = "bayesian"` when you want posterior summaries and
uncertainty intervals.

Use `latent_two_level` when the data structure contains real repeated
source/time observations of the same OD or OD-time flow. This model can
be more informative than `coverage_offset`, but only when the repeated
rows provide information about a shared latent state.

Use `reduced_form` when the goal is an MPD-scale counterfactual or a
compatibility check with earlier reduced-form analyses. Do not interpret
its `flow_adj` column as the same true-flow quantity returned by
`coverage_offset` or `latent_two_level`.
