---
title: "Bayesian Empirical Flow S1-S4 Validation"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: true
  message: true
---

## Purpose

This notebook is the manual RStudio validation runner for the empirical
source/time structures now available under `/Volumes/DEBIAS/data/outputs/flows`.
It uses the HTW flow outputs first because that branch contains mobile-phone
sources (`mapp1`, `mapp2`) and Census travel-to-work benchmarks at LAD/LTLA and
MSOA support.

The notebook tests whether `adjust_multilevel_bayes()` can run on real
empirical S1-S4 inputs, records runtime and diagnostics, and compares the
Bayesian outputs with deterministic baselines using the exported `debiasR`
validation and visualisation functions. The default run is intentionally small.
It runs LTLA frequentist smoke tests and leaves Bayesian sampling off until the
schema audit succeeds.

The empirical scenario constructions are:

- S1: `mapp1` month, one source and one aggregation window.
- S2: `mapp1` week plus month, one source and multiple aggregation windows.
- S3: `mapp1` month plus `mapp2` month, multiple sources and one aggregation
  window.
- S4: all available `mapp1` and `mapp2` week/month files for the selected
  geography, multiple sources and multiple aggregation windows. This is
  currently unbalanced because no `mapp2` week file is present in the inspected
  HTW folder.

Treat `mpd_time` here as the empirical time or aggregation-window identifier
available in the files. If a later metadata file supplies actual dates or
period IDs, replace the filename-derived `mpd_time` values before fitting S2
and S4.

## Manual Run Protocol

1. Run the notebook once with the defaults. Confirm that the file manifest,
   scenario audit, deterministic baselines, and frequentist smoke fits complete.
2. If the frequentist smoke tests pass, set `run_bayesian_pilots <- TRUE` and
   keep `area_grid <- c(6, 8)` or another small grid.
3. If the coverage-offset Bayesian pilots pass for S3 and S4, set
   `run_latent_bayesian_pilots <- TRUE` to test the experimental
   `latent_two_level` backend.
4. Increase `area_grid`, `iter_exploratory`, and `chains_exploratory` only after
   runtime and diagnostics are acceptable on the smaller grids.
5. Do not commit raw flow files, local caches, or bulky rendered outputs from
   this external volume.

## Controls

```{r}
flow_root <- "/Volumes/DEBIAS/data/outputs/flows"

install_from_github <- FALSE
use_local_source <- FALSE

geography <- "lad" # "lad" or "msoa"
include_self_flows <- TRUE
prediction_scope_default <- "observed" # use "complete_grid" only after smoke tests pass

scenario_ids <- c("s1", "s2", "s3", "s4")
area_grid <- c(6, 8)
plot_scenario <- "s4"
plot_n_areas <- max(area_grid)

primary_source <- "mapp1"
primary_time <- "month"

run_deterministic_baselines <- TRUE
run_frequentist_smoke <- TRUE
run_bayesian_pilots <- FALSE
run_latent_bayesian_pilots <- FALSE
run_random_intercept_specs <- TRUE
run_random_slope_specs <- TRUE
run_complete_grid_sensitivity <- FALSE

iter_exploratory <- 200
chains_exploratory <- 2
iter_confirmatory <- 1000
chains_confirmatory <- 2
use_confirmatory_settings <- FALSE

flow_adj_summary <- "median"
seed_base <- 25062026
max_model_runtime_minutes <- Inf
validation_drop_zeros <- TRUE

selection_r_grid <- seq(0, 3, by = 0.25)
selection_k_grid <- seq(0.1, 5, by = 0.25)
coverage_user_floor <- 1
coverage_cap <- 0.98
```

## Setup

```{r}
if (isTRUE(install_from_github)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    stop("Install `remotes` before setting `install_from_github <- TRUE`.")
  }
  remotes::install_github(
    "de-bias/debiasR",
    dependencies = c("Depends", "Imports", "LinkingTo"),
    upgrade = "never",
    build_vignettes = FALSE
  )
}

if (isTRUE(use_local_source)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install `devtools` or set `use_local_source <- FALSE`.")
  }
  devtools::load_all(".", quiet = TRUE)
} else {
  library(debiasR)
}

required_packages <- c("dplyr", "tibble", "tidyr", "purrr", "ggplot2")
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0L) {
  stop("Install required package(s): ", paste(missing_packages, collapse = ", "))
}

library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)

if (!dir.exists(flow_root)) {
  stop("The external flow root is not mounted or does not exist: ", flow_root)
}

if (!geography %in% c("lad", "msoa")) {
  stop("`geography` must be either 'lad' or 'msoa'.")
}
```

```{r}
package_version_or_na <- function(pkg) {
  if (requireNamespace(pkg, quietly = TRUE)) {
    as.character(utils::packageVersion(pkg))
  } else {
    NA_character_
  }
}

tibble(
  item = c(
    "R",
    "debiasR",
    "rstanarm",
    "rstan",
    "brms",
    "platform",
    "flow_root",
    "geography"
  ),
  value = c(
    as.character(getRversion()),
    package_version_or_na("debiasR"),
    package_version_or_na("rstanarm"),
    package_version_or_na("rstan"),
    package_version_or_na("brms"),
    R.version$platform,
    flow_root,
    geography
  )
)
```

## Helper Functions

```{r}
`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0L) y else x
}

compact_formula <- function(x) {
  if (is.null(x)) return(NA_character_)
  paste(deparse(x), collapse = " ")
}

capture_run <- function(expr) {
  warnings <- character()
  messages <- character()
  start <- Sys.time()

  value <- tryCatch(
    withCallingHandlers(
      force(expr),
      warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w))
        invokeRestart("muffleWarning")
      },
      message = function(m) {
        messages <<- c(messages, conditionMessage(m))
        invokeRestart("muffleMessage")
      }
    ),
    error = function(e) {
      structure(list(message = conditionMessage(e)), class = "captured_error")
    }
  )

  elapsed_seconds <- as.numeric(difftime(Sys.time(), start, units = "secs"))
  success <- !inherits(value, "captured_error")

  list(
    success = success,
    result = if (success) value else NULL,
    error = if (success) NA_character_ else value$message,
    warnings = warnings,
    messages = messages,
    elapsed_seconds = elapsed_seconds
  )
}

read_flow_csv <- function(path, n_max = Inf) {
  nrows <- if (is.finite(n_max)) as.integer(n_max) else -1L
  out <- utils::read.csv(path, nrows = nrows, check.names = FALSE)
  blank_names <- !nzchar(names(out))
  if (any(blank_names)) {
    names(out)[blank_names] <- paste0("unnamed_", which(blank_names))
  }
  tibble::as_tibble(out)
}

count_csv_rows <- function(path) {
  nrow(read_flow_csv(path))
}

first_existing_col <- function(df, candidates, label) {
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0L) {
    stop(
      "Could not find ", label, " column. Candidate names were: ",
      paste(candidates, collapse = ", "),
      ". Available names are: ",
      paste(names(df), collapse = ", ")
    )
  }
  hit[[1]]
}

valid_area_code <- function(x) {
  !is.na(x) & grepl("^E[0-9]+$", as.character(x))
}

flow_file_manifest <- function(root) {
  htw_root <- file.path(root, "htw")
  files <- list.files(htw_root, pattern = "csv$", recursive = TRUE, full.names = TRUE)
  files <- files[!grepl("/census/", files)]

  tibble(path = files) |>
    mutate(
      file = basename(.data$path),
      source_folder = basename(dirname(.data$path)),
      mpd_source = case_when(
        grepl("mapp1", .data$file, ignore.case = TRUE) ~ "mapp1",
        grepl("mapp2", .data$file, ignore.case = TRUE) ~ "mapp2",
        TRUE ~ NA_character_
      ),
      geography = case_when(
        grepl("_lad_", .data$file, ignore.case = TRUE) ~ "lad",
        grepl("_msoa_", .data$file, ignore.case = TRUE) ~ "msoa",
        grepl("_lsoa_", .data$file, ignore.case = TRUE) ~ "lsoa",
        TRUE ~ NA_character_
      ),
      mpd_time = case_when(
        grepl("_week", .data$file, ignore.case = TRUE) ~ "week",
        grepl("_month", .data$file, ignore.case = TRUE) ~ "month",
        TRUE ~ NA_character_
      )
    ) |>
    filter(!is.na(.data$mpd_source), !is.na(.data$geography), !is.na(.data$mpd_time)) |>
    arrange(.data$geography, .data$mpd_source, .data$mpd_time, .data$file)
}

census_path_for_geography <- function(root, geography) {
  suffix <- switch(
    geography,
    lad = "LTLA",
    msoa = "MSOA",
    stop("Unsupported geography: ", geography)
  )
  path <- file.path(root, "htw", "census", paste0("ODWP01EW_", suffix, ".csv"))
  if (!file.exists(path)) {
    stop("Missing Census benchmark file: ", path)
  }
  path
}

normalise_mpd_htw <- function(path, mpd_source, mpd_time, geography, include_self_flows = TRUE) {
  raw <- read_flow_csv(path)

  if (geography == "lad") {
    origin_col <- first_existing_col(raw, c("LAD22CD_home", "home_LAD22CD", "origin", "home"), "MPD origin")
    destination_col <- first_existing_col(raw, c("LAD22CD_work", "work_LAD22CD", "destination", "work"), "MPD destination")
  } else if (geography == "msoa") {
    origin_col <- first_existing_col(raw, c("MSOA21CD_home", "home_MSOA21CD", "origin", "home"), "MPD origin")
    destination_col <- first_existing_col(raw, c("MSOA21CD_work", "work_MSOA21CD", "destination", "work"), "MPD destination")
  } else {
    stop("Unsupported geography: ", geography)
  }

  flow_col <- first_existing_col(raw, c("flow", "count", "trips", "total_flow"), "MPD flow")

  out <- raw |>
    transmute(
      origin = as.character(.data[[origin_col]]),
      destination = as.character(.data[[destination_col]]),
      flow = suppressWarnings(as.numeric(.data[[flow_col]])),
      mpd_source = mpd_source,
      mpd_time = mpd_time,
      geography = geography,
      source_file = basename(path)
    ) |>
    filter(
      valid_area_code(.data$origin),
      valid_area_code(.data$destination),
      is.finite(.data$flow),
      .data$flow >= 0
    )

  if (!isTRUE(include_self_flows)) {
    out <- out |>
      filter(.data$origin != .data$destination)
  }

  out |>
    group_by(.data$origin, .data$destination, .data$mpd_source, .data$mpd_time, .data$geography, .data$source_file) |>
    summarise(flow = sum(.data$flow, na.rm = TRUE), .groups = "drop") |>
    arrange(.data$mpd_source, .data$mpd_time, .data$origin, .data$destination)
}

normalise_census_htw <- function(path, geography, include_self_flows = TRUE) {
  raw <- read_flow_csv(path)

  if (geography == "lad") {
    origin_col <- first_existing_col(raw, c("Lower tier local authorities code", "LTLA code", "origin"), "Census origin")
    destination_col <- first_existing_col(raw, c("LTLA of workplace code", "workplace LTLA code", "destination"), "Census destination")
  } else if (geography == "msoa") {
    origin_col <- first_existing_col(raw, c("Middle layer Super Output Areas code", "MSOA code", "origin"), "Census origin")
    destination_col <- first_existing_col(raw, c("MSOA of workplace code", "workplace MSOA code", "destination"), "Census destination")
  } else {
    stop("Unsupported geography: ", geography)
  }

  flow_col <- first_existing_col(raw, c("flow", "Count", "count"), "Census flow")

  out <- raw |>
    transmute(
      origin = as.character(.data[[origin_col]]),
      destination = as.character(.data[[destination_col]]),
      flow = suppressWarnings(as.numeric(.data[[flow_col]]))
    ) |>
    filter(
      valid_area_code(.data$origin),
      valid_area_code(.data$destination),
      is.finite(.data$flow),
      .data$flow >= 0
    )

  if (!isTRUE(include_self_flows)) {
    out <- out |>
      filter(.data$origin != .data$destination)
  }

  out |>
    group_by(.data$origin, .data$destination) |>
    summarise(flow = sum(.data$flow, na.rm = TRUE), .groups = "drop") |>
    arrange(.data$origin, .data$destination)
}

manifest_for_scenario <- function(manifest, scenario, geography, primary_source = "mapp1", primary_time = "month") {
  manifest <- manifest |>
    filter(.data$geography == !!geography)

  out <- switch(
    scenario,
    s1 = manifest |>
      filter(.data$mpd_source == !!primary_source, .data$mpd_time == !!primary_time),
    s2 = manifest |>
      filter(.data$mpd_source == !!primary_source),
    s3 = manifest |>
      filter(.data$mpd_time == !!primary_time, .data$mpd_source %in% c("mapp1", "mapp2")),
    s4 = manifest |>
      filter(.data$mpd_source %in% c("mapp1", "mapp2")),
    stop("Unknown scenario: ", scenario)
  )

  out |>
    arrange(.data$mpd_source, .data$mpd_time, .data$file)
}

scenario_available <- function(manifest, scenario) {
  n_sources <- dplyr::n_distinct(manifest$mpd_source)
  n_times <- dplyr::n_distinct(manifest$mpd_time)
  switch(
    scenario,
    s1 = nrow(manifest) >= 1L && n_sources == 1L && n_times == 1L,
    s2 = nrow(manifest) >= 2L && n_sources == 1L && n_times >= 2L,
    s3 = nrow(manifest) >= 2L && n_sources >= 2L && n_times == 1L,
    s4 = nrow(manifest) >= 2L && n_sources >= 2L && n_times >= 2L,
    FALSE
  )
}

complete_benchmark_grid <- function(benchmark, areas, include_self_flows = TRUE) {
  grid <- tidyr::expand_grid(
    origin = areas,
    destination = areas
  )
  if (!isTRUE(include_self_flows)) {
    grid <- filter(grid, .data$origin != .data$destination)
  }

  grid |>
    left_join(benchmark, by = c("origin", "destination")) |>
    mutate(
      benchmark_observed = is.finite(.data$flow),
      benchmark_zero_filled = !.data$benchmark_observed,
      flow = coalesce(.data$flow, 0)
    ) |>
    arrange(.data$origin, .data$destination)
}

complete_mpd_grid <- function(mpd, areas, include_self_flows = TRUE, fill_missing_flow = 0) {
  scenario_grid <- mpd |>
    distinct(.data$mpd_source, .data$mpd_time, .data$geography, .data$source_file)

  grid <- tidyr::expand_grid(
    scenario_grid,
    origin = areas,
    destination = areas
  )
  if (!isTRUE(include_self_flows)) {
    grid <- filter(grid, .data$origin != .data$destination)
  }

  grid |>
    left_join(
      mpd,
      by = c("origin", "destination", "mpd_source", "mpd_time", "geography", "source_file")
    ) |>
    mutate(
      mpd_observed = is.finite(.data$flow),
      mpd_zero_filled = !.data$mpd_observed,
      mpd_row_status = if_else(.data$mpd_observed, "observed", "zero_filled"),
      flow = if_else(.data$mpd_observed, .data$flow, fill_missing_flow)
    ) |>
    arrange(.data$mpd_source, .data$mpd_time, .data$origin, .data$destination)
}

select_area_set <- function(mpd, benchmark, n_areas) {
  common_areas <- intersect(
    union(mpd$origin, mpd$destination),
    union(benchmark$origin, benchmark$destination)
  )
  if (length(common_areas) < 2L) {
    stop("Fewer than two overlapping areas were found between MPD and Census.")
  }

  scores <- bind_rows(
    mpd |> transmute(area = .data$origin, score = .data$flow),
    mpd |> transmute(area = .data$destination, score = .data$flow),
    benchmark |> transmute(area = .data$origin, score = .data$flow),
    benchmark |> transmute(area = .data$destination, score = .data$flow)
  ) |>
    filter(.data$area %in% common_areas) |>
    group_by(.data$area) |>
    summarise(score = sum(.data$score, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(.data$score), .data$area)

  if (!is.null(n_areas) && is.finite(n_areas)) {
    scores <- slice_head(scores, n = as.integer(n_areas))
  }

  sort(scores$area)
}

build_coverage_table <- function(mpd, benchmark_complete, areas, coverage_user_floor = 1, coverage_cap = 0.98) {
  population <- benchmark_complete |>
    group_by(.data$origin) |>
    summarise(population = sum(.data$flow, na.rm = TRUE), .groups = "drop") |>
    filter(.data$origin %in% areas)

  scenario_grid <- mpd |>
    distinct(.data$mpd_source, .data$mpd_time)

  user_counts <- mpd |>
    group_by(.data$origin, .data$mpd_source, .data$mpd_time) |>
    summarise(user_count_raw = sum(.data$flow, na.rm = TRUE), .groups = "drop")

  tidyr::expand_grid(
    origin = areas,
    scenario_grid
  ) |>
    left_join(population, by = "origin") |>
    left_join(user_counts, by = c("origin", "mpd_source", "mpd_time")) |>
    mutate(
      population = coalesce(.data$population, 0),
      user_count_raw = coalesce(.data$user_count_raw, 0),
      population = pmax(.data$population, coverage_user_floor + 1),
      user_count = pmax(.data$user_count_raw, coverage_user_floor),
      user_count = pmin(.data$user_count, .data$population * coverage_cap),
      coverage_score = .data$user_count / .data$population,
      coverage_bias = 1 - .data$coverage_score,
      coverage_was_floored = .data$user_count_raw < coverage_user_floor
    ) |>
    arrange(.data$mpd_source, .data$mpd_time, .data$origin)
}

build_covariates_table <- function(coverage) {
  coverage |>
    group_by(.data$origin) |>
    summarise(
      population = max(.data$population, na.rm = TRUE),
      user_count_mean = mean(.data$user_count, na.rm = TRUE),
      coverage_score_mean = mean(.data$coverage_score, na.rm = TRUE),
      coverage_score_min = min(.data$coverage_score, na.rm = TRUE),
      coverage_bias_mean = mean(.data$coverage_bias, na.rm = TRUE),
      .groups = "drop"
    ) |>
    transmute(
      area = .data$origin,
      population = .data$population,
      log_population = log(pmax(.data$population, .Machine$double.eps)),
      user_count_mean = .data$user_count_mean,
      coverage_score_mean = .data$coverage_score_mean,
      coverage_score_min = .data$coverage_score_min,
      coverage_bias_mean = .data$coverage_bias_mean
    )
}

build_scenario_example <- function(manifest,
                                   census_path,
                                   scenario,
                                   geography,
                                   n_areas,
                                   prediction_scope = "observed",
                                   include_self_flows = TRUE) {
  scenario_manifest <- manifest_for_scenario(
    manifest = manifest,
    scenario = scenario,
    geography = geography,
    primary_source = primary_source,
    primary_time = primary_time
  )

  if (!scenario_available(scenario_manifest, scenario)) {
    return(list(
      available = FALSE,
      scenario = scenario,
      n_areas = n_areas,
      prediction_scope = prediction_scope,
      error = paste0("Insufficient files to construct ", toupper(scenario), "."),
      manifest = scenario_manifest
    ))
  }

  mpd_all <- purrr::pmap_dfr(
    scenario_manifest,
    function(path, file, source_folder, mpd_source, geography, mpd_time, ...) {
      normalise_mpd_htw(
        path = path,
        mpd_source = mpd_source,
        mpd_time = mpd_time,
        geography = geography,
        include_self_flows = include_self_flows
      )
    }
  )

  benchmark <- normalise_census_htw(
    path = census_path,
    geography = geography,
    include_self_flows = include_self_flows
  )

  areas <- select_area_set(mpd_all, benchmark, n_areas = n_areas)

  mpd_selected <- mpd_all |>
    filter(.data$origin %in% areas, .data$destination %in% areas) |>
    group_by(.data$origin, .data$destination, .data$mpd_source, .data$mpd_time, .data$geography, .data$source_file) |>
    summarise(flow = sum(.data$flow, na.rm = TRUE), .groups = "drop")

  benchmark_complete <- benchmark |>
    filter(.data$origin %in% areas, .data$destination %in% areas) |>
    complete_benchmark_grid(areas = areas, include_self_flows = include_self_flows)

  mpd_model <- if (identical(prediction_scope, "complete_grid")) {
    complete_mpd_grid(
      mpd = mpd_selected,
      areas = areas,
      include_self_flows = include_self_flows
    )
  } else {
    mpd_selected |>
      mutate(
        mpd_observed = TRUE,
        mpd_zero_filled = FALSE,
        mpd_row_status = "observed"
      )
  }

  coverage <- build_coverage_table(
    mpd = mpd_selected,
    benchmark_complete = benchmark_complete,
    areas = areas,
    coverage_user_floor = coverage_user_floor,
    coverage_cap = coverage_cap
  )
  covariates <- build_covariates_table(coverage)

  list(
    available = TRUE,
    scenario = scenario,
    n_areas = length(areas),
    prediction_scope = prediction_scope,
    manifest = scenario_manifest,
    areas = areas,
    mpd_od = mpd_model,
    mpd_observed_only = mpd_selected,
    benchmark_od = benchmark_complete,
    coverage = coverage,
    covariates = covariates,
    distance = NULL
  )
}

scenario_repeated_observation <- function(scenario) {
  switch(
    scenario,
    s1 = "none",
    s2 = "time",
    s3 = "source",
    s4 = "source_time",
    "auto"
  )
}

scenario_random_intercepts <- function(scenario) {
  base <- c("none", "origin", "destination", "od")
  if (scenario %in% c("s3", "s4")) base <- c(base, "source")
  if (scenario %in% c("s2", "s4")) base <- c(base, "time")
  if (scenario == "s4") base <- c(base, "source_time")
  base
}
```

## File Manifest

```{r}
manifest <- flow_file_manifest(flow_root)
census_path <- census_path_for_geography(flow_root, geography)

manifest |>
  mutate(
    selected_geography = .data$geography == !!geography,
    file_exists = file.exists(.data$path)
  )
```

```{r}
manifest_audit <- manifest |>
  filter(.data$geography == !!geography) |>
  mutate(
    n_rows = purrr::map_int(.data$path, count_csv_rows)
  ) |>
  select(geography, mpd_source, mpd_time, file, n_rows, path)

manifest_audit
```

## Scenario Construction Audit

```{r}
prediction_scopes <- if (isTRUE(run_complete_grid_sensitivity)) {
  c(prediction_scope_default, "complete_grid") |> unique()
} else {
  prediction_scope_default
}

scenario_examples <- list()

for (scenario in scenario_ids) {
  for (n_area in area_grid) {
    for (scope in prediction_scopes) {
      key <- paste(scenario, paste0("n", n_area), scope, sep = "__")
      scenario_examples[[key]] <- build_scenario_example(
        manifest = manifest,
        census_path = census_path,
        scenario = scenario,
        geography = geography,
        n_areas = n_area,
        prediction_scope = scope,
        include_self_flows = include_self_flows
      )
    }
  }
}

scenario_audit <- bind_rows(lapply(names(scenario_examples), function(key) {
  ex <- scenario_examples[[key]]
  if (!isTRUE(ex$available)) {
    return(tibble(
      run_key = key,
      scenario = ex$scenario,
      n_areas = ex$n_areas,
      prediction_scope = ex$prediction_scope,
      available = FALSE,
      error = ex$error,
      n_mpd_rows = NA_integer_,
      n_benchmark_rows = NA_integer_,
      n_coverage_rows = NA_integer_,
      n_sources = dplyr::n_distinct(ex$manifest$mpd_source),
      n_time_periods = dplyr::n_distinct(ex$manifest$mpd_time),
      source_levels = paste(sort(unique(ex$manifest$mpd_source)), collapse = ", "),
      time_levels = paste(sort(unique(ex$manifest$mpd_time)), collapse = ", "),
      source_time_cells = NA_integer_,
      n_zero_filled = NA_integer_
    ))
  }

  tibble(
    run_key = key,
    scenario = ex$scenario,
    n_areas = ex$n_areas,
    prediction_scope = ex$prediction_scope,
    available = TRUE,
    error = NA_character_,
    n_mpd_rows = nrow(ex$mpd_od),
    n_benchmark_rows = nrow(ex$benchmark_od),
    n_coverage_rows = nrow(ex$coverage),
    n_sources = dplyr::n_distinct(ex$mpd_od$mpd_source),
    n_time_periods = dplyr::n_distinct(ex$mpd_od$mpd_time),
    source_levels = paste(sort(unique(ex$mpd_od$mpd_source)), collapse = ", "),
    time_levels = paste(sort(unique(ex$mpd_od$mpd_time)), collapse = ", "),
    source_time_cells = nrow(distinct(ex$mpd_od, .data$mpd_source, .data$mpd_time)),
    n_zero_filled = sum(ex$mpd_od$mpd_zero_filled %in% TRUE, na.rm = TRUE)
  )
}))

scenario_audit
```

```{r}
coverage_audit <- bind_rows(lapply(names(scenario_examples), function(key) {
  ex <- scenario_examples[[key]]
  if (!isTRUE(ex$available)) return(tibble())

  ex$coverage |>
    summarise(
      run_key = key,
      scenario = ex$scenario,
      n_areas = ex$n_areas,
      prediction_scope = ex$prediction_scope,
      min_coverage = min(.data$coverage_score, na.rm = TRUE),
      median_coverage = stats::median(.data$coverage_score, na.rm = TRUE),
      max_coverage = max(.data$coverage_score, na.rm = TRUE),
      floored_rows = sum(.data$coverage_was_floored, na.rm = TRUE),
      .groups = "drop"
    )
}))

coverage_audit
```

## Model Specification Registry

The coverage-offset models estimate Census-scale true flows while treating
source/time-specific coverage as a fixed observation offset. The reduced-form
models estimate a counterfactual MPD-to-benchmark correction from the fitted
bias covariate rather than using the coverage offset. The latent two-level
models are the novel experimental path. They estimate source-invariant latent
true OD or OD-time flow states and model observed source/time rows as noisy
coverage-scaled observations of those latent states.

```{r}
make_adjustment_specs <- function(scenario) {
  specs <- list()

  add_spec <- function(spec_id,
                       model_engine,
                       observation_model,
                       target_scale,
                       coverage_scale = "origin",
                       model_family = "poisson",
                       random_intercept = "none",
                       mobility_formula = ~ log_pop_o + log_pop_d + log_distance,
                       bias_formula = ~ bias_e_origin,
                       backend = "auto",
                       latent_flow_unit = "auto",
                       run = TRUE,
                       max_n_areas = Inf) {
    specs[[length(specs) + 1L]] <<- list(
      spec_id = spec_id,
      model_engine = model_engine,
      observation_model = observation_model,
      target_scale = target_scale,
      coverage_scale = coverage_scale,
      model_family = model_family,
      random_intercept = random_intercept,
      mobility_formula = mobility_formula,
      bias_formula = bias_formula,
      backend = backend,
      latent_flow_unit = latent_flow_unit,
      run = run,
      max_n_areas = max_n_areas
    )
  }

  if (isTRUE(run_frequentist_smoke)) {
    add_spec(
      spec_id = "frequentist__coverage_offset__fixed",
      model_engine = "frequentist",
      observation_model = "coverage_offset",
      target_scale = "true_flow"
    )

    add_spec(
      spec_id = "frequentist__reduced_form__fixed",
      model_engine = "frequentist",
      observation_model = "reduced_form",
      target_scale = "mpd_counterfactual",
      coverage_scale = NA_character_
    )

    if (isTRUE(run_random_intercept_specs)) {
      for (ri in setdiff(scenario_random_intercepts(scenario), c("none", "od"))) {
        add_spec(
          spec_id = paste("frequentist__coverage_offset__ri", ri, sep = "__"),
          model_engine = "frequentist",
          observation_model = "coverage_offset",
          target_scale = "true_flow",
          random_intercept = ri
        )
      }
    }

    if (isTRUE(run_random_slope_specs)) {
      add_spec(
        spec_id = "frequentist__coverage_offset__rs_log_distance_origin",
        model_engine = "frequentist",
        observation_model = "coverage_offset",
        target_scale = "true_flow",
        random_intercept = "none",
        mobility_formula = ~ log_pop_o + log_pop_d + log_distance + (1 + log_distance | origin),
        max_n_areas = max(area_grid)
      )
    }
  }

  if (isTRUE(run_bayesian_pilots)) {
    add_spec(
      spec_id = "bayesian__coverage_offset__fixed",
      model_engine = "bayesian",
      observation_model = "coverage_offset",
      target_scale = "true_flow",
      backend = "auto"
    )

    if (isTRUE(run_random_intercept_specs)) {
      add_spec(
        spec_id = "bayesian__coverage_offset__ri_origin",
        model_engine = "bayesian",
        observation_model = "coverage_offset",
        target_scale = "true_flow",
        random_intercept = "origin",
        backend = "auto",
        max_n_areas = min(max(area_grid), 8)
      )
    }
  }

  if (isTRUE(run_latent_bayesian_pilots) && scenario %in% c("s3", "s4")) {
    add_spec(
      spec_id = "bayesian__latent_two_level__fixed",
      model_engine = "bayesian",
      observation_model = "latent_two_level",
      target_scale = "true_flow",
      backend = "stan_latent",
      latent_flow_unit = if (scenario == "s4") "od_time" else "od",
      max_n_areas = min(max(area_grid), 8)
    )
  }

  specs
}

spec_registry <- bind_rows(lapply(scenario_ids, function(scenario) {
  bind_rows(lapply(make_adjustment_specs(scenario), function(spec) {
    tibble(
      scenario = scenario,
      spec_id = spec$spec_id,
      model_engine = spec$model_engine,
      observation_model = spec$observation_model,
      target_scale = spec$target_scale,
      coverage_scale = spec$coverage_scale,
      model_family = spec$model_family,
      random_intercept = spec$random_intercept,
      backend = spec$backend,
      latent_flow_unit = spec$latent_flow_unit,
      mobility_formula = compact_formula(spec$mobility_formula),
      bias_formula = compact_formula(spec$bias_formula),
      max_n_areas = spec$max_n_areas,
      run = spec$run
    )
  }))
}))

spec_registry
```

## Deterministic Baselines

The deterministic baselines are run within each source-time slice and then
bound back together. This keeps `adjust_selection_rate()` and
`adjust_selection_rate2()` in the comparison while avoiding unsupported
time-aware joins inside methods that were originally source-level.

```{r}
baseline_method_label <- c(
  raw_mpd = "Unadjusted raw MPD",
  inverse_penetration_both = "Inverse penetration",
  selection_rate = "Selection rate",
  selection_rate2 = "Selection rate II",
  raking_ratio = "Raking ratio",
  coefficient_ols = "Coefficient OLS"
)

as_adjusted_slice <- function(result, source_value, time_value) {
  result |>
    mutate(
      mpd_source = source_value,
      mpd_time = time_value
    ) |>
    select(origin, destination, mpd_source, mpd_time, flow, flow_adj, everything())
}

run_baselines_for_slice <- function(mpd_slice, coverage_slice, covariates, benchmark) {
  source_value <- unique(mpd_slice$mpd_source)[[1]]
  time_value <- unique(mpd_slice$mpd_time)[[1]]
  coverage_slice_source <- coverage_slice |>
    select(origin, mpd_source, population, user_count)

  baseline_calls <- list(
    raw_mpd = function() {
      mpd_slice |>
        mutate(flow_adj = .data$flow)
    },
    inverse_penetration_both = function() {
      debiasR::adjust_inverse_penetration(
        mpd_od_df = mpd_slice,
        coverage_df = coverage_slice_source,
        weight_by = "both"
      )
    },
    selection_rate = function() {
      debiasR::adjust_selection_rate(
        mpd_od_df = mpd_slice,
        coverage_df = coverage_slice_source,
        covariates_df = covariates,
        covariate_col = "coverage_score_mean",
        weight_by = "origin",
        benchmark_od_df = benchmark,
        calibration_aggregate = "origin",
        r_grid = selection_r_grid
      )
    },
    selection_rate2 = function() {
      debiasR::adjust_selection_rate2(
        mpd_od_df = mpd_slice,
        coverage_df = coverage_slice_source,
        weight_by = "origin",
        benchmark_od_df = benchmark,
        calibration_aggregate = "origin",
        k_grid = selection_k_grid
      )
    },
    raking_ratio = function() {
      debiasR::adjust_raking_ratio(
        mpd_od_df = mpd_slice,
        benchmark_od_df = benchmark
      )
    },
    coefficient_ols = function() {
      debiasR::adjust_coefficient(
        mpd_od_df = mpd_slice,
        benchmark_od_df = benchmark,
        model_family = "ols",
        level = "od"
      )
    }
  )

  lapply(names(baseline_calls), function(method_name) {
    run <- capture_run(baseline_calls[[method_name]]())
    if (isTRUE(run$success)) {
      run$result <- as_adjusted_slice(run$result, source_value, time_value)
    }
    list(method_name = method_name, run = run)
  }) |>
    stats::setNames(names(baseline_calls))
}

run_baselines_for_example <- function(example) {
  split_keys <- example$mpd_od |>
    distinct(.data$mpd_source, .data$mpd_time)

  slice_runs <- purrr::pmap(split_keys, function(mpd_source, mpd_time) {
    mpd_slice <- example$mpd_od |>
      filter(.data$mpd_source == !!mpd_source, .data$mpd_time == !!mpd_time)
    coverage_slice <- example$coverage |>
      filter(.data$mpd_source == !!mpd_source, .data$mpd_time == !!mpd_time)
    run_baselines_for_slice(
      mpd_slice = mpd_slice,
      coverage_slice = coverage_slice,
      covariates = example$covariates,
      benchmark = example$benchmark_od
    )
  })

  method_names <- names(slice_runs[[1]])
  stats::setNames(lapply(method_names, function(method_name) {
    runs <- lapply(slice_runs, `[[`, method_name)
    success <- all(vapply(runs, function(x) isTRUE(x$run$success), logical(1)))
    list(
      method_name = method_name,
      run = list(
        success = success,
        result = if (success) bind_rows(lapply(runs, function(x) x$run$result)) else NULL,
        error = if (success) NA_character_ else paste(na.omit(vapply(runs, function(x) x$run$error, character(1))), collapse = " | "),
        warnings = unlist(lapply(runs, function(x) x$run$warnings), use.names = FALSE),
        messages = unlist(lapply(runs, function(x) x$run$messages), use.names = FALSE),
        elapsed_seconds = sum(vapply(runs, function(x) x$run$elapsed_seconds, numeric(1)), na.rm = TRUE)
      )
    )
  }), method_names)
}

baseline_runs <- list()

if (isTRUE(run_deterministic_baselines)) {
  for (key in names(scenario_examples)) {
    ex <- scenario_examples[[key]]
    if (isTRUE(ex$available)) {
      baseline_runs[[key]] <- run_baselines_for_example(ex)
    }
  }
}

baseline_summary <- bind_rows(lapply(names(baseline_runs), function(key) {
  ex <- scenario_examples[[key]]
  bind_rows(lapply(baseline_runs[[key]], function(item) {
    tibble(
      run_key = key,
      scenario = ex$scenario,
      n_areas = ex$n_areas,
      prediction_scope = ex$prediction_scope,
      method = item$method_name,
      model_class = "deterministic_baseline",
      success = item$run$success,
      elapsed_seconds = item$run$elapsed_seconds,
      n_rows = if (isTRUE(item$run$success)) nrow(item$run$result) else NA_integer_,
      error = item$run$error,
      warnings = paste(unique(item$run$warnings), collapse = " | ")
    )
  }))
}))

baseline_summary
```

## Multilevel Runs

```{r}
run_multilevel_spec <- function(example, spec, spec_index) {
  if (!isTRUE(spec$run) || example$n_areas > spec$max_n_areas) {
    return(list(
      success = FALSE,
      result = NULL,
      error = "Skipped by notebook controls or max_n_areas.",
      warnings = character(),
      messages = character(),
      elapsed_seconds = 0
    ))
  }

  iter <- if (isTRUE(use_confirmatory_settings)) iter_confirmatory else iter_exploratory
  chains <- if (isTRUE(use_confirmatory_settings)) chains_confirmatory else chains_exploratory
  seed <- seed_base + spec_index + example$n_areas + match(example$scenario, scenario_ids) * 100L

  capture_run(
    debiasR::adjust_multilevel_bayes(
      mpd_od_df = example$mpd_od,
      coverage_df = example$coverage,
      covariates_df = example$covariates,
      distance_df = example$distance,
      source_col = "mpd_source",
      time_col = "mpd_time",
      scenario = example$scenario,
      repeated_observation = scenario_repeated_observation(example$scenario),
      random_intercept = spec$random_intercept,
      mobility_formula = spec$mobility_formula,
      bias_formula = spec$bias_formula,
      model_family = spec$model_family,
      model_engine = spec$model_engine,
      backend = spec$backend,
      target_scale = spec$target_scale,
      observation_model = spec$observation_model,
      coverage_scale = if (is.na(spec$coverage_scale)) "origin" else spec$coverage_scale,
      latent_flow_unit = spec$latent_flow_unit,
      prediction_scope = example$prediction_scope,
      flow_adj_summary = flow_adj_summary,
      iter = iter,
      chains = chains,
      seed = seed,
      refresh = 0,
      include_flow_adj_draws = FALSE
    )
  )
}

multilevel_runs <- list()

for (key in names(scenario_examples)) {
  ex <- scenario_examples[[key]]
  if (!isTRUE(ex$available)) next

  specs <- make_adjustment_specs(ex$scenario)
  for (spec_i in seq_along(specs)) {
    spec <- specs[[spec_i]]
    run_id <- paste(key, spec$spec_id, sep = "__")
    multilevel_runs[[run_id]] <- list(
      run_key = key,
      scenario = ex$scenario,
      n_areas = ex$n_areas,
      prediction_scope = ex$prediction_scope,
      spec = spec,
      run = run_multilevel_spec(ex, spec, spec_i)
    )

    if (is.finite(max_model_runtime_minutes)) {
      elapsed_minutes <- multilevel_runs[[run_id]]$run$elapsed_seconds / 60
      if (elapsed_minutes > max_model_runtime_minutes) {
        warning(
          "Model runtime exceeded `max_model_runtime_minutes`: ",
          run_id,
          " elapsed minutes = ",
          round(elapsed_minutes, 2)
        )
      }
    }
  }
}

summarise_multilevel_run <- function(run_id, item) {
  result <- item$run$result
  metadata <- if (!is.null(result)) attr(result, "result_metadata") else NULL
  diagnostics <- if (!is.null(result)) attr(result, "diagnostics") else NULL
  convergence <- if (!is.null(diagnostics)) diagnostics$convergence else NULL

  tibble(
    run_id = run_id,
    run_key = item$run_key,
    scenario = item$scenario,
    n_areas = item$n_areas,
    prediction_scope = item$prediction_scope,
    method = item$spec$spec_id,
    model_class = item$spec$model_engine,
    observation_model = item$spec$observation_model,
    random_intercept = item$spec$random_intercept,
    backend = item$spec$backend,
    latent_flow_unit = item$spec$latent_flow_unit,
    success = item$run$success,
    elapsed_seconds = item$run$elapsed_seconds,
    n_rows = if (!is.null(result)) nrow(result) else NA_integer_,
    n_fit_rows = metadata$n_fit_rows %||% NA_integer_,
    n_prediction_rows = metadata$n_prediction_rows %||% NA_integer_,
    n_sources = metadata$n_sources %||% NA_integer_,
    n_time_periods = metadata$n_time_periods %||% NA_integer_,
    resolved_scenario = metadata$scenario %||% NA_character_,
    rhat_max = convergence$rhat_max %||% diagnostics$rhat_max %||% NA_real_,
    ess_bulk_min = convergence$ess_bulk_min %||% diagnostics$ess_bulk_min %||% NA_real_,
    ess_tail_min = convergence$ess_tail_min %||% diagnostics$ess_tail_min %||% NA_real_,
    n_eff_min = convergence$n_eff_min %||% diagnostics$n_eff_min %||% NA_real_,
    error = item$run$error,
    warnings = paste(unique(item$run$warnings), collapse = " | ")
  )
}

multilevel_summary <- bind_rows(lapply(names(multilevel_runs), function(run_id) {
  summarise_multilevel_run(run_id, multilevel_runs[[run_id]])
}))

multilevel_summary
```

## Validation Against Census Benchmark

```{r}
collect_adjusted_results <- function() {
  baseline_results <- bind_rows(lapply(names(baseline_runs), function(key) {
    bind_rows(lapply(baseline_runs[[key]], function(item) {
      if (!isTRUE(item$run$success)) return(tibble())
      item$run$result |>
        mutate(
          run_key = key,
          method = item$method_name,
          model_class = "deterministic_baseline",
          scenario = scenario_examples[[key]]$scenario,
          n_areas = scenario_examples[[key]]$n_areas,
          prediction_scope = scenario_examples[[key]]$prediction_scope
        )
    }))
  }))

  multilevel_results <- bind_rows(lapply(multilevel_runs, function(item) {
    if (!isTRUE(item$run$success)) return(tibble())
    item$run$result |>
      mutate(
        run_key = item$run_key,
        method = item$spec$spec_id,
        model_class = item$spec$model_engine,
        scenario = item$scenario,
        n_areas = item$n_areas,
        prediction_scope = item$prediction_scope
      )
  }))

  bind_rows(baseline_results, multilevel_results)
}

adjusted_results <- collect_adjusted_results()

adjusted_results |>
  count(.data$scenario, .data$n_areas, .data$prediction_scope, .data$method, .data$model_class, name = "rows") |>
  arrange(.data$scenario, .data$n_areas, .data$prediction_scope, .data$model_class, .data$method)
```

```{r}
validate_one_method <- function(adj, benchmark, method_name) {
  overall <- capture_run(
    debiasR::validate_flow_overall(
      adj_df = adj,
      benchmark_od_df = benchmark,
      method_name = method_name,
      comparisons = "adjusted_vs_benchmark",
      drop_zeros = validation_drop_zeros
    )
  )

  residuals <- capture_run(
    debiasR::validate_flow_residuals(
      adj_df = adj,
      benchmark_od_df = benchmark,
      method_name = method_name,
      top_n = 10
    )
  )

  distribution <- capture_run(
    debiasR::validate_flow_distribution(
      adj_df = adj,
      benchmark_od_df = benchmark,
      method_name = method_name,
      comparisons = "adjusted_vs_benchmark",
      weight_by = "benchmark_origin_total"
    )
  )

  structure <- capture_run(
    debiasR::validate_flow_residual_structure(
      adj_df = adj,
      benchmark_od_df = benchmark,
      method_name = method_name,
      covariate_df = NULL,
      covariate_col = NULL
    )
  )

  list(
    overall = overall,
    residuals = residuals,
    distribution = distribution,
    structure = structure
  )
}

overall_summary_tbl <- function(result) {
  if (!is.null(result$summary) && inherits(result$summary, "data.frame")) {
    return(tibble::as_tibble(result$summary))
  }

  scalar_fields <- vapply(
    result,
    function(x) length(x) == 1L && !inherits(x, "data.frame"),
    logical(1)
  )
  tibble::as_tibble(result[scalar_fields])
}

residual_data_tbl <- function(result) {
  data <- result$residuals %||% result$data
  if (is.null(data)) return(tibble())
  tibble::as_tibble(data)
}

validation_runs <- list()
validation_metadata <- list()

for (key in names(scenario_examples)) {
  ex <- scenario_examples[[key]]
  if (!isTRUE(ex$available)) next

  method_data <- adjusted_results |>
    filter(.data$run_key == !!key) |>
    group_split(.data$method)

  for (method_tbl in method_data) {
    method_name <- unique(method_tbl$method)[[1]]
    validation_key <- paste(key, method_name, sep = "__")
    validation_runs[[validation_key]] <- validate_one_method(
      adj = method_tbl,
      benchmark = ex$benchmark_od,
      method_name = method_name
    )
    validation_metadata[[validation_key]] <- tibble(
      validation_key = validation_key,
      run_key = key,
      method = method_name
    )
  }
}

validation_metrics <- bind_rows(lapply(names(validation_runs), function(key) {
  run <- validation_runs[[key]]$overall
  meta <- validation_metadata[[key]]
  if (!isTRUE(run$success)) {
    return(meta |> mutate(validation_error = run$error))
  }
  overall_summary_tbl(run$result) |>
    mutate(
      validation_key = key,
      run_key = meta$run_key,
      .before = 1
    )
}))

validation_residuals <- bind_rows(lapply(names(validation_runs), function(key) {
  run <- validation_runs[[key]]$residuals
  meta <- validation_metadata[[key]]
  if (!isTRUE(run$success)) return(tibble())
  residual_data_tbl(run$result) |>
    mutate(
      validation_key = key,
      run_key = meta$run_key,
      .before = 1
    )
}))

validation_distribution <- bind_rows(lapply(names(validation_runs), function(key) {
  run <- validation_runs[[key]]$distribution
  meta <- validation_metadata[[key]]
  if (!isTRUE(run$success)) return(tibble())
  run$result$summary |>
    mutate(
      validation_key = key,
      run_key = meta$run_key,
      .before = 1
    )
}))

validation_structure <- bind_rows(lapply(names(validation_runs), function(key) {
  run <- validation_runs[[key]]$structure
  meta <- validation_metadata[[key]]
  if (!isTRUE(run$success)) return(tibble())
  run$result$summary |>
    mutate(
      validation_key = key,
      run_key = meta$run_key,
      .before = 1
    )
}))

validation_metrics <- validation_metrics |>
  left_join(
    adjusted_results |>
      distinct(.data$run_key, .data$scenario, .data$n_areas, .data$prediction_scope, .data$method, .data$model_class),
    by = c("run_key", "method")
  ) |>
  distinct()

validation_metrics |>
  arrange(.data$scenario, .data$n_areas, .data$prediction_scope, .data$mae)
```

## Runtime And Diagnostics

```{r}
runtime_table <- bind_rows(
  baseline_summary |>
    transmute(
      scenario,
      n_areas,
      prediction_scope,
      method,
      model_class,
      success,
      elapsed_seconds,
      n_rows,
      n_fit_rows = NA_integer_,
      n_prediction_rows = NA_integer_,
      backend = NA_character_,
      observation_model = NA_character_,
      random_intercept = NA_character_,
      error,
      warnings
    ),
  multilevel_summary |>
    transmute(
      scenario,
      n_areas,
      prediction_scope,
      method,
      model_class,
      success,
      elapsed_seconds,
      n_rows,
      n_fit_rows,
      n_prediction_rows,
      backend,
      observation_model,
      random_intercept,
      error,
      warnings
    )
) |>
  arrange(.data$scenario, .data$n_areas, .data$prediction_scope, desc(.data$success), .data$elapsed_seconds)

runtime_table
```

```{r}
diagnostic_table <- multilevel_summary |>
  filter(.data$model_class == "bayesian") |>
  select(
    scenario,
    n_areas,
    prediction_scope,
    method,
    backend,
    observation_model,
    random_intercept,
    success,
    rhat_max,
    ess_bulk_min,
    ess_tail_min,
    n_eff_min,
    elapsed_seconds,
    error,
    warnings
  )

diagnostic_table
```

## Visual Summary

The plots use exported `debiasR::plot_validation_*()` functions. Methods are
ordered from best to worst by MAE for the selected scenario and area grid, with
the unadjusted raw MPD baseline kept at the end.

```{r}
prepare_plot_bundle <- function(metrics, residuals, distribution, structure, scenario, n_areas) {
  metric_slice <- metrics |>
    filter(.data$scenario == !!scenario, .data$n_areas == !!n_areas)

  if (nrow(metric_slice) == 0L) {
    return(NULL)
  }

  method_rank <- metric_slice |>
    filter(.data$comparison == "adjusted_vs_benchmark") |>
    group_by(.data$method) |>
    summarise(mae = min(.data$mae, na.rm = TRUE), .groups = "drop") |>
    arrange(.data$mae, .data$method) |>
    mutate(
      is_raw = .data$method == "raw_mpd",
      rank = row_number()
    )

  raw_methods <- method_rank$method[method_rank$is_raw]
  ordered_methods <- c(setdiff(method_rank$method, raw_methods), raw_methods)
  selected_run_keys <- unique(metric_slice$run_key)
  method_labels <- baseline_method_label
  missing_labels <- setdiff(ordered_methods, names(method_labels))
  method_labels <- c(method_labels, stats::setNames(missing_labels, missing_labels))

  order_tbl <- tibble(
    method = ordered_methods,
    method_order = seq_along(ordered_methods)
  )

  list(
    method_labels = method_labels,
    metrics = metric_slice |>
      left_join(order_tbl, by = "method") |>
      arrange(.data$method_order),
    residuals = residuals |>
      filter(.data$run_key %in% selected_run_keys, .data$method %in% ordered_methods) |>
      left_join(order_tbl, by = "method") |>
      arrange(.data$method_order),
    distribution = distribution |>
      filter(.data$run_key %in% selected_run_keys, .data$method %in% ordered_methods) |>
      left_join(order_tbl, by = "method") |>
      arrange(.data$method_order),
    structure = structure |>
      filter(.data$run_key %in% selected_run_keys, .data$method %in% ordered_methods) |>
      left_join(order_tbl, by = "method") |>
      arrange(.data$method_order),
    ordered_methods = ordered_methods
  )
}

plot_bundle <- prepare_plot_bundle(
  metrics = validation_metrics,
  residuals = validation_residuals,
  distribution = validation_distribution,
  structure = validation_structure,
  scenario = plot_scenario,
  n_areas = plot_n_areas
)

if (is.null(plot_bundle)) {
  message("No validation results available for plot_scenario = ", plot_scenario, " and plot_n_areas = ", plot_n_areas)
} else {
  n_pairs_col <- if ("n_pairs" %in% names(plot_bundle$metrics)) "n_pairs" else "n"
  plot_bundle$metrics |>
    select(method, mae, rmse, mape, n_pairs = all_of(n_pairs_col)) |>
    distinct()
}
```

### Overall Error Matrix

```{r}
if (!is.null(plot_bundle) && nrow(plot_bundle$metrics) > 0L) {
  debiasR::plot_validation_metrics(
    plot_bundle$metrics,
    error_measures = c("mae", "rmse", "mape"),
    method_labels = plot_bundle$method_labels
  )
}
```

### Residual Distributions

```{r}
if (!is.null(plot_bundle) && nrow(plot_bundle$residuals) > 0L) {
  debiasR::plot_validation_residuals(
    plot_bundle$residuals,
    residual = "absolute",
    method_labels = plot_bundle$method_labels,
    max_points = 2500
  )
}
```

### Flow Scatterplots

```{r}
if (!is.null(plot_bundle) && nrow(plot_bundle$residuals) > 0L) {
  debiasR::plot_validation_scatter(
    plot_bundle$residuals,
    method_labels = plot_bundle$method_labels
  )
}
```

### Residual Severity Bands

```{r}
if (!is.null(plot_bundle) && nrow(plot_bundle$residuals) > 0L) {
  debiasR::plot_validation_residual_bands(
    plot_bundle$residuals,
    method_labels = plot_bundle$method_labels,
    band_method = "sd",
    orientation = "horizontal"
  )
}
```

### Distributional Allocation

```{r}
if (!is.null(plot_bundle) && nrow(plot_bundle$distribution) > 0L) {
  debiasR::plot_validation_distribution(
    plot_bundle$distribution,
    method_labels = plot_bundle$method_labels
  )
}
```

### Residual Structure

```{r}
if (!is.null(plot_bundle) && nrow(plot_bundle$structure) > 0L) {
  debiasR::plot_validation_structure(
    plot_bundle$structure,
    method_labels = plot_bundle$method_labels
  )
}
```

## Decision Notes To Record

After each manual run, record the following points before increasing the grid or
sampling budget:

- which S1-S4 constructions were available for the selected geography;
- whether S4 was balanced or unbalanced across source-time cells;
- whether deterministic baselines, including selection-rate methods, completed;
- which frequentist smoke specifications failed and why;
- which Bayesian coverage-offset pilots completed, with runtime, warnings,
  `rhat_max`, `ess_bulk_min`, and `ess_tail_min`;
- whether the latent two-level backend completed for S3/S4 and whether its
  identifiability diagnostics are acceptable;
- which validation plot and metric should define the practical best-to-worst
  ordering;
- the largest area grid that remains feasible for manual RStudio testing.
