---
title: "Validation Visual Prototype Notebook"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This short notebook illustrates the prototype validation plotting functions
added to `debiasR`. The goal is visual iteration, not a final user-facing
vignette.

Project decision, 2026-06-19: these plotting functions will be used inside the
existing validation vignette. They will not be documented through a separate
validation-visualisation vignette. This notebook and its rendered output are
internal design/review notes.

The notebook uses deterministic simulated package data and lightweight
deterministic adjustment methods. It does not run Bayesian models. The LISA map
section uses user-supplied `sf` boundaries when they are available. The local
review defaults point to shared DEBIAS LAD and region boundary files only to
preview the visual; `debiasR` does not provide or infer boundary files.
`debiasRdata::lad_centroids` is used only as a lightweight LAD name-code lookup
for this simulated review example.

## Setup

```{r}
find_pkg_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "DESCRIPTION"))) {
      return(path)
    }

    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not find package root containing DESCRIPTION.")
    }
    path <- parent
  }
}

pkg_root <- find_pkg_root()
setwd(pkg_root)

if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(debiasR)
}

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  cat("Install ggplot2 to render the validation visual prototypes.\n")
  knitr::knit_exit()
}

library(dplyr)
```

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")

method_labels <- c(
  adjust_inverse_penetration = "Inverse penetration",
  adjust_raking_ratio = "Raking ratio",
  adjust_coefficient = "Coefficient"
)

method_subset <- names(method_labels)
benchmark_comparisons <- c("adjusted_vs_benchmark", "raw_vs_benchmark")
```

## Build Validation Inputs

```{r}
adjusted_outputs <- list(
  adjust_inverse_penetration = adjust_inverse_penetration(
    mpd_od_df = simulated_mpd.od,
    coverage_df = simulated_coverage,
    weight_by = "both"
  ),
  adjust_raking_ratio = adjust_raking_ratio(
    mpd_od_df = simulated_mpd.od,
    benchmark_od_df = simulated_benchmark.od
  ),
  adjust_coefficient = adjust_coefficient(
    mpd_od_df = simulated_mpd.od,
    benchmark_od_df = simulated_benchmark.od,
    model_family = "ols",
    level = "od"
  )
)

overall_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_overall(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    comparisons = "all",
    return_joined = FALSE
  )
}) |>
  stats::setNames(names(adjusted_outputs))

residual_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residuals(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    top_n = 12
  )
}) |>
  stats::setNames(names(adjusted_outputs))

distribution_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_distribution(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    comparisons = "all",
    weight_by = "benchmark_origin_total",
    method_name = method_name
  )
}) |>
  stats::setNames(names(adjusted_outputs))
```

For residual-structure diagnostics, this notebook uses a simple ring neighbour
table over the simulated areas. This is only an illustrative neighbour graph for
visual review.

```{r}
areas <- sort(unique(c(simulated_mpd.od$origin, simulated_mpd.od$destination)))
area_neighbors <- bind_rows(
  tibble::tibble(area = areas, neighbor = dplyr::lead(areas, default = areas[1])),
  tibble::tibble(area = areas, neighbor = dplyr::lag(areas, default = areas[length(areas)]))
) |>
  filter(area != neighbor)

area_covariates <- simulated_benchmark.od |>
  group_by(area = origin) |>
  summarise(benchmark_origin_flow_total = sum(flow), .groups = "drop")

structure_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residual_structure(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    area_neighbors = area_neighbors,
    covariate_df = area_covariates,
    covariate_col = "benchmark_origin_flow_total",
    local_moran = TRUE,
    local_moran_nsim = 99,
    local_moran_alpha = 0.05,
    local_moran_p_adjust = "none",
    local_moran_seed = 123
  )
}) |>
  stats::setNames(names(adjusted_outputs))

validation_lisa_boundaries <- NULL
validation_lisa_outline_boundaries <- NULL
validation_lisa_boundary_note <- NULL
validation_lisa_boundary_path <- Sys.getenv(
  "DEBIAS_LAD_BOUNDARY_PATH",
  unset = file.path(
    "/Volumes/DEBIAS/data/inputs/geographies/boundaries",
    "LAD_Dec_2021_GB_BFC_2022_simplified.gpkg"
  )
)
validation_lisa_outline_path <- Sys.getenv(
  "DEBIAS_REGION_BOUNDARY_PATH",
  unset = file.path(
    "/Volumes/DEBIAS/data/inputs/geographies/boundaries",
    "RGN_Dec_2021_EN_BFC_2022.gpkg"
  )
)

if (!requireNamespace("sf", quietly = TRUE)) {
  validation_lisa_boundary_note <- "Install sf and supply an sf boundary object to map LISA clusters."
} else if (!requireNamespace("debiasRdata", quietly = TRUE)) {
  validation_lisa_boundary_note <- "Install debiasRdata or provide a LAD name-code lookup to map these simulated LAD names."
} else if (!file.exists(validation_lisa_boundary_path)) {
  validation_lisa_boundary_note <- paste(
    "Set DEBIAS_LAD_BOUNDARY_PATH to a simplified LAD boundary GeoPackage.",
    "The shared DEBIAS volume path was not found."
  )
} else {
  data("lad_centroids", package = "debiasRdata")

  simulated_lad_lookup <- lad_centroids |>
    transmute(area_code = .data$area, area = .data$name) |>
    filter(.data$area %in% areas)

  validation_lisa_boundary_read <- tryCatch(
    sf::st_read(
      validation_lisa_boundary_path,
      quiet = TRUE
    ),
    error = function(error) error
  )

  if (inherits(validation_lisa_boundary_read, "error")) {
    validation_lisa_boundary_note <- paste(
      "Could not read the simplified LAD boundary GeoPackage:",
      conditionMessage(validation_lisa_boundary_read)
    )
  } else {
    validation_lisa_boundaries <- validation_lisa_boundary_read |>
      rename(area_code = code) |>
      inner_join(simulated_lad_lookup, by = "area_code") |>
      select(area, area_code)
  }

  if (!is.null(validation_lisa_boundaries)) {
    missing_lisa_areas <- setdiff(areas, validation_lisa_boundaries$area)
  } else {
    missing_lisa_areas <- character()
  }
  if (length(missing_lisa_areas) > 0L) {
    validation_lisa_boundary_note <- paste(
      "The simplified LAD boundaries do not cover:",
      paste(missing_lisa_areas, collapse = ", ")
    )
    validation_lisa_boundaries <- NULL
  }

  if (file.exists(validation_lisa_outline_path)) {
    validation_lisa_outline_read <- tryCatch(
      sf::st_read(
        validation_lisa_outline_path,
        quiet = TRUE
      ),
      error = function(error) error
    )

    if (!inherits(validation_lisa_outline_read, "error")) {
      validation_lisa_outline_boundaries <- validation_lisa_outline_read
    }
  }
}
```

## Level 1: Overall Fit Metrics

This chart compares methods by error measure in a single matrix. Fill colour is
scaled within each metric, so the visual supports method comparison without
putting MAE, RMSE, and MAPE on one raw axis. The cell labels preserve the
range of the within-metric relative-error score, such as 0-10, 11-20, and
91-100. The low-error band starts with pale yellow rather than white, reserving
white for page background and missing values. Users can select error measures
with `error_measures` and adjustment methods with `methods`.

The review default passes `comparisons = c("adjusted_vs_benchmark",
"raw_vs_benchmark")`. In the plot, adjusted-versus-benchmark values appear once
for each adjustment method. Raw-MPD-versus-benchmark values appear once as the
`Unadjusted raw MPD` baseline row.

```{r fig.height=6.5}
plot_validation_metrics(
  overall_results,
  error_measures = c("mae", "rmse", "mape"),
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels,
  relative_error_breaks = seq(0, 100, by = 10)
)
```

## Level 2 And Level 3: Residual Distributions

The violin plot shows the full distribution of signed flow differences. By
default, it shows adjusted-versus-benchmark residuals alongside the raw-MPD
versus benchmark baseline. Users can select other comparisons with
`comparisons` and select adjustment methods with `methods`, using method IDs or
display labels.

For benchmark comparisons, the signed residual is the benchmark flow minus the
flow being assessed. Positive values mean the benchmark is larger than the
adjusted or raw MPD flow; negative values mean the adjusted or raw MPD flow is
larger than the benchmark. The package convention is:

- `adjusted_vs_benchmark`: x is adjusted, y is benchmark, signed residual is
  benchmark minus adjusted.
- `raw_vs_benchmark`: x is raw MPD, y is benchmark, signed residual is benchmark
  minus raw MPD.
- `raw_vs_adjusted`: x is raw MPD, y is adjusted, signed residual is adjusted
  minus raw MPD.

Positive signed residuals therefore mean the reference series is larger than
the assessed series. The style follows the review reference: translucent
violins, overlaid OD-pair points, and sample-size labels above each method. The
categorical palette uses the shared supplied hex values.

```{r fig.height=5.2}
plot_validation_residuals(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 2 And Level 3: Pairwise Flow Scatterplots

The scatterplot uses the same comparison selector as the violin plot. By
default, it shows adjusted-versus-benchmark flows next to the raw-MPD versus
benchmark baseline. Users can select other comparisons, or pass a vector of
comparisons for a faceted view. When both benchmark comparisons are shown, the
facet labels identify whether the plotted estimate is an adjusted flow or a raw
MPD flow. The available axis orientations are:

- adjusted on x versus benchmark on y
- raw MPD on x versus benchmark on y
- raw MPD on x versus adjusted on y

The colour scale is blue-white-coral and centred on zero, using supplied
`#0000ff` and `#fa8775` endpoints. The default scale uses
robust symmetric limits from the 95th percentile of absolute signed
differences, with more extreme OD pairs squished to the end colours so the blue
and coral patterns remain visible. The `white_band` argument controls how much of
the scale around zero is treated as neutral.
Points use a fine neutral outline by default so near-zero differences remain
visible on the light background without adding a heavy border.

```{r fig.height=7.2}
plot_validation_scatter(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels,
  white_band = 0.22,
  point_alpha = 0.6,
  point_size = 0.9
)
```

## Level 3: Residual Outlier Bands

These stacked bar plots summarise how much of each method's OD-pair residual
distribution falls into severity bands. Each bar represents all plotted OD
pairs for one method-comparison combination. Each coloured segment is the share
of those OD pairs whose absolute residual falls into that severity band. A
method with more mass in higher-severity bands has more large OD-pair errors.
The unadjusted raw MPD baseline is included as a reference bar. The
colour-blind-friendly sequential palette orders residual severity from lower to
higher bands. The default orientation is horizontal so method labels remain
readable.

The standard-deviation version groups residual magnitudes by distance from zero
in standard-deviation units. It is useful for identifying residuals that are far
from the selected residual scale.

```{r residual-bands-sd, fig.height=4.5}
plot_validation_residual_bands(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  band_method = "sd",
  method_labels = method_labels
)
```

The quantile version uses shared cut points from the absolute-residual
distribution, so the method shares remain comparable while making the upper tail
easier to inspect.

```{r residual-bands-quantile, fig.height=4.5}
plot_validation_residual_bands(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  band_method = "quantile",
  method_labels = method_labels
)
```

## Level 4: Distributional Allocation Summary

This heatmap compares origin-conditioned destination-share allocation using
Jensen-Shannon divergence. Lower values indicate closer allocation fidelity. In
this implementation JSD uses natural logarithms, so values range from `0` for
identical destination-share distributions to `log(2)`, about `0.693`, for
maximally separated distributions. A value such as `0.158` is therefore not a
percentage; it is about 23% of the theoretical maximum JSD scale. The
user-facing default should use benchmark-origin-total weighted means because
that answers the system-level validation question: how much allocation error
affects the flow system as a whole. Unweighted means and medians remain useful
supporting diagnostics for the typical origin.

```{r fig.height=4}
plot_validation_distribution(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 4: Distribution Divergence Comparison

This bar view compares distributional allocation divergence directly by method.
Lower JSD values mean the destination-share allocation is closer to the
reference distribution. Showing adjusted-versus-benchmark next to raw
MPD-versus-benchmark makes the main validation question easier to read: did the
adjustment reduce allocation divergence from the benchmark? The full pairwise
matrix remains available with `plot_type = "heatmap"` when a method-development
audit needs all reference-by-comparison cells. Use `sort = "ascending"` or
`sort = "descending"` to order methods by divergence.

```{r fig.height=4}
plot_validation_distribution_pairwise(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels,
  sort = "ascending"
)
```

## Level 5: Residual-Structure Diagnostics

This dot-and-line summary shows scalar residual-structure diagnostics returned by
`validate_flow_residual_structure()`. It is useful for comparing methods on
global residual structure, but it deliberately does not show where local
clusters occur.

This diagnostic uses the same residual-structure logic as the assessing-bias
workflow, but it applies that logic to a different object. In assessing bias,
`validate_bias_residual_structure()` diagnoses population-count representation
residuals such as coverage-score or user-count residuals. Here,
`validate_flow_residual_structure()` diagnoses OD-flow validation residuals,
such as adjusted-versus-benchmark flow differences aggregated to areas. The
covariate panel below uses benchmark origin-flow totals as a simple illustrative
area-level covariate, so all three Level 5 panels contain values.

The shaded band marks values close to zero, and point labels show the rounded
diagnostic values. Values further from zero indicate stronger remaining residual
structure for that diagnostic, rather than better performance by default.

```{r fig.height=4}
plot_validation_structure(
  structure_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 5: LISA Cluster Map

Local Moran/LISA indicators need a separate map because the diagnostic is
spatially local. `debiasR` only provides the plotting option for validation
residual and LISA indicator outputs; users must provide the `sf` boundaries to
map those outputs. The prototype below uses user-supplied LAD boundaries for the
mapped areas when available, with an optional user-supplied region boundary file
drawn as a white outline. Only areas with Local Moran p-values at or below 0.05
are coloured by LISA cluster; non-significant areas are shown in light grey. LAD
internal outlines are suppressed.

```{r fig.height=5}
if (is.null(validation_lisa_boundaries)) {
  cat(validation_lisa_boundary_note, "\n")
} else {
  plot_validation_lisa_map(
    structure_results,
    boundaries = validation_lisa_boundaries,
    boundary_area_col = "area",
    comparisons = "adjusted_vs_benchmark",
    methods = method_subset,
    method_labels = method_labels,
    p_value_col = "p_value",
    p_value_threshold = 0.05,
    boundary_colour = NA,
    boundary_linewidth = 0,
    outline_boundaries = validation_lisa_outline_boundaries,
    outline_colour = "white",
    outline_linewidth = 0.45
  )
}
```

## Iteration Notes

The current proposal keeps the Level 1 method-by-error-measure matrix as the
default because the metrics have different units and the matrix keeps the raw
values visible. The Level 4 distribution comparison uses the bar view as the
preferred reading aid, with the compact heatmap retained as a secondary summary
for users who want a matrix-style display.
