---
title: "Validation Visual Prototype Notebook"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This short notebook illustrates the prototype validation plotting functions
added to `debiasR`. The goal is visual iteration, not a final user-facing
vignette.

The notebook uses deterministic simulated package data and lightweight
deterministic adjustment methods. It does not use the optional `debiasRdata`
package or run Bayesian models.

## Setup

```{r}
find_pkg_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "DESCRIPTION"))) {
      return(path)
    }

    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not find package root containing DESCRIPTION.")
    }
    path <- parent
  }
}

pkg_root <- find_pkg_root()
setwd(pkg_root)

if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(debiasR)
}

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  cat("Install ggplot2 to render the validation visual prototypes.\n")
  knitr::knit_exit()
}

library(dplyr)
```

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")

method_labels <- c(
  adjust_inverse_penetration = "Inverse penetration",
  adjust_raking_ratio = "Raking ratio",
  adjust_coefficient = "Coefficient"
)

method_subset <- names(method_labels)
```

## Build Validation Inputs

```{r}
adjusted_outputs <- list(
  adjust_inverse_penetration = adjust_inverse_penetration(
    mpd_od_df = simulated_mpd.od,
    coverage_df = simulated_coverage,
    weight_by = "both"
  ),
  adjust_raking_ratio = adjust_raking_ratio(
    mpd_od_df = simulated_mpd.od,
    benchmark_od_df = simulated_benchmark.od
  ),
  adjust_coefficient = adjust_coefficient(
    mpd_od_df = simulated_mpd.od,
    benchmark_od_df = simulated_benchmark.od,
    model_family = "ols",
    level = "od"
  )
)

overall_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_overall(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    return_joined = FALSE
  )
}) |>
  stats::setNames(names(adjusted_outputs))

residual_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residuals(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    top_n = 12
  )
}) |>
  stats::setNames(names(adjusted_outputs))

distribution_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_distribution(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    comparisons = "all",
    weight_by = "benchmark_origin_total",
    method_name = method_name
  )
}) |>
  stats::setNames(names(adjusted_outputs))
```

For residual-structure diagnostics, this notebook uses a simple ring neighbour
table over the simulated areas. This is only an illustrative neighbour graph for
visual review.

```{r}
areas <- sort(unique(c(simulated_mpd.od$origin, simulated_mpd.od$destination)))
area_neighbors <- bind_rows(
  tibble::tibble(area = areas, neighbor = dplyr::lead(areas, default = areas[1])),
  tibble::tibble(area = areas, neighbor = dplyr::lag(areas, default = areas[length(areas)]))
) |>
  filter(area != neighbor)

structure_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residual_structure(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    area_neighbors = area_neighbors,
    local_moran = TRUE,
    local_moran_nsim = 99,
    local_moran_alpha = 1,
    local_moran_p_adjust = "none",
    local_moran_seed = 123
  )
}) |>
  stats::setNames(names(adjusted_outputs))

validation_lisa_boundaries <- NULL
if (requireNamespace("sf", quietly = TRUE)) {
  grid_width <- ceiling(sqrt(length(areas)))
  validation_lisa_boundaries <- tibble::tibble(
    area = areas,
    x = (seq_along(areas) - 1) %% grid_width,
    y = floor((seq_along(areas) - 1) / grid_width)
  ) |>
    mutate(
      wkt = sprintf(
        "POLYGON ((%s %s, %s %s, %s %s, %s %s, %s %s))",
        x, y,
        x + 0.95, y,
        x + 0.95, y + 0.95,
        x, y + 0.95,
        x, y
      )
    ) |>
    sf::st_as_sf(wkt = "wkt", crs = 27700)
}
```

## Level 1: Overall Fit Metrics

This chart compares methods by error measure in a single matrix. Fill colour is
scaled within each metric, so the visual supports method comparison without
putting MAE, RMSE, and MAPE on one raw axis. The cell labels preserve the
original values. The prototype uses the shared validation visual style:
white background, quiet axes, and a bright yellow-to-blue palette. The plotted
colour hex values are supplied from the visual review palette: `#ffd700`,
`#ffb14e`, `#fa8775`, `#ea5f94`, `#cd34b5`, `#9d02d7`, and `#0000ff`.
Users can select error measures with `error_measures` and adjustment methods
with `methods`.

```{r}
plot_validation_metrics(
  overall_results,
  error_measures = c("mae", "rmse", "mape"),
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 2 And Level 3: Residual Distributions

The violin plot shows the full distribution of signed flow differences. By
default, it focuses on adjusted versus benchmark flows because this is the main
validation comparison. Users can select other comparisons with `comparisons`
and select adjustment methods with `methods`, using method IDs or display
labels. The y-axis names the selected `Y - X` comparison explicitly. The
package convention is:

- `adjusted_vs_benchmark`: x is adjusted, y is benchmark, signed residual is
  benchmark minus adjusted.
- `raw_vs_benchmark`: x is raw MPD, y is benchmark, signed residual is benchmark
  minus raw MPD.
- `raw_vs_adjusted`: x is raw MPD, y is adjusted, signed residual is adjusted
  minus raw MPD.

Positive signed residuals therefore mean the y/reference series is larger than
the x/baseline series. The style follows the review reference: translucent
violins, overlaid OD-pair points, and sample-size labels above each method. The
categorical palette uses the shared supplied hex values.

```{r fig.height=5.2}
plot_validation_residuals(
  residual_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 2 And Level 3: Pairwise Flow Scatterplots

The scatterplot uses the same comparison selector as the violin plot. By
default, it focuses on adjusted versus benchmark flows. Users can select other
comparisons, or pass a vector of comparisons for a faceted view. For a single
comparison, the axis labels name the X and Y series directly; for multiple
comparisons, each facet header states the X and Y series. The available axis
orientations are:

- adjusted on x versus benchmark on y
- raw MPD on x versus benchmark on y
- raw MPD on x versus adjusted on y

The colour scale is blue-white-coral and centred on zero, using supplied
`#0000ff` and `#fa8775` endpoints. The default scale uses
robust symmetric limits from the 95th percentile of absolute `Y - X`
differences, with more extreme OD pairs squished to the end colours so the blue
and coral patterns remain visible. The `white_band` argument controls how much of
the scale around zero is treated as neutral.
Points use a fine neutral outline by default so near-zero differences remain
visible on the light background without adding a heavy border.

```{r fig.height=6.5}
plot_validation_scatter(
  residual_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels,
  white_band = 0.22,
  point_alpha = 0.6,
  point_size = 0.9
)
```

## Level 3: Residual Outlier Bands

These stacked bar plots summarise how much of each method's OD-pair residual
distribution falls into severity bands. The unadjusted raw MPD baseline is
included as a reference bar. The colour-blind-friendly sequential palette orders
residual severity from lower to higher bands. The default orientation is
horizontal so method labels remain readable.

The standard-deviation version is useful for identifying residuals that are far
from the pooled adjusted-residual scale.

```{r residual-bands-sd, fig.height=4.5}
plot_validation_residual_bands(
  residual_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  band_method = "sd",
  method_labels = method_labels
)
```

The quantile version uses shared absolute-residual cut points, so the method
shares remain comparable while making the upper tail easier to inspect.

```{r residual-bands-quantile, fig.height=4.5}
plot_validation_residual_bands(
  residual_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  band_method = "quantile",
  method_labels = method_labels
)
```

## Level 4: Distributional Allocation Summary

This heatmap compares origin-conditioned destination-share allocation using
Jensen-Shannon divergence. Lower values indicate closer allocation fidelity.
The user-facing default should use benchmark-origin-total weighted means because
that answers the system-level validation question: how much allocation error
affects the flow system as a whole. Unweighted means and medians remain useful
supporting diagnostics for the typical origin.

```{r fig.height=4}
plot_validation_distribution(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 4: Pairwise Distribution Heatmap

This view turns the requested pairwise comparisons into a reference-by-comparison
matrix. JSD is symmetric, so the prototype mirrors off-diagonal entries and
adds zeros on the diagonal. Passing `comparisons = "all"` gives the full matrix
over adjusted, raw MPD, and benchmark distributions.

```{r fig.height=5}
plot_validation_distribution_pairwise(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = "all",
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 5: Residual-Structure Diagnostics

This dot-and-line summary shows scalar residual-structure diagnostics returned by
`validate_flow_residual_structure()`. It is useful for comparing methods on
global residual structure, but it deliberately does not show where local
clusters occur.

```{r fig.height=4}
plot_validation_structure(
  structure_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 5: LISA Cluster Map

Local Moran/LISA indicators need a separate map because the diagnostic is
spatially local. `debiasR` should require users to supply boundaries, for
example LAD or MSOA polygons, and then join the LISA output by area identifier.
The toy boundaries below are only for visual review of the prototype. To expose
the full cluster palette in this deterministic simulated example, the notebook
uses a deliberately relaxed LISA threshold above; final user-facing examples
should use the statistical default unless they are presenting a sensitivity
check.

```{r fig.height=5}
if (is.null(validation_lisa_boundaries)) {
  cat("Install sf and supply an sf boundary object to map LISA clusters.\n")
} else {
  plot_validation_lisa_map(
    structure_results,
    boundaries = validation_lisa_boundaries,
    boundary_area_col = "area",
    comparisons = "adjusted_vs_benchmark",
    methods = method_subset,
    method_labels = method_labels
  )
}
```

## Iteration Notes

The remaining design choice to review is:

- Whether Level 1 should keep the method-by-error-measure matrix as the default
  or add an alternate grouped bar view for methods with shared units.
