---
title: "Validation Visual Prototype Notebook"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This short notebook illustrates the prototype validation plotting functions
added to `debiasR`. The goal is visual iteration, not a final user-facing
vignette.

The notebook uses deterministic simulated package data and lightweight
deterministic adjustment methods. It does not run Bayesian models. The LISA map
section uses user-supplied `sf` boundaries when they are available. The local
review defaults point to shared DEBIAS LAD and region boundary files only to
preview the visual; `debiasR` does not provide or infer boundary files.
`debiasRdata::lad_centroids` is used only as a lightweight LAD name-code lookup
for this simulated review example.

## Setup

```{r}
find_pkg_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "DESCRIPTION"))) {
      return(path)
    }

    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not find package root containing DESCRIPTION.")
    }
    path <- parent
  }
}

pkg_root <- find_pkg_root()
setwd(pkg_root)

if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(debiasR)
}

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  cat("Install ggplot2 to render the validation visual prototypes.\n")
  knitr::knit_exit()
}

library(dplyr)
```

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")

method_labels <- c(
  adjust_inverse_penetration = "Inverse penetration",
  adjust_raking_ratio = "Raking ratio",
  adjust_coefficient = "Coefficient"
)

method_subset <- names(method_labels)
benchmark_comparisons <- c("adjusted_vs_benchmark", "raw_vs_benchmark")
```

## Build Validation Inputs

```{r}
adjusted_outputs <- list(
  adjust_inverse_penetration = adjust_inverse_penetration(
    mpd_od_df = simulated_mpd.od,
    coverage_df = simulated_coverage,
    weight_by = "both"
  ),
  adjust_raking_ratio = adjust_raking_ratio(
    mpd_od_df = simulated_mpd.od,
    benchmark_od_df = simulated_benchmark.od
  ),
  adjust_coefficient = adjust_coefficient(
    mpd_od_df = simulated_mpd.od,
    benchmark_od_df = simulated_benchmark.od,
    model_family = "ols",
    level = "od"
  )
)

overall_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_overall(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    comparisons = "all",
    return_joined = FALSE
  )
}) |>
  stats::setNames(names(adjusted_outputs))

residual_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residuals(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    top_n = 12
  )
}) |>
  stats::setNames(names(adjusted_outputs))

distribution_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_distribution(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    comparisons = "all",
    weight_by = "benchmark_origin_total",
    method_name = method_name
  )
}) |>
  stats::setNames(names(adjusted_outputs))
```

For residual-structure diagnostics, this notebook uses a simple ring neighbour
table over the simulated areas. This is only an illustrative neighbour graph for
visual review.

```{r}
areas <- sort(unique(c(simulated_mpd.od$origin, simulated_mpd.od$destination)))
area_neighbors <- bind_rows(
  tibble::tibble(area = areas, neighbor = dplyr::lead(areas, default = areas[1])),
  tibble::tibble(area = areas, neighbor = dplyr::lag(areas, default = areas[length(areas)]))
) |>
  filter(area != neighbor)

structure_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residual_structure(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    area_neighbors = area_neighbors,
    local_moran = TRUE,
    local_moran_nsim = 99,
    local_moran_alpha = 0.05,
    local_moran_p_adjust = "none",
    local_moran_seed = 123
  )
}) |>
  stats::setNames(names(adjusted_outputs))

validation_lisa_boundaries <- NULL
validation_lisa_outline_boundaries <- NULL
validation_lisa_boundary_note <- NULL
validation_lisa_boundary_path <- Sys.getenv(
  "DEBIAS_LAD_BOUNDARY_PATH",
  unset = file.path(
    "/Volumes/DEBIAS/data/inputs/geographies/boundaries",
    "LAD_Dec_2021_GB_BFC_2022_simplified.gpkg"
  )
)
validation_lisa_outline_path <- Sys.getenv(
  "DEBIAS_REGION_BOUNDARY_PATH",
  unset = file.path(
    "/Volumes/DEBIAS/data/inputs/geographies/boundaries",
    "RGN_Dec_2021_EN_BFC_2022.gpkg"
  )
)

if (!requireNamespace("sf", quietly = TRUE)) {
  validation_lisa_boundary_note <- "Install sf and supply an sf boundary object to map LISA clusters."
} else if (!requireNamespace("debiasRdata", quietly = TRUE)) {
  validation_lisa_boundary_note <- "Install debiasRdata or provide a LAD name-code lookup to map these simulated LAD names."
} else if (!file.exists(validation_lisa_boundary_path)) {
  validation_lisa_boundary_note <- paste(
    "Set DEBIAS_LAD_BOUNDARY_PATH to a simplified LAD boundary GeoPackage.",
    "The shared DEBIAS volume path was not found."
  )
} else {
  data("lad_centroids", package = "debiasRdata")

  simulated_lad_lookup <- lad_centroids |>
    transmute(area_code = .data$area, area = .data$name) |>
    filter(.data$area %in% areas)

  validation_lisa_boundary_read <- tryCatch(
    sf::st_read(
      validation_lisa_boundary_path,
      quiet = TRUE
    ),
    error = function(error) error
  )

  if (inherits(validation_lisa_boundary_read, "error")) {
    validation_lisa_boundary_note <- paste(
      "Could not read the simplified LAD boundary GeoPackage:",
      conditionMessage(validation_lisa_boundary_read)
    )
  } else {
    validation_lisa_boundaries <- validation_lisa_boundary_read |>
      rename(area_code = code) |>
      inner_join(simulated_lad_lookup, by = "area_code") |>
      select(area, area_code)
  }

  if (!is.null(validation_lisa_boundaries)) {
    missing_lisa_areas <- setdiff(areas, validation_lisa_boundaries$area)
  } else {
    missing_lisa_areas <- character()
  }
  if (length(missing_lisa_areas) > 0L) {
    validation_lisa_boundary_note <- paste(
      "The simplified LAD boundaries do not cover:",
      paste(missing_lisa_areas, collapse = ", ")
    )
    validation_lisa_boundaries <- NULL
  }

  if (file.exists(validation_lisa_outline_path)) {
    validation_lisa_outline_read <- tryCatch(
      sf::st_read(
        validation_lisa_outline_path,
        quiet = TRUE
      ),
      error = function(error) error
    )

    if (!inherits(validation_lisa_outline_read, "error")) {
      validation_lisa_outline_boundaries <- validation_lisa_outline_read
    }
  }
}
```

## Level 1: Overall Fit Metrics

This chart compares methods by error measure in a single matrix. Fill colour is
scaled within each metric, so the visual supports method comparison without
putting MAE, RMSE, and MAPE on one raw axis. The cell labels preserve the
original values. The legend is discrete: each colour represents a 10-point
range of the within-metric relative-error score, such as 0-10, 11-20 and
91-100. The prototype uses the shared validation visual style: white
background, quiet axes and a yellow-to-blue palette. The lowest relative-error
band starts with a pale yellow rather than white, so low error is visible as
data rather than appearing as an empty cell. Users can select error measures
with `error_measures` and adjustment methods with `methods`. The review default
includes the raw MPD versus benchmark comparison as an unadjusted baseline row.

```{r fig.height=6.5}
plot_validation_metrics(
  overall_results,
  error_measures = c("mae", "rmse", "mape"),
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels,
  palette = c(
    "#fff7bc", "#ffd700", "#ffb14e", "#fa8775",
    "#ea5f94", "#cd34b5", "#9d02d7", "#0000ff"
  ),
  relative_error_breaks = seq(0, 100, by = 10)
) +
  ggplot2::scale_x_discrete(
    labels = c(
      "Mean absolute error" = "Mean absolute\nerror",
      "Root mean squared error" = "Root mean\nsquared error",
      "Mean absolute percentage error" = "Mean absolute\npercentage error"
    )
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      title = "Relative error\nscore (%)",
      ncol = 1,
      byrow = FALSE,
      override.aes = list(
        alpha = 1,
        colour = "#A7ADB7",
        linewidth = 0.25
      )
    )
  ) +
  ggplot2::theme(
    legend.position = "right",
    legend.box = "vertical"
  )
```

## Level 2 And Level 3: Residual Distributions

The violin plot shows the full distribution of signed flow differences. By
default, it shows adjusted versus benchmark flows alongside the raw MPD versus
benchmark baseline. Users can select other comparisons with `comparisons` and
select adjustment methods with `methods`, using method IDs or display labels.
The y-axis names the selected `Y - X` comparison explicitly. The
package convention is:

- `adjusted_vs_benchmark`: x is adjusted, y is benchmark, signed residual is
  benchmark minus adjusted.
- `raw_vs_benchmark`: x is raw MPD, y is benchmark, signed residual is benchmark
  minus raw MPD.
- `raw_vs_adjusted`: x is raw MPD, y is adjusted, signed residual is adjusted
  minus raw MPD.

Positive signed residuals therefore mean the y/reference series is larger than
the x/baseline series. The style follows the review reference: translucent
violins, overlaid OD-pair points, and sample-size labels above each method. The
categorical palette uses the shared supplied hex values.

```{r fig.height=5.2}
plot_validation_residuals(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels
)
```

## Level 2 And Level 3: Pairwise Flow Scatterplots

The scatterplot uses the same comparison selector as the violin plot. By
default, it shows adjusted versus benchmark flows next to the raw MPD versus
benchmark baseline. Users can select other comparisons, or pass a vector of
comparisons for a faceted view. For a single
comparison, the axis labels name the X and Y series directly; for multiple
comparisons, each facet header states the X and Y series. The available axis
orientations are:

- adjusted on x versus benchmark on y
- raw MPD on x versus benchmark on y
- raw MPD on x versus adjusted on y

The colour scale is blue-white-coral and centred on zero, using supplied
`#0000ff` and `#fa8775` endpoints. The default scale uses
robust symmetric limits from the 95th percentile of absolute `Y - X`
differences, with more extreme OD pairs squished to the end colours so the blue
and coral patterns remain visible. The `white_band` argument controls how much of
the scale around zero is treated as neutral.
Points use a fine neutral outline by default so near-zero differences remain
visible on the light background without adding a heavy border.

```{r fig.height=7.2}
plot_validation_scatter(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels,
  white_band = 0.22,
  point_alpha = 0.6,
  point_size = 0.9
) +
  ggplot2::labs(
    x = "Adjusted or Unadjusted MPD flow",
    y = "Benchmark flow"
  )
```

## Level 3: Residual Outlier Bands

These stacked bar plots summarise how much of each method's OD-pair residual
distribution falls into severity bands. The unadjusted raw MPD baseline is
included as a reference bar. The colour-blind-friendly sequential palette orders
residual severity from lower to higher bands. The default orientation is
horizontal so method labels remain readable.

The standard-deviation version is useful for identifying residuals that are far
from the pooled adjusted-residual scale.

```{r residual-bands-sd, fig.height=4.5}
plot_validation_residual_bands(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  band_method = "sd",
  method_labels = method_labels
) +
  ggplot2::guides(fill = ggplot2::guide_legend(ncol = 1)) +
  ggplot2::theme(
    legend.position = "right",
    legend.box = "vertical"
  )
```

The quantile version uses shared absolute-residual cut points, so the method
shares remain comparable while making the upper tail easier to inspect.

```{r residual-bands-quantile, fig.height=4.5}
plot_validation_residual_bands(
  residual_results,
  comparisons = benchmark_comparisons,
  methods = method_subset,
  band_method = "quantile",
  method_labels = method_labels
) +
  ggplot2::guides(fill = ggplot2::guide_legend(ncol = 1)) +
  ggplot2::theme(
    legend.position = "right",
    legend.box = "vertical"
  )
```

## Level 4: Distributional Allocation Summary

This heatmap compares origin-conditioned destination-share allocation using
Jensen-Shannon divergence. Lower values indicate closer allocation fidelity.
The user-facing default should use benchmark-origin-total weighted means because
that answers the system-level validation question: how much allocation error
affects the flow system as a whole. Unweighted means and medians remain useful
supporting diagnostics for the typical origin.

```{r fig.height=4}
suppressMessages({
  plot_validation_distribution(
    distribution_results,
    metric = "jsd",
    value = "weighted_mean",
    comparisons = benchmark_comparisons,
    methods = method_subset,
    method_labels = method_labels
  ) +
    ggplot2::scale_fill_gradient(
      low = "#fff7bc",
      high = "#0000ff",
      name = "JSD"
    )
})
```

## Level 4: Distribution Divergence Comparison

This view compares distributional allocation divergence directly by method.
Lower JSD values mean the destination-share allocation is closer to the
reference distribution. Showing adjusted-versus-benchmark next to raw
MPD-versus-benchmark makes the main validation question easier to read: did the
adjustment reduce allocation divergence from the benchmark? The full pairwise
matrix is still available with `plot_type = "heatmap"` when a method-development
audit needs all reference-by-comparison cells. Use `sort = "ascending"` or
`sort = "descending"` to order methods by divergence.

```{r fig.height=4}
plot_validation_distribution_pairwise(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = benchmark_comparisons,
  methods = method_subset,
  method_labels = method_labels,
  sort = "ascending"
)
```

## Level 5: Residual-Structure Diagnostics

This dot-and-line summary shows scalar residual-structure diagnostics returned by
`validate_flow_residual_structure()`. It is useful for comparing methods on
global residual structure, but it deliberately does not show where local
clusters occur.

```{r fig.height=4}
structure_axis_values <- unlist(lapply(structure_results, function(result) {
  summary <- result$summary
  as.numeric(unlist(summary[c(
    "pearson_residual_benchmark_flow",
    "moran_i",
    "pearson_residual_covariate"
  )]))
}))

structure_metric_labels <- c(
  pearson_residual_benchmark_flow = "Residual-flow correlation (r)",
  moran_i = "Moran's I of residuals",
  pearson_residual_covariate = "Residual-covariate correlation (r)"
)

structure_axis_limits <- lapply(names(structure_metric_labels), function(metric) {
  metric_values <- unlist(lapply(structure_results, function(result) {
    result$summary[[metric]]
  }))
  limit <- max(abs(metric_values), na.rm = TRUE)
  if (!is.finite(limit)) {
    limit <- 0.25
  }
  limit <- max(limit * 1.15, 0.25)
  limit <- min(limit, 1)
  tibble::tibble(
    comparison_label = "Adjusted vs benchmark",
    metric_label = structure_metric_labels[[metric]],
    x = c(-limit, limit),
    y = method_labels[[1]]
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(
    metric_label = factor(
      .data$metric_label,
      levels = c(
        "Moran's I of residuals",
        "Residual-flow correlation (r)",
        "Residual-covariate correlation (r)"
      )
    )
  )

structure_plot <- plot_validation_structure(
  structure_results,
  comparisons = "adjusted_vs_benchmark",
  methods = method_subset,
  method_labels = method_labels
)

structure_plot$data <- structure_plot$data |>
  dplyr::mutate(
    metric_label = dplyr::recode(
      .data$metric_label,
      "Moran's I" = "Moran's I of residuals",
      "Residual vs benchmark-flow r" = "Residual-flow correlation (r)",
      "Residual vs covariate r" = "Residual-covariate correlation (r)"
    ),
    metric_label = factor(
      .data$metric_label,
      levels = c(
        "Moran's I of residuals",
        "Residual-flow correlation (r)",
        "Residual-covariate correlation (r)"
      )
    )
  )

structure_plot +
  ggplot2::geom_blank(
    data = structure_axis_limits,
    ggplot2::aes(x = .data$x, y = .data$y),
    inherit.aes = FALSE
  )
```

## Level 5: LISA Cluster Map

Local Moran/LISA indicators need a separate map because the diagnostic is
spatially local. `debiasR` only provides the plotting option for validation
residual and LISA indicator outputs; users must provide the `sf` boundaries to
map those outputs. The prototype below uses user-supplied LAD boundaries for the
mapped areas when available, with an optional user-supplied region boundary file
drawn as a white outline. Only areas with Local Moran p-values at or below 0.05
are coloured by LISA cluster; non-significant areas are shown in light grey. LAD
internal outlines are suppressed.

```{r fig.height=5}
if (is.null(validation_lisa_boundaries)) {
  cat(validation_lisa_boundary_note, "\n")
} else {
  plot_validation_lisa_map(
    structure_results,
    boundaries = validation_lisa_boundaries,
    boundary_area_col = "area",
    comparisons = "adjusted_vs_benchmark",
    methods = method_subset,
    method_labels = method_labels,
    p_value_col = "p_value",
    p_value_threshold = 0.05,
    boundary_colour = NA,
    boundary_linewidth = 0,
    outline_boundaries = validation_lisa_outline_boundaries,
    outline_colour = "white",
    outline_linewidth = 0.45
  )
}
```

## Iteration Notes

The remaining design choice to review is:

- Whether Level 1 should keep the method-by-error-measure matrix as the default
  or add an alternate grouped bar view for methods with shared units.
