---
title: "debiasR: Method Comparison Walkthrough"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This vignette compares the implemented adjustment methods in `debiasR` and
includes the multilevel helper as the main experimental prototype:

1. `adjust_inverse_penetration`
2. `adjust_selection_rate`
3. `adjust_selection_rate2`
4. `adjust_raking_ratio`
5. `adjust_coefficient`
6. `adjust_multilevel_bayes` with `model_engine = "frequentist"` (temporary S1 placeholder for the Bayesian model)

It uses a fast but fair comparison protocol:

1. same stratified OD subset for all methods per replicate,
2. repeated over multiple seeds,
3. Method 6 fit with the frequentist development engine while S1-S4 Bayesian implementation is completed.

`adjust_multilevel_bayes()` is the main methodological innovation. Treat Method
6 results here as a fast scenario-contract check, not final Bayesian inference.
The example below uses S1 data: one MPD source observed during one time unit.

## Setup

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)
library(ggplot2)
library(forcats)
library(tidyr)
library(readr)
library(stringr)
library(purrr)

if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("Install 'patchwork' to render the multi-panel comparison plots.")
}
```

## Data

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")
data("simulated_covariates", package = "debiasR")
data("simulated_distance", package = "debiasR")

tibble::tibble(
  dataset = c(
    "simulated_mpd.od",
    "simulated_benchmark.od",
    "simulated_coverage",
    "simulated_covariates",
    "simulated_distance"
  ),
  rows = c(
    nrow(simulated_mpd.od),
    nrow(simulated_benchmark.od),
    nrow(simulated_coverage),
    nrow(simulated_covariates),
    nrow(simulated_distance)
  )
)

# coverage_df schema used by methods in this walkthrough
names(simulated_coverage)
```

## Simulation Fidelity Check

This historical notebook originally compared the packaged simulated fixtures
with private raw calibration CSVs. Those raw CSVs are no longer distributed in
the public repository. Use the bundled `simulated_*` objects for lightweight
method comparisons, and use `debiasRdata` for audited public empirical examples.

## Fair-Lite Settings

```{r}
n_sub <- 12000
fair_seeds <- c(123, 456, 789)

make_flow_strata <- function(flow_vec) {
  pos <- flow_vec[flow_vec > 0]
  q50 <- stats::quantile(pos, 0.50, na.rm = TRUE)
  q90 <- stats::quantile(pos, 0.90, na.rm = TRUE)

  dplyr::case_when(
    flow_vec <= 0 ~ "zero",
    flow_vec <= q50 ~ "low",
    flow_vec <= q90 ~ "mid",
    TRUE ~ "high"
  )
}

draw_fair_subset <- function(mpd_df, n_target, seed) {
  set.seed(seed)
  tmp <- mpd_df |>
    dplyr::mutate(stratum = make_flow_strata(.data$flow))

  counts <- tmp |>
    dplyr::count(.data$stratum, name = "n_stratum") |>
    dplyr::mutate(
      n_take = pmax(1L, round(n_target * .data$n_stratum / sum(.data$n_stratum)))
    )

  # Force exact sample size by adjusting the largest stratum.
  diff_n <- n_target - sum(counts$n_take)
  if (diff_n != 0) {
    idx <- which.max(counts$n_stratum)
    counts$n_take[idx] <- pmax(1L, counts$n_take[idx] + diff_n)
  }

  sampled <- lapply(counts$stratum, function(s) {
    df_s <- tmp[tmp$stratum == s, , drop = FALSE]
    n_take_s <- counts$n_take[counts$stratum == s][1]
    n_take_s <- min(n_take_s, nrow(df_s))
    df_s[sample.int(nrow(df_s), size = n_take_s), , drop = FALSE]
  }) |>
    dplyr::bind_rows()

  sampled |>
    dplyr::select(-.data$stratum)
}
```

## Validation functions

```{r}
collect_metrics <- function(method_name, adj_df, benchmark_df, seed_id) {
  v <- validate_flow_overall(
    adj_df = adj_df,
    benchmark_od_df = benchmark_df,
    method_name = method_name,
    return_joined = FALSE
  )

  tibble::tibble(
    seed = seed_id,
    method = v$method,
    n = v$n,
    pearson_r = v$pearson_r,
    spearman_rho = v$spearman_rho,
    rmse = v$rmse,
    mae = v$mae,
    mape = v$mape,
    r_squared = v$r_squared
  )
}

build_compare_df <- function(adj_df, mpd_df, benchmark_df) {
  adj_df |>
    dplyr::select(origin, destination, flow_adj) |>
    dplyr::inner_join(
      mpd_df |>
        dplyr::select(origin, destination, flow_mpd = flow),
      by = c("origin", "destination")
    ) |>
    dplyr::inner_join(
      benchmark_df |>
        dplyr::select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    )
}

plot_method_triptych <- function(method_label, adj_df, mpd_df, benchmark_df) {
  df <- build_compare_df(adj_df, mpd_df = mpd_df, benchmark_df = benchmark_df)

  p1 <- ggplot(df, aes(x = flow_mpd, y = flow_adj)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#2B8CBE") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "Adjusted vs MPD",
      x = "Mobile-phone-derived flow",
      y = "Adjusted flow"
    )

  p2 <- ggplot(df, aes(x = flow_bench, y = flow_adj)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#41AE76") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "Adjusted vs Benchmark",
      x = "Benchmark flow",
      y = "Adjusted flow"
    )

  p3 <- ggplot(df, aes(x = flow_bench, y = flow_mpd)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#756BB1") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "MPD vs Benchmark",
      x = "Benchmark flow",
      y = "Mobile-phone-derived flow"
    )

  patchwork::wrap_plots(p1, p2, p3, nrow = 1) +
    patchwork::plot_annotation(title = method_label)
}
```

## Run Methods (Fair-Lite)

```{r}
run_methods_once <- function(seed_id) {
  mpd_sub <- draw_fair_subset(simulated_mpd.od, n_target = n_sub, seed = seed_id)
  benchmark_sub <- simulated_benchmark.od |>
    dplyr::inner_join(
      mpd_sub |> dplyr::select(origin, destination),
      by = c("origin", "destination")
    )

  distance_sub <- simulated_distance |>
    dplyr::inner_join(
      mpd_sub |> dplyr::select(origin, destination),
      by = c("origin", "destination")
    )

  adj_m1 <- adjust_inverse_penetration(
    mpd_od_df = mpd_sub,
    coverage_df = simulated_coverage,
    weight_by = "both"
  )

  adj_m2 <- adjust_selection_rate(
    mpd_od_df = mpd_sub,
    coverage_df = simulated_coverage,
    covariates_df = simulated_covariates,
    income_col = "income_norm",
    weight_by = "origin",
    benchmark_od_df = benchmark_sub,
    calibration_aggregate = "origin"
  )

  adj_m3 <- adjust_selection_rate2(
    mpd_od_df = mpd_sub,
    coverage_df = simulated_coverage,
    weight_by = "origin",
    benchmark_od_df = benchmark_sub,
    calibration_aggregate = "origin"
  )

  adj_m4 <- adjust_raking_ratio(
    mpd_od_df = mpd_sub,
    benchmark_od_df = benchmark_sub
  )

  adj_m5 <- adjust_coefficient(
    mpd_od_df = mpd_sub,
    benchmark_od_df = benchmark_sub,
    model_family = "ols",
    level = "od"
  )

  mpd_s1 <- mpd_sub |>
    dplyr::mutate(
      mpd_source = "operator_a",
      mpd_time = "2021_q1"
    )
  coverage_s1 <- simulated_coverage |>
    dplyr::mutate(
      mpd_source = "operator_a",
      mpd_time = "2021_q1"
    )

  adj_m6 <- adjust_multilevel_bayes(
    mpd_od_df = mpd_s1,
    coverage_df = coverage_s1,
    covariates_df = simulated_covariates,
    distance_df = distance_sub,
    formula = flow ~ income_norm_o + income_norm_d + log_distance + bias_e_origin,
    model_engine = "frequentist",
    scenario = "s1",
    source_col = "mpd_source",
    time_col = "mpd_time",
    repeated_observation = "none",
    random_intercept = "none",
    model_family = "poisson",
    flow_adj_summary = "median"
  )

  metrics <- dplyr::bind_rows(
    collect_metrics("adjust_inverse_penetration", adj_m1, benchmark_sub, seed_id),
    collect_metrics("adjust_selection_rate", adj_m2, benchmark_sub, seed_id),
    collect_metrics("adjust_selection_rate2", adj_m3, benchmark_sub, seed_id),
    collect_metrics("adjust_raking_ratio", adj_m4, benchmark_sub, seed_id),
    collect_metrics("adjust_coefficient", adj_m5, benchmark_sub, seed_id),
    collect_metrics("adjust_multilevel_bayes", adj_m6, benchmark_sub, seed_id)
  )

  list(
    mpd_sub = mpd_sub,
    benchmark_sub = benchmark_sub,
    adj = list(
      adjust_inverse_penetration = adj_m1,
      adjust_selection_rate = adj_m2,
      adjust_selection_rate2 = adj_m3,
      adjust_raking_ratio = adj_m4,
      adjust_coefficient = adj_m5,
      adjust_multilevel_bayes = adj_m6
    ),
    metrics = metrics
  )
}

res_list <- purrr::map(fair_seeds, run_methods_once)
metrics_tbl <- dplyr::bind_rows(purrr::map(res_list, "metrics"))

metrics_summary <- metrics_tbl |>
  dplyr::group_by(.data$method) |>
  dplyr::summarise(
    n_mean = mean(.data$n),
    pearson_r_mean = mean(.data$pearson_r),
    pearson_r_sd = stats::sd(.data$pearson_r),
    rmse_mean = mean(.data$rmse),
    rmse_sd = stats::sd(.data$rmse),
    r_squared_mean = mean(.data$r_squared),
    r_squared_sd = stats::sd(.data$r_squared),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$pearson_r_mean))

metrics_summary
```

## Replicate-Level Metrics

```{r}
metrics_tbl
```

## Scatterplot Rows By Method (Seed 123)

```{r}
vis <- res_list[[1]]
```

```{r}
plot_method_triptych(
  "adjust_inverse_penetration",
  vis$adj$adjust_inverse_penetration,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_selection_rate`

```{r}
plot_method_triptych(
  "adjust_selection_rate",
  vis$adj$adjust_selection_rate,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_selection_rate2`

```{r}
plot_method_triptych(
  "adjust_selection_rate2",
  vis$adj$adjust_selection_rate2,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_raking_ratio`

```{r}
plot_method_triptych(
  "adjust_raking_ratio",
  vis$adj$adjust_raking_ratio,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_coefficient`

```{r}
plot_method_triptych(
  "adjust_coefficient",
  vis$adj$adjust_coefficient,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_multilevel_bayes` frequentist placeholder

```{r}
plot_method_triptych(
  "adjust_multilevel_bayes (frequentist placeholder)",
  vis$adj$adjust_multilevel_bayes,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

## Plot Comparison (Mean Across Seeds)

```{r}
metrics_summary |>
  dplyr::mutate(method = forcats::fct_reorder(method, pearson_r_mean)) |>
  ggplot(aes(x = method, y = pearson_r_mean)) +
  geom_col(fill = "#2B8CBE", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Pearson Correlation by Method (Mean Across Seeds)",
    x = "Method",
    y = "Pearson r"
  )
```

```{r}
metrics_summary |>
  dplyr::mutate(method = forcats::fct_reorder(method, r_squared_mean)) |>
  ggplot(aes(x = method, y = r_squared_mean)) +
  geom_col(fill = "#41AE76", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "R-squared by Method (Mean Across Seeds)",
    x = "Method",
    y = "R-squared"
  )
```

## Inspect Method 6 Attributes (Seed 123)

```{r}
m6_vis <- vis$adj$adjust_multilevel_bayes
head(attr(m6_vis, "coefficients"))
attr(m6_vis, "formula")
attr(m6_vis, "backend")
attr(m6_vis, "model_engine")
attr(m6_vis, "scenario")
attr(m6_vis, "model_family")
```

The example above is S1: one source and one time unit. For repeated
observations, keep the same function but change `scenario`,
`source_col`, `time_col`, and `repeated_observation`: use `s2` and
`repeated_observation = "time"` for one source over multiple periods, `s3` and
`"source"` for multiple sources in one period, and `s4` plus `"source_time"`
when both source and time vary.

## Render

```bash
quarto render vignettes/method-comparison.qmd
```
