---
title: "debiasR: Empirical Methods Walkthrough"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This walkthrough runs and visualises the implemented `debiasR` correction methods using the empirical MSOA travel-to-work data from `debiasRdata`.

## Setup

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(tidyverse)
source(file.path(pkg_root, "style", "data-visualisation_theme.R"))
```

```{r}
has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work walkthrough.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 30)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work

  example_mpd_od <- msoa_OD_travel2work
  example_benchmark_od <- census_msoa_OD_travel2work
  example_coverage <- example_data$coverage
  example_covariates <- example_data$covariates
  example_distance <- tibble(origin = character(), destination = character(), distance_km = numeric())
  example_active_users <- example_data$active_users
  example_population <- example_data$population

  tibble(
    dataset = c(
      "msoa_OD_travel2work",
      "census_msoa_OD_travel2work",
      "coverage",
      "covariates",
      "active_users",
      "population"
    ),
    rows = c(
      nrow(example_mpd_od),
      nrow(example_benchmark_od),
      nrow(example_coverage),
      nrow(example_covariates),
      nrow(example_active_users),
      nrow(example_population)
    )
  )
}
```

## Empirical Distribution Check

```{r}
tibble(
  series = c("benchmark_flow", "mpd_flow"),
  zero_share = c(
    mean(example_benchmark_od$flow == 0),
    mean(example_mpd_od$flow == 0)
  ),
  q50_log1p = c(
    quantile(log1p(example_benchmark_od$flow), 0.5),
    quantile(log1p(example_mpd_od$flow), 0.5)
  ),
  q90_log1p = c(
    quantile(log1p(example_benchmark_od$flow), 0.9),
    quantile(log1p(example_mpd_od$flow), 0.9)
  ),
  q99_log1p = c(
    quantile(log1p(example_benchmark_od$flow), 0.99),
    quantile(log1p(example_mpd_od$flow), 0.99)
  )
)
```

## Plot Helpers

```{r}
plot_origin_totals <- function(df, flow_col = "flow", title_text) {
  df |>
    group_by(origin) |>
    summarise(total_flow = sum(.data[[flow_col]], na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = reorder(origin, total_flow), y = total_flow)) +
    geom_col(fill = "#121282", alpha = 0.9) +
    coord_flip() +
    labs(
      title = title_text,
      x = "Origin",
      y = "Total flow"
    ) +
    theme_plot_tufte()
}

plot_adj_vs_benchmark <- function(adj_df, benchmark_df, title_text) {
  cmp <- adj_df |>
    select(origin, destination, flow_adj) |>
    inner_join(
      benchmark_df |>
        select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    )

  ggplot(cmp, aes(x = flow_bench, y = flow_adj)) +
    geom_point(color = "#121282", alpha = 0.75, size = 2) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#FE7105") +
    labs(
      title = title_text,
      x = "Benchmark flow",
      y = "Adjusted flow"
    ) +
    theme_plot_tufte()
}

collect_metrics <- function(method_name, adj_df) {
  v <- validate_flow_overall(
    adj_df = adj_df,
    benchmark_od_df = example_benchmark_od,
    method_name = method_name,
    return_joined = FALSE
  )

  tibble(
    method = v$method,
    n = v$n,
    pearson_r = v$pearson_r,
    spearman_rho = v$spearman_rho,
    rmse = v$rmse,
    mae = v$mae,
    mape = v$mape,
    r_squared = v$r_squared
  )
}
```

## Baseline Visuals

```{r}
plot_origin_totals(example_mpd_od, "flow", "Observed MPD Flows by Origin")
```

```{r}
plot_origin_totals(example_benchmark_od, "flow", "Benchmark Flows by Origin")
```

## 1) `adjust_inverse_penetration`

```{r}
adj_m1 <- adjust_inverse_penetration(
  mpd_od_df = example_mpd_od,
  coverage_df = example_coverage,
  weight_by = "both"
)

collect_metrics("adjust_inverse_penetration", adj_m1)
```

```{r}
plot_adj_vs_benchmark(adj_m1, example_benchmark_od, "Method 1: Adjusted vs Benchmark")
```

## 2) `adjust_selection_rate`

```{r}
adj_m2 <- adjust_selection_rate(
  mpd_od_df = example_mpd_od,
  coverage_df = example_coverage,
  covariates_df = example_covariates,
  income_col = "income_norm",
  weight_by = "origin",
  benchmark_od_df = example_benchmark_od,
  calibration_aggregate = "origin"
)

list(r_global = attr(adj_m2, "r_global"))
```

```{r}
plot_adj_vs_benchmark(adj_m2, example_benchmark_od, "Method 2: Adjusted vs Benchmark")
```

## 3) `adjust_selection_rate2`

```{r}
adj_m3 <- adjust_selection_rate2(
  mpd_od_df = example_mpd_od,
  coverage_df = example_coverage,
  weight_by = "origin",
  benchmark_od_df = example_benchmark_od,
  calibration_aggregate = "origin"
)

list(k = attr(adj_m3, "k"))
```

```{r}
plot_adj_vs_benchmark(adj_m3, example_benchmark_od, "Method 3: Adjusted vs Benchmark")
```

## 4) `adjust_raking_ratio`

```{r}
adj_m4 <- adjust_raking_ratio(
  mpd_od_df = example_mpd_od,
  benchmark_od_df = example_benchmark_od
)

list(
  ipf_converged = attr(adj_m4, "ipf_converged"),
  ipf_iterations = attr(adj_m4, "ipf_iterations")
)
```

```{r}
plot_adj_vs_benchmark(adj_m4, example_benchmark_od, "Method 4: Adjusted vs Benchmark")
```

## 5) `adjust_coefficient`

```{r}
adj_m5 <- adjust_coefficient(
  mpd_od_df = example_mpd_od,
  benchmark_od_df = example_benchmark_od,
  model_family = "ols",
  level = "od"
)

list(coef = attr(adj_m5, "coef"))
```


```{r}
plot_adj_vs_benchmark(adj_m5, example_benchmark_od, "Method 5: Adjusted vs Benchmark")
```

## 6) `adjust_multilevel_bayes` (prototype)

```{r}
adj_m6 <- NULL

if (requireNamespace("rstanarm", quietly = TRUE) && nrow(example_distance) > 0L) {
  adj_m6 <- adjust_multilevel_bayes(
    mpd_od_df = example_mpd_od,
    coverage_df = example_coverage,
    covariates_df = example_covariates,
    distance_df = example_distance,
    random_intercept = "origin",
    model_family = "poisson",
    flow_adj_summary = "median",
    iter = 500,
    chains = 2,
    seed = 123,
    refresh = 0
  )

  attr(adj_m6, "coefficients")
} else {
  "Empirical distance data are not available in this example, so the Bayesian prototype is skipped."
}
```

```{r}
if (!is.null(adj_m6)) {
  plot_adj_vs_benchmark(adj_m6, example_benchmark_od, "Method 6: Adjusted vs Benchmark")
}
```

## Compare Validation Metrics Across Methods

```{r}
metrics_tbl <- bind_rows(
  collect_metrics("adjust_inverse_penetration", adj_m1),
  collect_metrics("adjust_selection_rate", adj_m2),
  collect_metrics("adjust_selection_rate2", adj_m3),
  collect_metrics("adjust_raking_ratio", adj_m4),
  collect_metrics("adjust_coefficient", adj_m5),
  if (!is.null(adj_m6)) collect_metrics("adjust_multilevel_bayes", adj_m6)
) |>
  arrange(desc(pearson_r))

metrics_tbl
```

```{r}
metrics_tbl |>
  mutate(method = fct_reorder(method, pearson_r)) |>
  ggplot(aes(x = method, y = pearson_r)) +
  geom_col(fill = "#41AE76", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Pearson Correlation by Method",
    x = "Method",
    y = "Pearson r"
  ) +
  theme_plot_tufte()
```

## Validation Helper Examples

### `validate_flow_overall`

```{r}
val_benchmark_m1 <- validate_flow_overall(
  adj_df = adj_m1,
  benchmark_od_df = example_benchmark_od,
  method_name = "adjust_inverse_penetration",
  return_joined = FALSE
)

val_benchmark_m1
```

### `validate_flow_pairs`

```{r}
val_all_m1 <- validate_flow_pairs(
  adj_df = adj_m1,
  benchmark_od_df = example_benchmark_od
)

val_all_m1 |>
  select(origin, destination, mpd_flow, adj_flow, benchmark_flow, diff_mpd_adj, diff_adj_benchmark) |>
  slice_head(n = 10)
```

### `validate_flow_residuals`

```{r}
residual_m1 <- validate_flow_residuals(
  adj_df = adj_m1,
  benchmark_od_df = example_benchmark_od,
  method_name = "adjust_inverse_penetration"
)

residual_m1$summary
```

### `validate_flow_residual_structure`

```{r}
structure_m1 <- validate_flow_residual_structure(
  adj_df = adj_m1,
  benchmark_od_df = example_benchmark_od,
  method_name = "adjust_inverse_penetration",
  covariate_df = example_covariates,
  covariate_col = "income_norm"
)

structure_m1$summary
```

### `validate_flow_distribution`

```{r}
distribution_m1 <- validate_flow_distribution(
  adj_df = adj_m1,
  benchmark_od_df = example_benchmark_od,
  method_name = "adjust_inverse_penetration",
  weight_by = "benchmark_origin_total"
)

distribution_m1$summary
```

## `measure_bias` Example

```{r}
coverage_for_bias <- example_coverage |>
  select(origin, population, user_count)

bias_df <- measure_bias(coverage_for_bias)
bias_df
```

```{r}
bias_df |>
  ggplot(aes(x = reorder(origin, coverage_score), y = coverage_score)) +
  geom_col(fill = "#2B8CBE", alpha = 0.9) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#D95F0E") +
  coord_flip() +
  labs(
    title = "Coverage Score by Origin",
    x = "Origin",
    y = "coverage_score = user_count / population"
  ) +
  theme_plot_tufte()
```

## Stage 3 Bias Residual Diagnostics

```{r}
areas <- sort(unique(example_coverage$origin))

area_neighbors <- tibble(
  area = areas,
  neighbor = dplyr::lead(areas, default = areas[1])
) |>
  bind_rows(tibble(
    area = areas,
    neighbor = dplyr::lag(areas, default = areas[length(areas)])
  ))

bias_structure <- validate_bias_residual_structure(
  coverage_df = example_coverage,
  benchmark_od_df = example_benchmark_od,
  area_neighbors = area_neighbors,
  covariate_df = example_covariates,
  covariate_col = "internet_access"
)

bias_structure$summary
```

```{r}
bias_structure$area_level |>
  select(
    area,
    population,
    user_count,
    coverage_score,
    expected_user_count,
    user_count_residual,
    coverage_score_residual,
    population_lm_expected_user_count,
    population_lm_residual,
    selected_residual
  ) |>
  arrange(desc(abs(selected_residual))) |>
  slice_head(n = 10)
```

```{r}
bias_structure$benchmark_flow_correlation
```

## Render

```bash
quarto render vignettes/testing/empirical-methods-walkthrough.qmd
```
