---
title: "debiasR: Method Comparison Walkthrough"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This vignette compares the stable deterministic adjustment methods in `debiasR`
and includes the Bayesian helper as the main experimental prototype:

1. `adjust_inverse_penetration`
2. `adjust_selection_rate`
3. `adjust_selection_rate2`
4. `adjust_raking_ratio`
5. `adjust_coefficient`
6. `adjust_multilevel_bayes` (stage-1 prototype with observed and complete-grid prediction scopes)

It uses a fast but fair comparison protocol:

1. same stratified OD subset for all methods per replicate,
2. repeated over multiple seeds,
3. Bayesian prototype with lighter MCMC settings for illustration.

`adjust_multilevel_bayes()` is not part of the stable deterministic path. Treat
Method 6 results here as exploratory unless the optional Bayesian diagnostics,
complete-grid checks, and runtime gate are explicitly reviewed for the rendered
data.

## Setup

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)
library(ggplot2)
library(forcats)
library(tidyr)
library(readr)
library(stringr)
library(purrr)

# Parallel setup for Stan chains
options(mc.cores = max(1L, parallel::detectCores(logical = FALSE) - 1L))
Sys.setenv(STAN_NUM_THREADS = 1)

if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("Install 'patchwork' to render the multi-panel comparison plots.")
}
```

## Data

```{r}
has_debiasRdata <- requireNamespace("debiasRdata", quietly = TRUE)
if (!has_debiasRdata) {
  cat("Install `debiasRdata` to run the empirical MSOA travel-to-work comparison.\n")
  knitr::knit_exit()
}

if (has_debiasRdata) {
  example_data <- debiasR_example_data(n_areas = 40)
  msoa_OD_travel2work <- example_data$msoa_OD_travel2work
  census_msoa_OD_travel2work <- example_data$census_msoa_OD_travel2work

  example_mpd_od <- msoa_OD_travel2work
  example_benchmark_od <- census_msoa_OD_travel2work
  example_coverage <- example_data$coverage
  example_covariates <- example_data$covariates
  example_distance <- tibble::tibble(
    origin = character(),
    destination = character(),
    distance_km = numeric()
  )

  tibble::tibble(
    dataset = c(
      "msoa_OD_travel2work",
      "census_msoa_OD_travel2work",
      "coverage",
      "covariates"
    ),
    rows = c(
      nrow(example_mpd_od),
      nrow(example_benchmark_od),
      nrow(example_coverage),
      nrow(example_covariates)
    )
  )

  # coverage_df schema used by methods in this walkthrough
  names(example_coverage)
}
```

## Empirical Data Check

```{r}
example_data$metadata

tibble::tibble(
  series = c("benchmark_flow", "mpd_flow", "coverage_score"),
  n = c(
    nrow(example_benchmark_od),
    nrow(example_mpd_od),
    nrow(example_coverage)
  ),
  median = c(
    stats::median(example_benchmark_od$flow),
    stats::median(example_mpd_od$flow),
    stats::median(example_coverage$user_count / example_coverage$population)
  ),
  q90 = c(
    stats::quantile(example_benchmark_od$flow, 0.9),
    stats::quantile(example_mpd_od$flow, 0.9),
    stats::quantile(example_coverage$user_count / example_coverage$population, 0.9)
  )
)
```

## Fair-Lite Settings

```{r}
n_sub <- 12000
fair_seeds <- c(123, 456, 789)
m6_iter <- 200
m6_chains <- 1

make_flow_strata <- function(flow_vec) {
  pos <- flow_vec[flow_vec > 0]
  q50 <- stats::quantile(pos, 0.50, na.rm = TRUE)
  q90 <- stats::quantile(pos, 0.90, na.rm = TRUE)

  dplyr::case_when(
    flow_vec <= 0 ~ "zero",
    flow_vec <= q50 ~ "low",
    flow_vec <= q90 ~ "mid",
    TRUE ~ "high"
  )
}

draw_fair_subset <- function(mpd_df, n_target, seed) {
  set.seed(seed)
  tmp <- mpd_df |>
    dplyr::mutate(stratum = make_flow_strata(.data$flow))

  counts <- tmp |>
    dplyr::count(.data$stratum, name = "n_stratum") |>
    dplyr::mutate(
      n_take = pmax(1L, round(n_target * .data$n_stratum / sum(.data$n_stratum)))
    )

  # Force exact sample size by adjusting the largest stratum.
  diff_n <- n_target - sum(counts$n_take)
  if (diff_n != 0) {
    idx <- which.max(counts$n_stratum)
    counts$n_take[idx] <- pmax(1L, counts$n_take[idx] + diff_n)
  }

  sampled <- lapply(counts$stratum, function(s) {
    df_s <- tmp[tmp$stratum == s, , drop = FALSE]
    n_take_s <- counts$n_take[counts$stratum == s][1]
    n_take_s <- min(n_take_s, nrow(df_s))
    df_s[sample.int(nrow(df_s), size = n_take_s), , drop = FALSE]
  }) |>
    dplyr::bind_rows()

  sampled |>
    dplyr::select(-.data$stratum)
}
```

## Validation functions

```{r}
collect_metrics <- function(method_name, adj_df, benchmark_df, seed_id) {
  v <- validate_flow_overall(
    adj_df = adj_df,
    benchmark_od_df = benchmark_df,
    method_name = method_name,
    return_joined = FALSE
  )

  tibble::tibble(
    seed = seed_id,
    method = v$method,
    n = v$n,
    pearson_r = v$pearson_r,
    spearman_rho = v$spearman_rho,
    rmse = v$rmse,
    mae = v$mae,
    mape = v$mape,
    r_squared = v$r_squared
  )
}

build_compare_df <- function(adj_df, mpd_df, benchmark_df) {
  adj_df |>
    dplyr::select(origin, destination, flow_adj) |>
    dplyr::inner_join(
      mpd_df |>
        dplyr::select(origin, destination, flow_mpd = flow),
      by = c("origin", "destination")
    ) |>
    dplyr::inner_join(
      benchmark_df |>
        dplyr::select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    )
}

plot_method_triptych <- function(method_label, adj_df, mpd_df, benchmark_df) {
  df <- build_compare_df(adj_df, mpd_df = mpd_df, benchmark_df = benchmark_df)

  p1 <- ggplot(df, aes(x = flow_mpd, y = flow_adj)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#2B8CBE") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "Adjusted vs MPD",
      x = "Mobile-phone-derived flow",
      y = "Adjusted flow"
    )

  p2 <- ggplot(df, aes(x = flow_bench, y = flow_adj)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#41AE76") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "Adjusted vs Benchmark",
      x = "Benchmark flow",
      y = "Adjusted flow"
    )

  p3 <- ggplot(df, aes(x = flow_bench, y = flow_mpd)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#756BB1") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "MPD vs Benchmark",
      x = "Benchmark flow",
      y = "Mobile-phone-derived flow"
    )

  patchwork::wrap_plots(p1, p2, p3, nrow = 1) +
    patchwork::plot_annotation(title = method_label)
}
```

## Run Methods (Fair-Lite)

```{r}
run_methods_once <- function(seed_id) {
  mpd_sub <- draw_fair_subset(example_mpd_od, n_target = n_sub, seed = seed_id)
  benchmark_sub <- example_benchmark_od |>
    dplyr::inner_join(
      mpd_sub |> dplyr::select(origin, destination),
      by = c("origin", "destination")
    )

  distance_sub <- example_distance |>
    dplyr::inner_join(
      mpd_sub |> dplyr::select(origin, destination),
      by = c("origin", "destination")
    )

  adj_m1 <- adjust_inverse_penetration(
    mpd_od_df = mpd_sub,
    coverage_df = example_coverage,
    weight_by = "both"
  )

  adj_m2 <- adjust_selection_rate(
    mpd_od_df = mpd_sub,
    coverage_df = example_coverage,
    covariates_df = example_covariates,
    income_col = "income_norm",
    weight_by = "origin",
    benchmark_od_df = benchmark_sub,
    calibration_aggregate = "origin"
  )

  adj_m3 <- adjust_selection_rate2(
    mpd_od_df = mpd_sub,
    coverage_df = example_coverage,
    weight_by = "origin",
    benchmark_od_df = benchmark_sub,
    calibration_aggregate = "origin"
  )

  adj_m4 <- adjust_raking_ratio(
    mpd_od_df = mpd_sub,
    benchmark_od_df = benchmark_sub
  )

  adj_m5 <- adjust_coefficient(
    mpd_od_df = mpd_sub,
    benchmark_od_df = benchmark_sub,
    model_family = "ols",
    level = "od"
  )

  adj_m6 <- NULL
  if (nrow(distance_sub) > 0L) {
    adj_m6 <- adjust_multilevel_bayes(
      mpd_od_df = mpd_sub,
      coverage_df = example_coverage,
      covariates_df = example_covariates,
      distance_df = distance_sub,
      random_intercept = "origin",
      model_family = "poisson",
      backend = "auto",
      flow_adj_summary = "median",
      iter = m6_iter,
      chains = m6_chains,
      seed = seed_id,
      refresh = 0
    )
  }

  metrics <- dplyr::bind_rows(
    collect_metrics("adjust_inverse_penetration", adj_m1, benchmark_sub, seed_id),
    collect_metrics("adjust_selection_rate", adj_m2, benchmark_sub, seed_id),
    collect_metrics("adjust_selection_rate2", adj_m3, benchmark_sub, seed_id),
    collect_metrics("adjust_raking_ratio", adj_m4, benchmark_sub, seed_id),
    collect_metrics("adjust_coefficient", adj_m5, benchmark_sub, seed_id),
    if (!is.null(adj_m6)) {
      collect_metrics("adjust_multilevel_bayes", adj_m6, benchmark_sub, seed_id)
    }
  )

  list(
    mpd_sub = mpd_sub,
    benchmark_sub = benchmark_sub,
    adj = list(
      adjust_inverse_penetration = adj_m1,
      adjust_selection_rate = adj_m2,
      adjust_selection_rate2 = adj_m3,
      adjust_raking_ratio = adj_m4,
      adjust_coefficient = adj_m5,
      adjust_multilevel_bayes = adj_m6
    ),
    metrics = metrics
  )
}

res_list <- purrr::map(fair_seeds, run_methods_once)
metrics_tbl <- dplyr::bind_rows(purrr::map(res_list, "metrics"))

metrics_summary <- metrics_tbl |>
  dplyr::group_by(.data$method) |>
  dplyr::summarise(
    n_mean = mean(.data$n),
    pearson_r_mean = mean(.data$pearson_r),
    pearson_r_sd = stats::sd(.data$pearson_r),
    rmse_mean = mean(.data$rmse),
    rmse_sd = stats::sd(.data$rmse),
    r_squared_mean = mean(.data$r_squared),
    r_squared_sd = stats::sd(.data$r_squared),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$pearson_r_mean))

metrics_summary
```

## Replicate-Level Metrics

```{r}
metrics_tbl
```

## Scatterplot Rows By Method (Seed 123)

```{r}
vis <- res_list[[1]]
```

```{r}
plot_method_triptych(
  "adjust_inverse_penetration",
  vis$adj$adjust_inverse_penetration,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_selection_rate`

```{r}
plot_method_triptych(
  "adjust_selection_rate",
  vis$adj$adjust_selection_rate,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_selection_rate2`

```{r}
plot_method_triptych(
  "adjust_selection_rate2",
  vis$adj$adjust_selection_rate2,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_raking_ratio`

```{r}
plot_method_triptych(
  "adjust_raking_ratio",
  vis$adj$adjust_raking_ratio,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_coefficient`

```{r}
plot_method_triptych(
  "adjust_coefficient",
  vis$adj$adjust_coefficient,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_multilevel_bayes`

```{r}
if (!is.null(vis$adj$adjust_multilevel_bayes)) {
  plot_method_triptych(
    "adjust_multilevel_bayes",
    vis$adj$adjust_multilevel_bayes,
    mpd_df = vis$mpd_sub,
    benchmark_df = vis$benchmark_sub
  )
} else {
  "Empirical distance data are not available in this example, so the Bayesian prototype is skipped."
}
```

## Plot Comparison (Mean Across Seeds)

```{r}
metrics_summary |>
  dplyr::mutate(method = forcats::fct_reorder(method, pearson_r_mean)) |>
  ggplot(aes(x = method, y = pearson_r_mean)) +
  geom_col(fill = "#2B8CBE", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Pearson Correlation by Method (Mean Across Seeds)",
    x = "Method",
    y = "Pearson r"
  )
```

```{r}
metrics_summary |>
  dplyr::mutate(method = forcats::fct_reorder(method, r_squared_mean)) |>
  ggplot(aes(x = method, y = r_squared_mean)) +
  geom_col(fill = "#41AE76", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "R-squared by Method (Mean Across Seeds)",
    x = "Method",
    y = "R-squared"
  )
```

## Inspect Method 6 Attributes (Seed 123)

```{r}
m6_vis <- vis$adj$adjust_multilevel_bayes
if (!is.null(m6_vis)) {
  head(attr(m6_vis, "coefficients"))
  attr(m6_vis, "formula")
  attr(m6_vis, "backend")
  attr(m6_vis, "model_family")
} else {
  "Empirical distance data are not available in this example, so Method 6 was skipped."
}
```

## Render

```bash
quarto render vignettes/method-comparison.qmd
```
